# Statistical Testing

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(knitr)
library(kableExtra)
library(patchwork)
```

---

# Introduction {.unnumbered}

::: {.callout-note icon=false}
## Why Does This Matter?

In health research, we constantly ask questions that require a decision:

- Does this new drug lower blood pressure *more* than the standard treatment?
- Is there a relationship between smoking and lung disease?
- Do recovery times differ between two surgical techniques?

We cannot answer these questions simply by looking at sample means or
proportions — random chance alone can create apparent differences between
groups. **Statistical testing** gives us a formal, objective framework
for deciding whether an observed difference is likely to be real or
simply due to chance.

This is the most widely used tool in health research, and understanding
it is essential for reading and interpreting scientific literature.
:::

---

# Concept of Hypothesis Testing

## The Logic of Hypothesis Testing

Statistical testing works by assuming the opposite of what we want to prove,
then asking: *"If nothing were really happening, how surprising would our
data be?"*

If the data would be very surprising under the assumption that nothing is
happening, we conclude that something probably *is* happening.

---

## The Two Hypotheses

Every hypothesis test involves two competing statements:

### Null Hypothesis ($H_0$)

The **null hypothesis** states that there is **no effect, no difference,
or no relationship** in the population. It is the assumption we start with
and attempt to disprove.

> *Examples:*
> - The mean blood pressure is 120 mmHg. ($H_0: \mu = 120$)
> - There is no difference in recovery time between the two groups. ($H_0: \mu_1 = \mu_2$)
> - There is no association between smoking status and disease. ($H_0:$ no association)

### Alternative Hypothesis ($H_1$ or $H_a$)

The **alternative hypothesis** states that there **is** an effect,
difference, or relationship. This is usually what the researcher
believes or wants to demonstrate.

> *Examples:*
> - The mean blood pressure is not 120 mmHg. ($H_1: \mu \neq 120$)
> - Recovery time differs between the two groups. ($H_1: \mu_1 \neq \mu_2$)
> - There is an association between smoking and disease. ($H_1:$ association exists)

---

## One-Tailed vs. Two-Tailed Tests

The alternative hypothesis can be **directional** or **non-directional**:

```{r}
#| label: tbl-tails
#| tbl-cap: "One-tailed vs. Two-tailed Tests"

tails <- data.frame(
  Type = c("Two-tailed", "One-tailed (right)", "One-tailed (left)"),
  `H1` = c("mu != mu0  (different from)", "mu > mu0  (greater than)", "mu < mu0  (less than)"),
  `Use When` = c(
    "You want to detect a difference in either direction (most common)",
    "You specifically expect the value to be higher",
    "You specifically expect the value to be lower"
  )
)

kable(tails, col.names = c("Test Type", "H1", "Use When"),
      align = c("l","l","l")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = 13) |>
  column_spec(1, bold = TRUE, color = "#2c7bb6") |>
  row_spec(1, background = "#eaf4fb")
```

::: {.callout-tip icon=false}
## Practical Advice
In most health research, use a **two-tailed test** unless you have a
strong, pre-specified reason to test in only one direction.
Two-tailed tests are more conservative and more widely accepted.
:::

---

## The p-value

The **p-value** is the probability of obtaining results *at least as extreme
as those observed*, assuming the null hypothesis is true.

$$p\text{-value} = P(\text{observed result or more extreme} \mid H_0 \text{ is true})$$

- A **small p-value** means the observed data would be very unlikely if
 $H_0$ were true evidence against $H_0$
- A **large p-value** means the observed data are consistent with $H_0$
  insufficient evidence to reject $H_0$

---

## The Significance Level (α)

The **significance level** $\alpha$ is the threshold we set in advance
for deciding whether the p-value is "small enough" to reject $H_0$.

The most common choice in health research is $\alpha = 0.05$.

| Decision Rule | |
|---|---|
| If $p\text{-value} \leq \alpha$ | **Reject $H_0$** — result is statistically significant |
| If $p\text{-value} > \alpha$ | **Fail to reject $H_0$** — insufficient evidence |

::: {.callout-warning icon=false}
## Common Misinterpretations

**"Failing to reject $H_0$" is NOT the same as proving $H_0$ is true.**
It simply means the data do not provide strong enough evidence against it.

**"Statistically significant" does NOT necessarily mean clinically important.**
With a very large sample, even a trivially small difference can become
statistically significant. Always consider the practical meaning of results.
:::

---

## Type I and Type II Errors

When making a decision about $H_0$, two types of errors are possible:

```{r}
#| label: tbl-errors
#| tbl-cap: "Type I and Type II Errors in Hypothesis Testing"

errors <- data.frame(
  ` ` = c("Reject H0", "Fail to reject H0"),
 `H0 is actually TRUE` = c("Type I Error (alpha) - False Positive",
                            "Correct Decision"),
  `H0 is actually FALSE` = c("Correct Decision (Power = 1 - beta)",
                             "Type II Error (beta) - False Negative")
)

kable(errors, col.names = c("Our Decision", "H0 is TRUE", "H0 is FALSE"),
      align = c("l","c","c")) |>
  kable_styling(bootstrap_options = c("bordered"),
                full_width = FALSE, font_size = 13) |>
  column_spec(1, bold = TRUE) |>
  column_spec(2, background = c("#fde0d9","#d9f0d3")) |>
  column_spec(3, background = c("#d9f0d3","#fde0d9"))
```

- **Type I error (α):** Concluding there is an effect when there is none
  (*false positive*). Controlled by choosing α (usually 0.05).
- **Type II error (β):** Missing a real effect (*false negative*).
  The probability of correctly detecting a real effect is called **power** = $1 - \beta$.

---

## General Steps in Hypothesis Testing

::: {.callout-important icon=false}
## The 5-Step Framework

**Step 1 — State the hypotheses:** Write $H_0$ and $H_1$ clearly.

**Step 2 — Choose the significance level:** Usually $\alpha = 0.05$.

**Step 3 — Select and compute the test statistic:**
A number calculated from the sample that measures how far the data
are from what $H_0$ predicts.

**Step 4 — Find the p-value:**
The probability of getting a test statistic this extreme if $H_0$ is true.

**Step 5 — Make a decision and interpret:**
Compare p-value to α. State your conclusion in the context of the problem.
:::

---

# Statistical Testing on Categorical Data

## Chi-Square Test for Two Independent Groups

### When to Use

The **chi-square test of independence** is used when:

- The outcome variable is **categorical (nominal)**
- We are comparing **two or more independent groups**
- We want to test whether there is an **association** between two categorical variables

### The Contingency Table

Data are arranged in a **contingency table** (also called a cross-tabulation):

$$\begin{array}{|l|c|c|c|}
\hline
& \textbf{Outcome: Yes} & \textbf{Outcome: No} & \textbf{Row Total} \\
\hline
\textbf{Group 1} & a & b & a+b \\
\textbf{Group 2} & c & d & c+d \\
\hline
\textbf{Column Total} & a+c & b+d & n \\
\hline
\end{array}$$

### The Test Statistic

$$\chi^2 = \sum \frac{(O - E)^2}{E}$$

Where:
- $O$ = observed frequency (actual count in each cell)
- $E$ = expected frequency = $\dfrac{\text{Row total} \times \text{Column total}}{n}$

The sum is taken over **all cells** in the table.

The test statistic follows a **chi-square distribution** with
$df = (\text{rows} - 1) \times (\text{columns} - 1)$.

For a $2 \times 2$ table: $df = (2-1)(2-1) = 1$.

::: {.callout-warning icon=false}
## Condition for Valid Chi-Square Test
All **expected frequencies** must be $\geq 5$.
If this condition is violated, use **Fisher's exact test** instead
(available in SPSS).
:::

### Chi-Square Critical Values

```{r}
#| label: tbl-chisq
#| tbl-cap: "Chi-Square Critical Values at alpha = 0.05"

df_chi  <- 1:10
cv_chi  <- round(qchisq(0.95, df = df_chi), 3)

chi_tab <- data.frame(df = df_chi, critical = cv_chi)

kable(chi_tab, col.names = c("Degrees of Freedom (df)", "chi2 critical value (alpha = 0.05)"),
      align = c("c","c")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = 13) |>
  column_spec(1, bold = TRUE, color = "#2c7bb6") |>
  row_spec(1, background = "#eaf4fb")
```

---

# Statistical Testing on Rank Data

## Mann-Whitney U Test for Two Independent Groups

### When to Use

The **Mann-Whitney U test** (also called the Wilcoxon rank-sum test) is used when:

- The outcome variable is **ordinal**, or
- The outcome is **continuous but not normally distributed** (especially with
  small samples or obvious skewness)
- We are comparing **two independent groups**

It is a **non-parametric** test — it does not assume the data follow
any particular distribution.

### How It Works

Instead of comparing means, the Mann-Whitney test compares the **ranks**
of observations between the two groups.

**Steps:**

1. Combine all observations from both groups and **rank them** from
   smallest (rank 1) to largest.
2. If there are **tied values**, assign the average rank to each tied observation.
3. Calculate the **sum of ranks** for each group: $W_1$ and $W_2$.
4. The test statistic $U$ is computed from the rank sums.
5. Compare to the critical value or obtain the p-value from software (SPSS).

::: {.callout-tip icon=false}
## Intuition Behind the Test
If the two groups come from the same distribution, their ranks should be
**randomly mixed** between the groups — neither group should consistently
have higher or lower ranks. A significant result means one group
consistently has higher ranks (i.e., higher values) than the other.
:::

### The U Statistic

$$U_1 = n_1 n_2 + \frac{n_1(n_1+1)}{2} - W_1$$

$$U_2 = n_1 n_2 + \frac{n_2(n_2+1)}{2} - W_2$$

$$U = \min(U_1, U_2)$$

Where $n_1$, $n_2$ are the sample sizes and $W_1$, $W_2$ are the rank sums
for groups 1 and 2 respectively.

For large samples, $U$ is approximately normally distributed and a Z-score
can be computed. In practice, **SPSS computes the p-value directly**.

## Wilcoxon Signed-Ranks Test for Two Related Groups

### When to Use

The **Wilcoxon signed-ranks test** (also called the Wilcoxon matched-pairs
test) is the non-parametric alternative to the **paired samples t-test**.
It is used when:

- The outcome variable is **ordinal**, or
- The **differences** between pairs are **not normally distributed**
  (especially with small samples)
- The two sets of measurements are **related (paired)** — for example,
  the same patients measured before and after treatment, or two
  measurements taken on the same blood sample

::: {.callout-important icon=false}
## Mann-Whitney or Wilcoxon?
Both tests work with ranks, so they are easy to confuse. The difference
is the **study design**, exactly as with the two t-tests:

- **Two independent groups** (different people in each group)
  $\rightarrow$ **Mann-Whitney U test**
- **Two related groups** (the same people measured twice)
  $\rightarrow$ **Wilcoxon signed-ranks test**
:::

### How It Works

The Wilcoxon test looks at the **differences within each pair**, and then
ranks the **sizes** of those differences.

**Steps:**

1. For each pair, calculate the difference $d_i = x_{i,1} - x_{i,2}$.
2. **Discard** any pair whose difference is zero.
3. Rank the **absolute** differences $|d_i|$ from smallest to largest,
   assigning average ranks to ties.
4. Attach to each rank the **sign** of its original difference.
5. Add the ranks belonging to the positive differences ($W^{+}$) and the
   ranks belonging to the negative differences ($W^{-}$).

### The Test Statistic

$$W = \min(W^{+},\; W^{-})$$

If the two conditions really are the same, the positive and negative
ranks should roughly balance, so $W^{+}$ and $W^{-}$ should be similar.
A **small** value of $W$ means the differences fall mostly in one
direction, which is evidence against $H_0$.

For larger samples the test statistic is converted to a Z-score, and this
is what **SPSS reports** together with the p-value:

$$H_0: \text{the two related measurements are equal}$$
$$H_1: \text{the two related measurements are different}$$

::: {.callout-tip icon=false}
## Reading the SPSS Output
Choose **Analyze** $\rightarrow$ **Nonparametric Tests** $\rightarrow$
**2 Related Samples**, and tick **Wilcoxon**. The *Test Statistics*
table reports $Z$ and the 2-tailed significance. As always, decide by
comparing the p-value with $\alpha$ — the sign of $Z$ only indicates the
direction of the ranks, not the importance of the result.
:::

### Worked Example

Blood clotting time was measured for **8 patients** using two solutions.
Blood was taken from each patient and divided into two parts, so the two
measurements are **paired**.

```{r}
#| label: tbl-wilcoxon
#| tbl-cap: "Blood clotting time using two solutions (n = 8 patients)"

pt   <- 1:8
sol1 <- c(3, 5, 5, 4, 3, 7, 8, 7)
gsol <- c(7, 6, 3, 8, 5, 9, 7, 9)
dif  <- sol1 - gsol

wil <- data.frame(Patient = pt, Shrimp = sol1, Standard = gsol, d = dif)

kable(wil,
      col.names = c("Patient", "Shrimp-shell extract",
                    "Standard solution", "Difference d"),
      align = c("c","c","c","c")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = 12) |>
  row_spec(0, bold = TRUE, color = "white", background = "#2c7bb6")
```

Because the sample is small ($n = 8$) and the outcome is a clotting time
that need not be normally distributed, the **Wilcoxon signed-ranks test**
is used rather than the paired t-test. SPSS reports:

$$Z = -1.706, \qquad p\text{-value} = 0.088$$

Since $p = 0.088 > \alpha = 0.05$, we **do not reject** $H_0$. There is
not enough evidence to say that the clotting times produced by the two
solutions differ, at the 0.05 level of significance.

---

# Statistical Testing on Quantitative Data

## Independent Samples t-Test

### When to Use

The **independent samples t-test** is used when:

- The outcome variable is **continuous (numerical)**
- We are comparing the **means of two independent groups**
- The data are approximately **normally distributed** in each group
  (or sample sizes are large enough, $n \geq 30$ per group)

### Hypotheses

$$H_0: \mu_1 = \mu_2 \quad \text{(no difference in population means)}$$
$$H_1: \mu_1 \neq \mu_2 \quad \text{(the population means differ)}$$

### The Test Statistic

$$t = \frac{\bar{x}_1 - \bar{x}_2}{SE_{\bar{x}_1 - \bar{x}_2}}$$

The standard error of the difference depends on whether the two groups
have **equal or unequal variances** — SPSS performs **Levene's test**
automatically to help you decide which version to use.

**Equal variances assumed (pooled):**

$$SE = s_p\sqrt{\frac{1}{n_1}+\frac{1}{n_2}}, \qquad
s_p = \sqrt{\frac{(n_1-1)s_1^2 + (n_2-1)s_2^2}{n_1+n_2-2}}$$

**Equal variances not assumed (Welch's):**

$$SE = \sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}$$

The test statistic follows a **t-distribution** with $df = n_1 + n_2 - 2$
(equal variances) or an approximated df (Welch's).

---

## Paired Samples t-Test

### When to Use

The **paired samples t-test** is used when:

- The outcome variable is **continuous (numerical)**
- Observations come in **natural pairs** — the same subject measured twice
  (e.g., before and after treatment), or matched pairs of subjects
- We want to test whether the **mean difference** within pairs is zero

### Hypotheses

$$H_0: \mu_d = 0 \quad \text{(mean difference is zero)}$$
$$H_1: \mu_d \neq 0 \quad \text{(mean difference is not zero)}$$

Where $\mu_d$ is the population mean of the within-pair differences.

### The Test Statistic

For each pair $i$, compute the difference: $d_i = x_{1i} - x_{2i}$

Then:

$$t = \frac{\bar{d}}{s_d / \sqrt{n}}$$

Where:
- $\bar{d}$ = mean of the differences
- $s_d$ = standard deviation of the differences
- $n$ = number of pairs

The test statistic follows a **t-distribution** with $df = n - 1$.

---

## Choosing the Right Test

```{r}
#| label: tbl-test-choice
#| tbl-cap: "Guide to Choosing the Correct Statistical Test (Two Groups)"

tests <- data.frame(
  `Data Type` = c("Categorical (Nominal)",
                   "Ordinal or non-normal continuous",
                   "Ordinal or non-normal continuous",
                   "Continuous (normal)",
                   "Continuous (normal)"),
  `Groups` = c("2 independent groups",
                "2 independent groups",
                "2 related (paired) groups",
                "2 independent groups",
                "2 related (paired) groups"),
  `Correct Test` = c("Chi-square test",
                      "Mann-Whitney U test",
                      "Wilcoxon signed-ranks test",
                      "Independent samples t-test",
                      "Paired samples t-test"),
 `SPSS Procedure` = c("Analyze Descriptive Statistics Crosstabs",
 "Analyze Nonparametric Legacy Dialogs 2 Independent Samples",
 "Analyze Nonparametric Legacy Dialogs 2 Related Samples",
 "Analyze Compare Means Independent Samples T Test",
 "Analyze Compare Means Paired Samples T Test")
)

kable(tests,
      col.names = c("Outcome Data Type", "Study Design",
                    "Statistical Test", "SPSS Procedure"),
      align = c("l","l","l","l")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = 12) |>
  column_spec(3, bold = TRUE, color = "#2c7bb6")
```

```{r}
#| label: fig-test-flowchart
#| fig-cap: "Decision flowchart for selecting the correct test (two-group comparisons)"
#| fig-align: center
#| fig-height: 5.4
#| fig-width: 9

boxes <- data.frame(
  x = c(5.5, 5.5, 2.0, 8.0, 2.0, 8.0, 5.6, 9.4, 4.2, 6.6, 8.6, 10.4),
  y = c(9.2, 7.7, 6.1, 6.1, 4.3, 4.3, 2.6, 2.6, 0.8, 0.8, 0.8, 0.8),
  label = c("Comparing 2 Groups", "Outcome variable type?",
            "Nominal\n(Categorical)", "Continuous\nor Ordinal",
            "Chi-square\nTest", "Groups independent\nor paired?",
            "Normal\ndistribution?", "Normal\ndistribution?",
            "Independent\nt-test", "Mann-Whitney\nU Test",
            "Paired\nt-test", "Wilcoxon\nsigned-ranks"),
  fill = c("#dce8f5","#dce8f5","#abd9e9","#abd9e9","#2c7bb6","#dce8f5",
           "#dce8f5","#dce8f5","#1a9641","#fdae61","#d7191c","#fdae61"),
  col  = c("black","black","black","black","white","black",
           "black","black","white","black","white","black")
)

seg <- data.frame(
  x    = c(5.5, 5.5, 5.5, 2.0, 8.0, 8.0, 9.4, 5.6, 5.6, 9.4, 9.4),
  xend = c(5.5, 2.2, 7.8, 2.0, 8.0, 5.8, 9.4, 4.4, 6.5, 8.7, 10.3),
  y    = c(8.9, 7.35, 7.35, 5.6, 5.6, 3.9, 3.9, 2.2, 2.2, 2.2, 2.2),
  yend = c(8.0, 6.6,  6.6,  4.7, 4.7, 3.0, 3.0, 1.2, 1.2, 1.2, 1.2)
)

edge_lab <- data.frame(
  x = c(3.4, 7.7, 6.4, 9.8, 4.3, 6.6, 8.5, 10.5),
  y = c(7.1, 7.1, 3.6, 3.6, 1.8, 1.8, 1.8, 1.8),
  label = c("Nominal","Continuous/\nOrdinal","Independent","Paired",
            "Yes","No","Yes","No")
)

ggplot() +
  geom_segment(data = seg, aes(x = x, xend = xend, y = y, yend = yend),
               arrow = arrow(length = unit(0.18, "cm")), color = "gray40") +
  geom_text(data = edge_lab, aes(x = x, y = y, label = label),
            size = 2.6, color = "gray30") +
  geom_label(data = boxes,
             aes(x = x, y = y, label = label, fill = fill, color = col),
             size = 2.8, label.r = unit(0.3, "lines"),
             label.padding = unit(0.35, "lines"), show.legend = FALSE) +
  scale_fill_identity() + scale_color_identity() +
  scale_x_continuous(limits = c(0.6, 11.6)) +
  scale_y_continuous(limits = c(0.1, 9.8)) +
  theme_void()
```

---

# Key Takeaways

::: {.callout-note icon=false}
## Chapter 5 Summary

1. Hypothesis testing uses sample data to decide between two competing
   claims: the **null hypothesis ($H_0$)** and the
   **alternative hypothesis ($H_1$)**.

2. The **p-value** is the probability of the observed data (or more extreme)
   if $H_0$ is true. A p-value $\leq \alpha$ (usually 0.05) leads us to
   **reject $H_0$**.

3. **Type I error (α):** rejecting a true $H_0$ (false positive).
   **Type II error (β):** failing to reject a false $H_0$ (false negative).

4. **Chi-square test:** for association between two categorical (nominal)
   variables in independent groups.

5. **Mann-Whitney U test:** for comparing two independent groups when the
   outcome is ordinal or the normality assumption is not met.

6. **Wilcoxon signed-ranks test:** for comparing two **related** measurements
   when the outcome is ordinal or the differences are not normally
   distributed. It is the non-parametric alternative to the paired t-test.

7. **Independent samples t-test:** for comparing means of two independent
   groups with continuous, approximately normal data.

8. **Paired samples t-test:** for comparing means of two related measurements
   (before/after, or matched pairs) with continuous data.

9. The two rank tests differ by **design**: Mann-Whitney for *independent*
   groups, Wilcoxon signed-ranks for *related* pairs.

10. Always choose the test based on **data type** and **study design** —
    using the wrong test leads to invalid conclusions.
:::

---

# Examples

## Example 1: Chi-Square Test

A study investigates whether **smoking status** is associated with
**hypertension** in a sample of 200 adults.

|  | Hypertension | No Hypertension | Total |
|---|:---:|:---:|:---:|
| Smoker | 45 | 55 | 100 |
| Non-smoker | 25 | 75 | 100 |
| **Total** | **70** | **130** | **200** |

Test at $\alpha = 0.05$ whether smoking is associated with hypertension.

::: {.callout-tip icon=false}
## Solution

**Step 1 — Hypotheses:**

$H_0$: Smoking status and hypertension are independent (no association).

$H_1$: Smoking status and hypertension are associated.

**Step 2 —** $\alpha = 0.05$

**Step 3 — Expected frequencies:**

$$E = \frac{\text{Row total} \times \text{Column total}}{n}$$

```{r}
#| label: tbl-expected
#| tbl-cap: "Observed (O) and Expected (E) Frequencies"

obs <- matrix(c(45, 55, 25, 75), nrow=2, byrow=TRUE)
row_totals <- rowSums(obs)
col_totals <- colSums(obs)
n <- sum(obs)

exp_freq <- outer(row_totals, col_totals) / n

result_tab <- data.frame(
  Group = c("Smoker", "Non-smoker"),
  O_hyp  = obs[,1],
  E_hyp  = exp_freq[,1],
  O_no   = obs[,2],
  E_no   = exp_freq[,2]
)

kable(result_tab,
      col.names = c("Group","O (Hyp.)","E (Hyp.)","O (No Hyp.)","E (No Hyp.)"),
      align = rep("c",5), digits = 2) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = 13) |>
  column_spec(1, bold = TRUE)
```

All expected frequencies $\geq 5$

**Step 4 — Test statistic:**

$$\chi^2 = \frac{(45-35)^2}{35} + \frac{(55-65)^2}{65} +
           \frac{(25-35)^2}{35} + \frac{(75-65)^2}{65}$$

$$= \frac{100}{35} + \frac{100}{65} + \frac{100}{35} + \frac{100}{65}
= 2.857 + 1.538 + 2.857 + 1.538 = \mathbf{8.791}$$

$df = (2-1)(2-1) = 1$

**Step 5 — Decision:**

From the chi-square table: $\chi^2_{\text{critical}} = 3.841$ at $df=1$, $\alpha=0.05$.

Since $8.791 > 3.841$, we **reject $H_0$**.

The p-value $= 0.003 < 0.05$.

**Conclusion:** There is a statistically significant association between
smoking status and hypertension ($\chi^2 = 8.79$, $df = 1$, $p = 0.003$).
Smokers have a higher prevalence of hypertension than non-smokers.
:::

---

## Example 2: Independent Samples t-Test

A clinical trial compares the mean reduction in systolic blood pressure (mmHg)
between patients receiving a **new drug** vs. a **placebo** after 8 weeks.

| Group | n | Mean reduction | SD |
|---|:---:|:---:|:---:|
| New drug | 40 | 12.4 | 6.8 |
| Placebo | 40 | 7.1 | 7.2 |

Test at $\alpha = 0.05$ whether the new drug produces a greater mean
reduction in SBP than placebo.

::: {.callout-tip icon=false}
## Solution

**Step 1 — Hypotheses:**

$H_0: \mu_1 = \mu_2$ (no difference in mean reduction)

$H_1: \mu_1 \neq \mu_2$ (two-tailed; difference exists)

**Step 2 —** $\alpha = 0.05$

**Step 3 — Pooled standard error** (assuming equal variances):

$$s_p = \sqrt{\frac{(40-1)(6.8)^2 + (40-1)(7.2)^2}{40+40-2}}
= \sqrt{\frac{39(46.24) + 39(51.84)}{78}}
= \sqrt{\frac{1803.36 + 2021.76}{78}}
= \sqrt{49.041} = 7.003$$

$$SE = 7.003 \times \sqrt{\frac{1}{40}+\frac{1}{40}}
= 7.003 \times \sqrt{0.05}
= 7.003 \times 0.2236 = 1.566$$

**Step 4 — Test statistic:**

$$t = \frac{12.4 - 7.1}{1.566} = \frac{5.3}{1.566} = \mathbf{3.384}$$

$df = 40 + 40 - 2 = 78$

**Step 5 — Decision:**

$t_{\text{critical}} = 1.991$ at $df = 78$, $\alpha = 0.05$ (two-tailed).

Since $3.384 > 1.991$, we **reject $H_0$**.

The p-value $= 0.001 < 0.05$.

**Conclusion:** The new drug produces a significantly greater reduction
in systolic blood pressure compared to placebo
($t = 3.38$, $df = 78$, $p = 0.001$).
The mean difference is 5.3 mmHg (95% CI: 2.2 to 8.4 mmHg).
:::

---

## Example 3: Paired Samples t-Test

A study measures the **diastolic blood pressure (DBP)** of 10 patients
**before** and **after** 4 weeks of a lifestyle intervention.

```{r}
#| label: tbl-paired
#| tbl-cap: "DBP Before and After Lifestyle Intervention (n = 10 patients)"

set.seed(7)
before <- c(92, 88, 95, 100, 85, 97, 90, 94, 88, 96)
after  <- c(85, 82, 89, 91,  80, 88, 84, 87, 83, 88)
d      <- before - after

pair_tab <- data.frame(
  Patient = paste0("Patient ", 1:10),
  Before  = before,
  After   = after,
  d       = d
)

kable(pair_tab,
      col.names = c("Patient", "DBP Before (mmHg)",
                    "DBP After (mmHg)", "Difference d = Before - After"),
      align = c("l","c","c","c")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = 13) |>
  column_spec(4, bold = TRUE, color = "#2c7bb6") |>
  row_spec(0, bold = TRUE, color = "white", background = "#2c7bb6")
```

Test at $\alpha = 0.05$ whether the intervention significantly reduces DBP.

::: {.callout-tip icon=false}
## Solution

**Step 1 — Hypotheses:**

$H_0: \mu_d = 0$ (no mean change in DBP)

$H_1: \mu_d \neq 0$ (mean DBP has changed)

**Step 2 —** $\alpha = 0.05$

**Step 3 — Compute $\bar{d}$ and $s_d$:**

$$\bar{d} = \frac{`r sum(d)`}{10} = `r mean(d)` \text{ mmHg}$$

$$s_d = \sqrt{\frac{\sum(d_i - \bar{d})^2}{n-1}} = `r round(sd(d), 3)`$$

**Step 4 — Test statistic:**

$$t = \frac{\bar{d}}{s_d/\sqrt{n}} = \frac{`r mean(d)`}{`r round(sd(d),3)`/\sqrt{10}}
= \frac{`r mean(d)`}{`r round(sd(d)/sqrt(10),3)`} = `r round(mean(d)/(sd(d)/sqrt(10)),3)`$$

$df = n - 1 = 9$

**Step 5 — Decision:**

$t_{\text{critical}} = 2.262$ at $df = 9$, $\alpha = 0.05$ (two-tailed).

Since $`r round(mean(d)/(sd(d)/sqrt(10)),3)` > 2.262$, we **reject $H_0$**.

**Conclusion:** The lifestyle intervention produced a statistically
significant reduction in diastolic blood pressure
($t = `r round(mean(d)/(sd(d)/sqrt(10)),2)`$, $df = 9$, $p < 0.05$).
The mean reduction was **`r mean(d)` mmHg**.
:::

---

# Exercises

::: {.callout-caution icon=false}
## Exercise 1 — Chi-Square Test

A researcher investigates whether **physical activity level** is associated
with **diabetes status** in 300 adults:

|  | Diabetes | No Diabetes | Total |
|---|:---:|:---:|:---:|
| Active | 18 | 132 | 150 |
| Inactive | 42 | 108 | 150 |
| **Total** | **60** | **240** | **300** |

**(a)** State $H_0$ and $H_1$.

**(b)** Calculate all expected frequencies and verify the chi-square
condition is met.

**(c)** Calculate the chi-square test statistic.

**(d)** At $\alpha = 0.05$, what is your conclusion?

**(e)** In plain language, what does your result mean?
:::

::: {.callout-caution icon=false}
## Exercise 2 — Independent Samples t-Test

A study compares the **mean hemoglobin levels (g/dL)** between
male and female patients:

| Group | n | $\bar{x}$ | s |
|---|:---:|:---:|:---:|
| Male | 35 | 14.8 | 1.6 |
| Female | 35 | 12.9 | 1.4 |

**(a)** State $H_0$ and $H_1$ (two-tailed).

**(b)** Calculate the pooled standard error.

**(c)** Calculate the t test statistic.

**(d)** At $\alpha = 0.05$ with appropriate df, state your conclusion.
:::

::: {.callout-caution icon=false}
## Exercise 3 — Paired Samples t-Test

A pain management study measures the **pain score (0–10)** of 8 patients
**before** and **one hour after** receiving an analgesic:

| Patient | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 |
|---|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| Before | 7 | 8 | 6 | 9 | 7 | 8 | 5 | 9 |
| After  | 4 | 5 | 4 | 6 | 3 | 5 | 3 | 6 |

**(a)** Calculate the difference $d_i$ for each patient.

**(b)** Calculate $\bar{d}$ and $s_d$.

**(c)** Calculate the paired t statistic.

**(d)** At $\alpha = 0.05$, is the analgesic effective in reducing pain?
State your conclusion clearly.
:::

::: {.callout-caution icon=false}
## Exercise 4 — Choosing the Right Test

For each scenario below, state which statistical test is most appropriate
and briefly explain why:

**(a)** Comparing the proportion of post-operative complications between
patients who received standard care vs. enhanced recovery protocol.

**(b)** Comparing patient satisfaction scores (rated: Poor / Fair / Good /
Excellent) between two hospital wards.

**(c)** Measuring the effect of a new diet on body weight by recording
each participant's weight before and after a 12-week program.

**(d)** Comparing the mean length of hospital stay (days) between two
age groups (under 60 vs. 60 and over), where length-of-stay data is
strongly right-skewed.

**(e)** Comparing pain scores (rated 0--10) recorded for the **same**
patients before and 24 hours after starting a new analgesic, where the
differences are clearly not normally distributed.
:::

---

*End of Chapter 5*

---

::: {.callout-note icon=false}
## Coming Up Next
**Chapter 6: Introduction to Statistical Modeling** — We take our first
steps beyond comparing two groups, exploring how one variable can be used
to explain or predict another through **linear regression**.
:::
