# Data in Health Science

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(knitr)
library(kableExtra)
```

---

# Introduction {.unnumbered}

::: {.callout-note icon=false}
## Why Does This Matter for Nurses?

Every day, nurses make decisions based on data — patient vital signs, lab results,
medication dosages, and clinical outcomes. Understanding **what data is** and
**how to classify it correctly** is the very first step toward making
safe, evidence-based decisions in clinical practice.

In this chapter, we lay the foundation for everything that follows in this course.
:::

---

# Data Types

## What is Data?

**Data** are pieces of information collected from observations or measurements.
In health science, data come from patients, experiments, surveys, or medical records.

> **Example:** A nurse records the blood pressure, age, gender, pain score,
> and discharge status of every patient on the ward. Each of these recorded
> pieces of information is *data*.

Data can be broadly divided into two major types:

- **Categorical Data** — data that represent categories or groups
- **Numerical Data** — data that represent measurable quantities

---

## Categorical Data

Categorical data place individuals into groups. There are two subtypes:

### Nominal Data

Nominal data have **no natural order** among categories.
You can only say things are *different*, not *bigger* or *smaller*.

::: {.callout-tip icon=false}
## Nominal Examples in Nursing
- Blood type: A, B, AB, O
- Gender: Male, Female
- Disease diagnosis: Diabetes, Hypertension, Asthma
- Ward type: Medical, Surgical, Pediatric, ICU
:::

### Ordinal Data

Ordinal data have **a natural order**, but the *distance* between categories
is not necessarily equal.

::: {.callout-tip icon=false}
## Ordinal Examples in Nursing
- Pain scale: None (0), Mild (1–3), Moderate (4–6), Severe (7–10)
- Patient satisfaction: Very dissatisfied Dissatisfied Neutral Satisfied Very satisfied
- Level of consciousness: Alert Confused Drowsy Unresponsive
- Education level: Primary Secondary Bachelor Graduate
:::

---

## Numerical Data

Numerical data represent actual measured or counted quantities. There are two subtypes:

### Discrete Data

Discrete data are **counted** and can only take **whole number** values.
There are no meaningful values between two consecutive counts.

::: {.callout-tip icon=false}
## Discrete Examples in Nursing
- Number of patients admitted per day: 0, 1, 2, 3, ...
- Number of medication errors per month
- Number of children a patient has
- Number of hospital readmissions
:::

### Continuous Data

Continuous data are **measured** and can take **any value** within a range,
including decimals.

::: {.callout-tip icon=false}
## Continuous Examples in Nursing
- Body temperature: 36.5°C, 37.2°C, 38.8°C
- Blood pressure: 120/80 mmHg, 135/85 mmHg
- Body weight: 58.3 kg, 72.5 kg
- Blood glucose level: 95.4 mg/dL
:::

---

## Summary: Data Type Classification

```{r}
#| label: fig-data-types
#| fig-cap: "Classification of Data Types in Health Science"
#| fig-align: center
#| fig-height: 4
#| fig-width: 8

# Simple diagram using ggplot
library(ggplot2)

df <- data.frame(
  x =    c(5,   2,   8,   1.2, 2.8,  6.5, 9.5),
  y =    c(4,   2.5, 2.5, 1,   1,    1,   1),
  label = c("Data",
            "Categorical",
            "Numerical",
            "Nominal",
            "Ordinal",
            "Discrete",
            "Continuous"),
  size = c(6, 5, 5, 4, 4, 4, 4)
)

arrows_df <- data.frame(
  x =    c(5,   5,   2,   2,   8,   8),
  xend = c(2,   8,   1.2, 2.8, 6.5, 9.5),
  y =    c(3.7, 3.7, 2.2, 2.2, 2.2, 2.2),
  yend = c(2.8, 2.8, 1.3, 1.3, 1.3, 1.3)
)

ggplot() +
  geom_segment(data = arrows_df,
               aes(x = x, xend = xend, y = y, yend = yend),
               arrow = arrow(length = unit(0.2, "cm")),
               color = "gray50", linewidth = 0.8) +
  geom_label(data = df,
             aes(x = x, y = y, label = label),
             fill = c("#2c7bb6","#abd9e9","#abd9e9",
                      "#d7191c","#fdae61","#1a9641","#a6d96a"),
             color = "white",
             fontface = "bold",
             size = df$size - 1,
             label.r = unit(0.4, "lines"),
             label.padding = unit(0.5, "lines")) +
  xlim(0, 11) + ylim(0.3, 4.8) +
  theme_void()
```

---

# Population and Sample

## Population

A **population** is the **complete set** of all individuals or observations
that we are interested in studying.

In health research, populations are usually very large and it is
**impractical or impossible** to collect data from everyone.

> **Example:** All patients with Type 2 diabetes in Thailand.

## Sample

A **sample** is a **subset** of the population that is actually studied.
We use the sample to draw conclusions (*make inferences*) about the
whole population.

> **Example:** 200 patients with Type 2 diabetes selected from 5 hospitals
> in Bangkok.

::: {.callout-important icon=false}
## Key Principle
A good sample must be **representative** of the population —
meaning it should reflect the characteristics of the whole group,
not just one part of it.
:::

---

## Key Terms

```{r}
#| label: tbl-terms
#| tbl-cap: "Important Statistical Terms"

terms <- data.frame(
  Term = c("Population",
           "Sample",
           "Parameter",
           "Statistic",
           "Variable",
           "Observation"),
  Definition = c(
    "The entire group of interest",
    "A selected subset of the population",
    "A numerical summary describing the population (e.g., population mean μ)",
    "A numerical summary calculated from the sample (e.g., sample mean x̄)",
    "A characteristic being measured or observed (e.g., age, blood pressure)",
    "A single recorded value for one individual"
  ),
  `Health Example` = c(
    "All nurses working in Thailand",
    "500 nurses randomly selected from 10 hospitals",
    "The true average age of ALL nurses in Thailand",
    "The average age calculated from our 500 nurses",
    "Age, gender, years of experience, ward type",
    "Nurse A: age = 28, female, 3 years experience"
  )
)

kable(terms, col.names = c("Term", "Definition", "Health Science Example"),
      align = c("l","l","l")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = TRUE, font_size = 13) |>
  column_spec(1, bold = TRUE, color = "#2c7bb6")
```

---

## Population vs. Sample: Visual Concept

```{r}
#| label: fig-pop-sample
#| fig-cap: "The sample is drawn from the population and used to make inferences"
#| fig-align: center
#| fig-height: 4
#| fig-width: 7

set.seed(42)
n_pop <- 200
pop <- data.frame(
  x = runif(n_pop, 0, 10),
  y = runif(n_pop, 0, 10),
  group = "Population"
)

sample_idx <- sample(1:n_pop, 30)
pop$group[sample_idx] <- "Sample"

ggplot(pop, aes(x = x, y = y, color = group, size = group, shape = group)) +
  geom_point(alpha = 0.8) +
  scale_color_manual(values = c("Population" = "#b0c4de", "Sample" = "#d7191c")) +
  scale_size_manual(values = c("Population" = 2, "Sample" = 4)) +
  scale_shape_manual(values = c("Population" = 16, "Sample" = 17)) +
  annotate("text", x = 5, y = 10.5,
           label = "Population (All Patients)",
           size = 4.5, fontface = "bold", color = "#2c7bb6") +
  annotate("text", x = 5, y = -0.8,
 label = " Red triangles = selected sample (n = 30)",
           size = 3.5, color = "#d7191c") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        panel.border = element_rect(color = "#2c7bb6", fill = NA, linewidth = 1.5)) +
  xlim(-0.5, 10.5) + ylim(-1.2, 11)
```

---

# Key Takeaways

::: {.callout-note icon=false}
## Chapter 1 Summary

1. **Data** are recorded pieces of information collected from patients or research subjects.

2. Data are classified into two main types:
   - **Categorical**: Nominal (no order) and Ordinal (has order)
   - **Numerical**: Discrete (counted, whole numbers) and Continuous (measured, any value)

3. Identifying the **correct data type** is critical — it determines which
   statistical methods are appropriate.

4. A **population** is the entire group of interest; a **sample** is the
   subset we actually study.

5. We use **statistics** calculated from samples to estimate **parameters**
   of the population — this is the core idea behind all of biostatistics.
:::

---

# Examples

## Example 1: Classify Each Variable

A nurse researcher collects the following information from 50 patients
admitted to a medical ward:

| Variable | Recorded Values |
|---|---|
| Patient ID | 001, 002, 003, ... |
| Age | 45, 62, 38, ... (years) |
| Gender | Male, Female |
| Blood type | A, B, AB, O |
| Systolic blood pressure | 118, 135, 142, ... (mmHg) |
| Pain score | 0–10 scale |
| Number of medications | 1, 2, 3, ... |
| Discharge status | Recovered, Transferred, Deceased |

::: {.callout-tip icon=false}
## Solution

| Variable | Data Type | Reason |
|---|---|---|
| Patient ID | — (Identifier, not analyzed) | Just a label |
| Age | **Numerical – Continuous** | Measured, can be 45.5 years |
| Gender | **Categorical – Nominal** | Two categories, no order |
| Blood type | **Categorical – Nominal** | Four categories, no order |
| Systolic BP | **Numerical – Continuous** | Measured in mmHg, any value |
| Pain score | **Categorical – Ordinal** | Has order (0 < 1 < ... < 10), but gaps unequal |
| No. of medications | **Numerical – Discrete** | Counted, whole numbers only |
| Discharge status | **Categorical – Nominal** | Three categories, no natural order |
:::

---

## Example 2: Population or Sample?

For each situation below, identify whether the group described is a
**population** or a **sample**:

**(a)** A hospital wants to know the average job satisfaction of all 320 nurses
working there. The HR department surveys all 320 nurses.

**(b)** A researcher wants to study pain levels in post-operative patients
across Thailand. She recruits 150 patients from 3 hospitals in Chiang Mai.

**(c)** A nursing student studies the blood glucose levels of the 8 diabetic
patients currently assigned to her ward.

::: {.callout-tip icon=false}
## Solution

**(a)** This is the **Population** — all 320 nurses are surveyed;
no one is left out. The result is a *parameter*.

**(b)** This is a **Sample** — 150 patients from 3 hospitals cannot
represent all post-operative patients in Thailand.
The result is a *statistic* used to estimate the population parameter.

**(c)** This is a **Sample** — the 8 patients on the ward are a
small subset of all diabetic patients.
:::

---

# Exercises

::: {.callout-caution icon=false}
## Exercise 1

A public health nurse collects the following data from patients visiting
a community health clinic:

- Weight (kg)
- Presence of hypertension (Yes / No)
- Severity of hypertension (Mild / Moderate / Severe)
- Number of clinic visits in the past year
- Ethnicity
- Body temperature (°C)
- Diastolic blood pressure (mmHg)

**Classify each variable** into the correct data type and explain your reasoning.
:::

::: {.callout-caution icon=false}
## Exercise 2

A researcher is studying the prevalence of depression among elderly patients
in nursing homes across the country. She selects 200 patients from 4 nursing
homes in the central region.

**(a)** What is the population in this study?

**(b)** What is the sample?

**(c)** If the average depression score of the 200 patients is 14.2,
is this number a *parameter* or a *statistic*? Explain.
:::

::: {.callout-caution icon=false}
## Exercise 3

Look at the following list and decide whether each is an example of
a **discrete** or **continuous** variable:

1. The number of patients who fell during a hospital shift
2. The time (in minutes) a patient waited before being seen by a doctor
3. The number of syringes used during a procedure
4. A patient's oxygen saturation level (SpO₂ %)
5. The number of newborns delivered in a hospital per week
:::

---

*End of Chapter 1*

---

::: {.callout-note icon=false}
## Coming Up Next
**Chapter 2: Data Summarization** — How do we make sense of a large set of
patient data? We will learn to calculate means, medians, standard deviations,
and draw charts that tell a clear story.
:::
