# Data Summarization

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(knitr)
library(kableExtra)
library(dplyr)
```

---

# Introduction {.unnumbered}

::: {.callout-note icon=false}
## Why Does This Matter for Nurses?

Imagine you have blood pressure readings from 80 patients on your ward.
Looking at 80 individual numbers tells you very little. But if someone says
*"the average systolic BP is 138 mmHg and most patients fall between
120 and 155 mmHg"* — now you have a clear picture.

**Data summarization** is the process of turning raw data into meaningful,
easy-to-understand information using numbers and charts. This is one of the
most practical skills in clinical nursing and health research.
:::

---

# Numerical Summarization

Numerical summarization uses **numbers** to describe the key features of data.
There are two important aspects we want to describe:

1. **Central tendency** — Where is the "middle" or "typical" value?
2. **Variability** — How spread out are the values?

---

## Measures of Central Tendency

### Mean (Arithmetic Average)

The **mean** is calculated by adding all values and dividing by the number of observations.

$$\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n} = \frac{x_1 + x_2 + \cdots + x_n}{n}$$

Where:

- $\bar{x}$ = sample mean (read as "x-bar")
- $x_i$ = each individual value
- $n$ = total number of observations
- $\sum$ = "sum of all"

::: {.callout-warning icon=false}
## When to Use the Mean
The mean works best for **continuous or discrete numerical data**
that is **roughly symmetric** (not heavily skewed) and has **no extreme outliers**.
A single extreme value can pull the mean far from the typical value.
:::

---

### Median

The **median** is the **middle value** when all observations are arranged in order.

- If $n$ is **odd**: the median is the value at position $\dfrac{n+1}{2}$
- If $n$ is **even**: the median is the **average of the two middle values**
  at positions $\dfrac{n}{2}$ and $\dfrac{n}{2}+1$

::: {.callout-warning icon=false}
## When to Use the Median
The median is preferred when data is **skewed** or has **outliers**,
because it is not affected by extreme values.
It is also the appropriate measure for **ordinal data**.
:::

---

### Mode

The **mode** is the value that appears **most frequently** in the data.

- A dataset can have **one mode** (unimodal), **two modes** (bimodal),
  or **no mode** (all values appear equally).
- The mode is the only measure of central tendency appropriate for **nominal data**.

---

### Choosing the Right Measure

```{r}
#| label: tbl-central
#| tbl-cap: "Choosing the Right Measure of Central Tendency"

ct <- data.frame(
  Measure = c("Mean", "Median", "Mode"),
  `Best For` = c(
    "Numerical data (continuous or discrete), symmetric distribution",
    "Numerical data with skewness or outliers; Ordinal data",
    "Nominal data; finding the most common category"
  ),
  `Nursing Example` = c(
    "Average patient age, average blood glucose",
    "Average length of hospital stay (skewed by long-stay patients)",
    "Most common blood type on a ward, most frequent pain score"
  )
)

kable(ct, col.names = c("Measure", "Best For", "Nursing Example"),
      align = c("l","l","l")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = 13) |>
  column_spec(1, bold = TRUE, color = "#2c7bb6")
```

---

## Measures of Variability

Knowing the average alone is not enough. Two wards might have the same average
patient age but very different spreads — one ward may have all patients aged
55–65, another may have patients aged 20–90. Variability tells us how **spread
out** the data are.

### Range

The **range** is the simplest measure of spread:

$$\text{Range} = \text{Maximum value} - \text{Minimum value}$$

::: {.callout-warning icon=false}
## Limitation
The range only uses two values (max and min) and is very sensitive to outliers.
:::

---

### Variance

The **variance** measures the average squared distance of each value from the mean.

**Population variance:**
$$\sigma^2 = \frac{\sum_{i=1}^{N}(x_i - \mu)^2}{N}$$

**Sample variance** (used more often in practice):
$$s^2 = \frac{\sum_{i=1}^{n}(x_i - \bar{x})^2}{n-1}$$

Where:

- $\mu$ = population mean, $\bar{x}$ = sample mean
- $N$ = population size, $n$ = sample size
- $n-1$ in the denominator corrects for the fact that we're using a sample

---

### Standard Deviation

The **standard deviation (SD)** is the square root of the variance.
It is expressed in the **same units as the original data**, making it
much easier to interpret than variance.

$$s = \sqrt{s^2} = \sqrt{\frac{\sum_{i=1}^{n}(x_i - \bar{x})^2}{n-1}}$$

::: {.callout-tip icon=false}
## Interpreting Standard Deviation

A **small SD** means values are clustered closely around the mean.
A **large SD** means values are widely spread.

In clinical practice, you may see lab reports written as:
**Mean ± SD**, for example: Hemoglobin = **12.5 ± 1.8 g/dL**
:::

---

### Summary: Numerical Measures

```{r}
#| label: tbl-variability
#| tbl-cap: "Summary of Numerical Summary Measures"

vm <- data.frame(
  Measure = c("Mean", "Median", "Mode", "Range", "Variance", "Standard Deviation"),
  Symbol = c("x̄", "Md", "Mo", "Max − Min", "s²", "s"),
  Purpose = c("Central tendency", "Central tendency",
               "Central tendency", "Spread",
               "Spread (squared units)", "Spread (original units)"),
  `Data Type` = c("Numerical", "Numerical / Ordinal",
                   "Any", "Numerical",
                   "Numerical", "Numerical")
)

kable(vm, col.names = c("Measure","Symbol","Purpose","Appropriate Data Type"),
      align = c("l","c","l","l")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = 13) |>
  column_spec(1, bold = TRUE, color = "#2c7bb6")
```

---

# Pictorial Summarization

Charts and graphs turn numbers into visual stories. The correct chart depends
on the **data type**.

## Bar Chart

Used for **categorical data** (nominal or ordinal).
Each bar represents a category; the height shows the count or percentage.

```{r}
#| label: fig-bar
#| fig-cap: "Bar Chart: Blood Types of 120 Patients in a Medical Ward"
#| fig-align: center
#| fig-height: 4
#| fig-width: 6

blood <- data.frame(
  type = c("A", "B", "AB", "O"),
  count = c(38, 27, 15, 40)
)

ggplot(blood, aes(x = type, y = count, fill = type)) +
  geom_bar(stat = "identity", width = 0.6, color = "white") +
  geom_text(aes(label = count), vjust = -0.5, size = 4.5, fontface = "bold") +
  scale_fill_manual(values = c("#2c7bb6","#abd9e9","#d7191c","#fdae61")) +
  labs(x = "Blood Type", y = "Number of Patients") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank())
```

---

## Histogram

Used for **continuous numerical data**.
Bars are drawn over class intervals (ranges of values); bars touch each other
because the data is continuous.

```{r}
#| label: fig-hist
#| fig-cap: "Histogram: Systolic Blood Pressure of 80 Patients"
#| fig-align: center
#| fig-height: 4
#| fig-width: 6

set.seed(123)
bp_data <- data.frame(sbp = round(rnorm(80, mean = 130, sd = 15)))

ggplot(bp_data, aes(x = sbp)) +
  geom_histogram(binwidth = 10, fill = "#2c7bb6", color = "white", alpha = 0.85) +
  labs(x = "Systolic Blood Pressure (mmHg)", y = "Number of Patients") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

::: {.callout-tip icon=false}
## Bar Chart vs Histogram
| | Bar Chart | Histogram |
|---|---|---|
| Data type | Categorical | Continuous numerical |
| Bars | Separated (gaps) | Touching (no gaps) |
| X-axis | Categories | Numerical ranges |
:::

---

## Box Plot (Box-and-Whisker Plot)

A **box plot** summarizes the **distribution** of numerical data using
5 key values (the **five-number summary**):

| Value | Description |
|---|---|
| Minimum | Smallest value (excluding outliers) |
| Q1 | 25th percentile (lower quartile) |
| Median (Q2) | 50th percentile — middle value |
| Q3 | 75th percentile (upper quartile) |
| Maximum | Largest value (excluding outliers) |

The **Interquartile Range (IQR) = Q3 − Q1** represents the middle 50% of data.
Points beyond **1.5 × IQR** from Q1 or Q3 are marked as **outliers ()**.

```{r}
#| label: fig-boxplot
#| fig-cap: "Box Plot: Comparing Length of Hospital Stay Between Two Wards"
#| fig-align: center
#| fig-height: 4
#| fig-width: 6

set.seed(99)
stay <- data.frame(
  days = c(rexp(50, 0.3), rexp(50, 0.5)),
  ward = rep(c("Medical Ward", "Surgical Ward"), each = 50)
)

ggplot(stay, aes(x = ward, y = days, fill = ward)) +
  geom_boxplot(width = 0.5, outlier.color = "#d7191c",
               outlier.shape = 1, outlier.size = 3, alpha = 0.8) +
  scale_fill_manual(values = c("#abd9e9","#fdae61")) +
  labs(x = "", y = "Length of Stay (days)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank())
```

---

## Choosing the Right Chart

```{r}
#| label: tbl-charts
#| tbl-cap: "Guide to Choosing the Right Chart"

charts <- data.frame(
  `Chart Type` = c("Bar Chart", "Pie Chart", "Histogram", "Box Plot"),
  `Data Type` = c("Categorical (Nominal/Ordinal)",
                   "Categorical (Nominal) — proportions",
                   "Numerical (Continuous)",
                   "Numerical (Continuous/Discrete)"),
  `What It Shows` = c(
    "Frequency or percentage of each category",
    "Proportion of each category as a slice of the whole",
    "Distribution shape, spread, and central tendency",
    "Median, spread, quartiles, and outliers"
  ),
  `Nursing Example` = c(
    "Number of patients by blood type",
    "Proportion of patients by diagnosis",
    "Distribution of patient ages or blood pressure",
    "Comparing recovery time between two wards"
  )
)

kable(charts,
      col.names = c("Chart Type","Data Type","What It Shows","Nursing Example"),
      align = c("l","l","l","l")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = 13) |>
  column_spec(1, bold = TRUE, color = "#2c7bb6")
```

---

# Key Takeaways

::: {.callout-note icon=false}
## Chapter 2 Summary

1. **Numerical summarization** uses measures of *central tendency*
   (mean, median, mode) and *variability* (range, variance, SD) to
   describe data with numbers.

2. The **mean** is best for symmetric numerical data;
   the **median** is better when data is skewed or has outliers;
   the **mode** is used for categorical data.

3. The **standard deviation (SD)** is the most commonly reported measure
   of spread in clinical research, expressed in the same units as the data.

4. **Pictorial summarization** uses charts: bar charts for categorical data,
   histograms for continuous data, and box plots to compare groups and
   identify outliers.

5. Always choose the **correct summary measure and chart** based on the
   **data type** — using the wrong one leads to misleading conclusions.
:::

---

# Examples

## Example 1: Calculating Mean, Median, and SD

A nurse records the resting heart rate (beats per minute) of **9 patients**:

$$72, \; 68, \; 85, \; 74, \; 90, \; 68, \; 78, \; 72, \; 110$$

Calculate the **(a) mean**, **(b) median**, and **(c) standard deviation**.

::: {.callout-tip icon=false}
## Solution

**Step 1 — Arrange data in order:**

$$68, \; 68, \; 72, \; 72, \; 74, \; 78, \; 85, \; 90, \; 110$$

---

**(a) Mean:**

$$\bar{x} = \frac{68+68+72+72+74+78+85+90+110}{9} = \frac{717}{9} = 79.7 \text{ bpm}$$

---

**(b) Median:**

$n = 9$ (odd), so median is at position $\dfrac{9+1}{2} = 5$th value.

Counting from the ordered list: the 5th value = **74 bpm**

Notice that the median (74) is lower than the mean (79.7) because the
outlier value of 110 pulls the mean upward.

---

**(c) Standard Deviation:**

First, find each $(x_i - \bar{x})^2$:

```{r}
#| label: tbl-sd-calc
#| tbl-cap: "Step-by-step SD calculation"

hr <- c(68, 68, 72, 72, 74, 78, 85, 90, 110)
xbar <- mean(hr)

sd_table <- data.frame(
  `Patient` = paste0("Patient ", 1:9),
  `Heart Rate (x)` = hr,
  `x − x̄` = round(hr - xbar, 1),
  `(x − x̄)²` = round((hr - xbar)^2, 2)
)

kable(sd_table,
      col.names = c("Patient", "Heart Rate (x)", "x − x̄", "(x − x̄)²"),
      align = c("l","c","c","c")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = 13) |>
  row_spec(9, bold = TRUE) |>
  row_spec(0, bold = TRUE, color = "white", background = "#2c7bb6")
```

$$\sum(x_i - \bar{x})^2 = `r round(sum((hr - xbar)^2), 2)`$$

$$s^2 = \frac{`r round(sum((hr-xbar)^2),2)`}{9-1} = \frac{`r round(sum((hr-xbar)^2),2)`}{8} = `r round(var(hr),2)`$$

$$s = \sqrt{`r round(var(hr),2)`} = `r round(sd(hr),2)` \text{ bpm}$$

**Interpretation:** The average heart rate is **79.7 ± 13.5 bpm**.
The relatively large SD reflects the wide spread, largely influenced
by the patient with 110 bpm.
:::

---

## Example 2: Choosing the Right Summary and Chart

A hospital administrator records the **diagnosis** of 200 patients admitted
last month and wants to summarize the data. A research nurse records the
**length of stay (days)** for the same 200 patients.

::: {.callout-tip icon=false}
## Solution

| Variable | Data Type | Best Summary Measure | Best Chart |
|---|---|---|---|
| Diagnosis | Categorical – Nominal | Mode (most common diagnosis) | Bar Chart |
| Length of stay | Numerical – Continuous (likely skewed) | Median + IQR | Box Plot or Histogram |

**Why median for length of stay?**
Hospital stays are almost always right-skewed — most patients stay a few days,
but a small number may stay for weeks or months. These long-stay patients
would pull the mean upward, making it misleading. The median better
represents the "typical" patient.
:::

---

# Exercises

::: {.callout-caution icon=false}
## Exercise 1

A nurse records the body temperature (°C) of **10 patients** at 8:00 AM:

$$37.2, \; 36.8, \; 38.5, \; 37.0, \; 39.1, \; 36.8, \; 37.5, \; 38.0, \; 37.2, \; 40.2$$

**(a)** Calculate the **mean** body temperature.

**(b)** Find the **median** body temperature.

**(c)** What is the **mode**?

**(d)** Calculate the **range**.

**(e)** Calculate the **standard deviation**. Show all steps.

**(f)** One patient has a temperature of 40.2°C. How does this value
affect the mean compared to the median? Which measure would you
report to the physician, and why?
:::

::: {.callout-caution icon=false}
## Exercise 2

A community health nurse surveys **50 elderly patients** and records:

- **Gender** (Male / Female)
- **Systolic blood pressure** (mmHg)
- **Self-rated health** (Poor / Fair / Good / Very Good / Excellent)

For each variable, state:

**(a)** The data type.

**(b)** The most appropriate measure of central tendency.

**(c)** The most appropriate chart to display the data.
:::

::: {.callout-caution icon=false}
## Exercise 3

The number of patient falls recorded each month over **12 months**
in a hospital ward are:

$$2, \; 0, \; 3, \; 1, \; 4, \; 2, \; 1, \; 0, \; 3, \; 2, \; 5, \; 1$$

**(a)** Calculate the mean and median number of falls per month.

**(b)** Draw a **bar chart** (by hand or describe what it would look like)
showing the frequency of each number of falls.

**(c)** The ward manager wants to set a monthly target.
Should she use the mean or median as the baseline? Explain.
:::

---

*End of Chapter 2*

---

::: {.callout-note icon=false}
## Coming Up Next
**Chapter 3: Random Variable and Probability Distribution** —
We will learn what a random variable is, how probabilities work,
and why the **normal distribution** (bell curve) is so important
in health science research.
:::
