# Estimation

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(knitr)
library(kableExtra)
library(patchwork)
# Table helpers (used by the book and the slides)
# tbl_fs(): font size; PDF tables use 10pt so they fit the page width
# pdf_w():  fixed column width in the PDF only, so long text wraps
tbl_fs <- function(html_size = 13) if (knitr::is_latex_output()) 10 else html_size
pdf_w  <- function(x) if (knitr::is_latex_output()) x else NULL
```

---

# Introduction {.unnumbered}

::: {.callout-note icon=false}
## Why Does This Matter?

In practice, we almost never have access to an entire population.
We collect data from a **sample** and use it to draw conclusions about
the **population** — a process called **statistical inference**.

**Estimation** is one of the two main branches of statistical inference
(the other being hypothesis testing, covered in Chapter 5).

In this chapter, we answer questions like:

- The mean serum cholesterol in our sample is 198 mg/dL —
  what can we say about the mean for the *entire* population?
- A survey found that 34% of respondents have hypertension —
  what is a reasonable range for the true population proportion?

The key tool for answering these questions is the **confidence interval**.
:::

---

# Concept of Estimation

## Point Estimate vs. Interval Estimate

There are two ways to estimate a population parameter from sample data:

### Point Estimate

A **point estimate** is a **single value** calculated from the sample
that serves as our best guess for the population parameter.

| Population Parameter | Symbol | Point Estimate | Symbol |
|---|:---:|---|:---:|
| Population mean | $\mu$ | Sample mean | $\bar{x}$ |
| Population proportion | $p$ | Sample proportion | $\hat{p}$ |
| Population SD | $\sigma$ | Sample SD | $s$ |

A point estimate is simple and easy to communicate, but it gives us
**no information about how accurate or reliable** that single number is.

---

### Interval Estimate (Confidence Interval)

An **interval estimate** gives a **range of plausible values** for the
population parameter, along with a stated level of confidence.

$$\text{Point Estimate} \pm \text{Margin of Error}$$

This range is called a **confidence interval (CI)**.

::: {.callout-important icon=false}
## Key Concept: Confidence Level

The **confidence level** (commonly **90%, 95%, or 99%**) tells us:

> *"If we repeated this study many times and calculated a confidence interval
> each time, approximately X% of those intervals would contain the true
> population parameter."*

In health research, **95% confidence** is the most widely used standard.
:::

---

## The Sampling Distribution of the Mean

If we take **one** sample of 40 patients and calculate $\bar{x}$, we get
one number. If a colleague takes a **different** sample of 40 patients from
the same population, she will get a slightly different $\bar{x}$.
The sample mean itself varies from sample to sample.

Imagine repeating this many times and drawing a histogram of all the
sample means. This histogram is called the
**sampling distribution of the mean**. It has three important features:

1. It is **centred at the population mean** $\mu$.
2. Its spread is **smaller** than the spread of individual values:
   its standard deviation is $\sigma / \sqrt{n}$.
3. **Central Limit Theorem:** if the sample size is large enough
   (a common rule of thumb is $n \geq 30$), the sampling distribution of
   $\bar{x}$ is **approximately normal** — *even when the individual
   values are skewed*. If the individual values are already normal, it is
   normal for any $n$.

```{r}
#| label: fig-clt
#| fig-cap: "Left: length of stay of individual patients is right-skewed. Right: the means of many samples of 40 patients are approximately normal."
#| fig-align: center
#| fig-height: 3.5
#| fig-width: 9

set.seed(2024)
los_pop <- rexp(20000, rate = 1/5)            # skewed population, mean 5 days
xbars   <- replicate(2000, mean(sample(los_pop, 40)))

p_ind <- ggplot(data.frame(x = los_pop), aes(x)) +
  geom_histogram(binwidth = 1, boundary = 0, fill = "#fdae61", color = "white") +
  labs(title = "Individual patients", x = "Length of stay (days)", y = "Count") +
  coord_cartesian(xlim = c(0, 30)) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())

p_mean <- ggplot(data.frame(x = xbars), aes(x)) +
  geom_histogram(binwidth = 0.2, fill = "#2c7bb6", color = "white") +
  labs(title = "Means of samples of n = 40", x = "Sample mean (days)", y = "Count") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())

p_ind + p_mean
```

This is why we can use the **normal distribution** (Chapter 3) to build
confidence intervals for a mean, and why the rule "$n \geq 30$" appears
in the conditions below.

## The Role of the Standard Error

The standard deviation of the sampling distribution has its own name:
the **standard error (SE)**. It measures how much the sample estimate
varies from sample to sample. Because we rarely know $\sigma$, we
replace it with the sample SD $s$.

### Standard Error of the Mean

$$SE_{\bar{x}} = \frac{s}{\sqrt{n}}$$

Where $s$ is the sample standard deviation and $n$ is the sample size.

### Standard Error of a Proportion

$$SE_{\hat{p}} = \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

::: {.callout-tip icon=false}
## SE vs. SD — What Is the Difference?

| | Standard Deviation (SD) | Standard Error (SE) |
|---|---|---|
| **Describes** | Variability of individual observations | Variability of the sample mean |
| **Gets smaller with** | More homogeneous data | Larger sample size |
| **Used for** | Describing the spread of data | Constructing confidence intervals |

The SE always gets smaller as $n$ increases — larger samples give
more precise estimates.
:::

```{r}
#| label: fig-se
#| fig-cap: "As sample size increases, the standard error decreases and estimates become more precise"
#| fig-align: center
#| fig-height: 3.5
#| fig-width: 6

n_vals <- seq(10, 500, by = 10)
se_vals <- 15 / sqrt(n_vals)   # SD = 15, typical for SBP

df_se <- data.frame(n = n_vals, se = se_vals)

ggplot(df_se, aes(x = n, y = se)) +
  geom_line(color = "#2c7bb6", linewidth = 1.3) +
  geom_hline(yintercept = c(1, 2), linetype = "dashed",
             color = "#d7191c", alpha = 0.6) +
  annotate("text", x = 480, y = 2.3, label = "SE = 2", color = "#d7191c", size = 3.5) +
  annotate("text", x = 480, y = 1.3, label = "SE = 1", color = "#d7191c", size = 3.5) +
  labs(x = "Sample Size (n)", y = "Standard Error",
       title = "Standard Error vs. Sample Size (SD = 15)") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

---

# Confidence Interval for a Mean

## When the Population SD Is Known (Z-interval)

If the population standard deviation $\sigma$ is known and either
(a) the population is normal, or (b) the sample size is large ($n \geq 30$),
we use the **Z-interval**:

$$\bar{x} \pm z_{1-\alpha/2} \cdot \frac{\sigma}{\sqrt{n}}$$

The **critical value** $z_{1-\alpha/2}$ is written using the **left cumulative
probability** (the area to the *left* of the value, matching the standard
normal table and SPSS `IDF.NORMAL`). It depends on the confidence level:

```{r}
#| label: tbl-z-critical
#| tbl-cap: "Critical Values for Common Confidence Levels"

crit <- data.frame(
  `Confidence Level` = c("90%", "95%", "99%"),
  `alpha` = c("0.10", "0.05", "0.01"),
  `1 - alpha/2` = c("0.950", "0.975", "0.995"),
  `z(1-alpha/2)` = c("1.645", "1.960", "2.576")
)

kable(crit,
      col.names = c("Confidence Level", "alpha", "1 - alpha/2 (left area)", "z(1-alpha/2)"),
      align = c("c","c","c","c")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = tbl_fs(13)) |>
  column_spec(1, bold = TRUE, color = "#2c7bb6") |>
  row_spec(2, background = "#eaf4fb")
```

::: {.callout-tip icon=false}
## The Most Used Value
In health research, the 95% CI with $z_{0.975} = 1.96$ is by far the
most commonly reported. You will see it in almost every clinical paper.
:::

---

## When the Population SD Is Unknown (t-interval)

In practice, $\sigma$ is almost never known. When we use the sample SD $s$
as an estimate, we must use the **t-distribution** instead of the Z-distribution:

$$\bar{x} \pm t_{1-\alpha/2,\, n-1} \cdot \frac{s}{\sqrt{n}}$$

Where $t_{1-\alpha/2,\, n-1}$ is the critical value from the **t-distribution**
with $n - 1$ **degrees of freedom (df)**. As with $z$, the subscript
$1-\alpha/2$ is the **left cumulative probability** (e.g., $0.975$ for a 95% CI).

### The t-Distribution

The t-distribution is similar to the standard normal distribution but has
**heavier tails** — reflecting the extra uncertainty from estimating $\sigma$
with $s$. As $n$ increases, the t-distribution approaches the Z-distribution.

```{r}
#| label: fig-t-dist
#| fig-cap: "t-distributions with different degrees of freedom compared to Z"
#| fig-align: center
#| fig-height: 4
#| fig-width: 7

x <- seq(-4, 4, length = 500)
df_plot <- data.frame(
  x = rep(x, 4),
  y = c(dnorm(x),
        dt(x, df = 3),
        dt(x, df = 10),
        dt(x, df = 30)),
  dist = rep(c("Z (normal)", "t (df=3)", "t (df=10)", "t (df=30)"), each = 500)
)

df_plot$dist <- factor(df_plot$dist,
                       levels = c("t (df=3)","t (df=10)","t (df=30)","Z (normal)"))

ggplot(df_plot, aes(x = x, y = y, color = dist, linetype = dist)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("#d7191c","#fdae61","#abd9e9","#2c7bb6")) +
  scale_linetype_manual(values = c("dashed","dashed","dashed","solid")) +
  labs(x = "Value", y = "Density", color = "Distribution", linetype = "Distribution") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right",
        panel.grid.minor = element_blank())
```

### t Critical Values

```{r}
#| label: tbl-t-critical
#| tbl-cap: "t Critical Values by left cumulative probability (use the 0.975 column for a 95% CI)"

df_vals <- c(1:30, 35, 40, 50, 60, 70, 80, 100, 120, Inf)
probs   <- c(0.95, 0.975, 0.99, 0.995)

t_tab <- data.frame(
  df = c(as.character(df_vals[-length(df_vals)]), "Inf (Z)")
)
for (p in probs) {
  t_tab[[as.character(p)]] <- formatC(qt(p, df = df_vals), format = "f", digits = 3)
}

kable(t_tab,
      col.names = c("df", "0.95", "0.975", "0.99", "0.995"),
      align = rep("c", 5)) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE, font_size = tbl_fs(12)) |>
  add_header_above(c(" " = 1, "Left cumulative probability" = 4)) |>
  column_spec(1, bold = TRUE, color = "#2c7bb6") |>
  column_spec(3, bold = TRUE, background = "#eaf4fb")
```

**How to use the table:** for a 95% CI, use the column **0.975** and the row
$df = n - 1$. For example, with $n = 25$: $t_{0.975,\,24} = 2.064$.

If your df is **not listed**, use the **next smaller** df in the table.
This gives a slightly larger critical value, so the interval is slightly
wider — the safe direction.

As df increases, the t critical value converges to **1.960** (the Z value).
For $n \geq 30$ the difference is small, which is why some textbooks
use Z as an approximation. **SPSS always uses t** when $\sigma$ is unknown,
and so do we in this course.

---

## Interpreting a Confidence Interval

::: {.callout-important icon=false}
## Correct Interpretation

> *"We are 95% confident that the true population mean lies between
> [lower bound] and [upper bound]."*

**What this means:** If we repeated the sampling process many times,
95% of the resulting confidence intervals would contain the true parameter.

**What this does NOT mean:** There is NOT a 95% probability that the
true parameter lies in this specific interval — the true parameter is fixed;
it either is or is not in the interval.
:::

::: {.callout-tip icon=false}
## Using a CI to Judge a Target Value
A 95% CI contains the **plausible** values of the parameter. If a target
or claimed value (for example, a hospital's target readmission rate)
lies **outside** the interval, the data suggest the true value differs
from that target. We will formalise this idea as a hypothesis test in
Chapter 5.
:::

```{r}
#| label: fig-ci-simulation
#| fig-cap: "Simulation: 50 confidence intervals from repeated samples. Green intervals contain the true mean (μ = 130); red ones do not."
#| fig-align: center
#| fig-height: 6
#| fig-width: 7

set.seed(42)
true_mu <- 130
sigma    <- 15
n        <- 40
n_sims   <- 50

sims <- replicate(n_sims, {
  x    <- rnorm(n, mean = true_mu, sd = sigma)
  xbar <- mean(x)
  se   <- sd(x) / sqrt(n)
  c(xbar, xbar - 1.96*se, xbar + 1.96*se)
})

ci_df <- data.frame(
  sim   = 1:n_sims,
  mean  = sims[1,],
  lower = sims[2,],
  upper = sims[3,],
  contains = sims[2,] <= true_mu & sims[3,] >= true_mu
)

ggplot(ci_df, aes(y = sim, color = contains)) +
  geom_segment(aes(x = lower, xend = upper, yend = sim), linewidth = 0.8) +
  geom_point(aes(x = mean), size = 1.5) +
  geom_vline(xintercept = true_mu, color = "black",
             linetype = "dashed", linewidth = 1) +
  scale_color_manual(values = c("TRUE" = "#1a9641", "FALSE" = "#d7191c"),
                     labels = c("TRUE" = "Contains μ", "FALSE" = "Does not contain μ"),
                     name = "") +
  annotate("text", x = true_mu + 0.5, y = 52,
           label = "True μ = 130", hjust = 0, size = 3.5) +
  labs(x = "Value", y = "Sample Number") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
```

The figure above shows 50 confidence intervals from 50 different samples
of the same population. Most (≈95%) contain the true mean; a few do not —
this is expected and not an error.

---

## Example 1: 95% CI for a Mean (σ unknown)

A study measured the fasting blood glucose (mg/dL) of a random sample of
**36 adults** and found:

$$\bar{x} = 105.4 \text{ mg/dL}, \qquad s = 18.6 \text{ mg/dL}$$

Construct a **95% confidence interval** for the population mean blood glucose.

::: {.callout-tip icon=false}
## Solution

**Given:** $n = 36$, $\bar{x} = 105.4$, $s = 18.6$, confidence level = 95%

**Step 1 — Standard Error:**
$$SE = \frac{s}{\sqrt{n}} = \frac{18.6}{\sqrt{36}} = \frac{18.6}{6} = 3.10$$

**Step 2 — Critical value:**

$df = n - 1 = 35$. From the t-table at left cumulative probability
$1 - \alpha/2 = 0.975$: $t_{0.975,\, 35} = 2.030$

*(df = 35 is in the t-table. Since $n = 36$ is fairly large, the value is
already close to $z_{0.975} = 1.96$.)*

**Step 3 — Confidence Interval:**
$$105.4 \pm 2.030 \times 3.10 = 105.4 \pm 6.29$$

$$\textbf{95\% CI: } (99.1, \; 111.7) \text{ mg/dL}$$

**Interpretation:** We are 95% confident that the true mean fasting blood
glucose of adults in this population is between **99.1 and 111.7 mg/dL**.
:::

---

## Example 2: Effect of Sample Size on CI Width

Using the same data as Example 1 ($\bar{x} = 105.4$, $s = 18.6$),
compare the 95% CI when $n = 36$, $n = 100$, and $n = 400$.

::: {.callout-tip icon=false}
## Solution

```{r}
#| label: tbl-ci-width
#| tbl-cap: "Effect of Sample Size on CI Width (xbar = 105.4, s = 18.6, 95% CI)"

ns   <- c(36, 100, 400)
se   <- 18.6 / sqrt(ns)
tcr  <- qt(0.975, df = ns - 1)
me   <- tcr * se
lo   <- 105.4 - me
hi   <- 105.4 + me
wid  <- hi - lo

ci_tab <- data.frame(
  n     = ns,
  SE    = round(se, 3),
  t     = round(tcr, 3),
  ME    = round(me, 2),
  Lower = round(lo, 1),
  Upper = round(hi, 1),
  Width = round(wid, 2)
)

kable(ci_tab,
      col.names = c("n", "SE", "t(0.975, n-1)", "Margin of Error",
                    "Lower Bound", "Upper Bound", "CI Width"),
      align = rep("c", 7)) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = tbl_fs(13)) |>
  column_spec(1, bold = TRUE, color = "#2c7bb6") |>
  column_spec(7, bold = TRUE)
```

The first row is exactly the CI from Example 1.

**Conclusion:** Quadrupling the sample size (from 100 to 400) roughly
**halves** the CI width — because SE depends on $\sqrt{n}$, not $n$ directly
(the small extra change comes from the t value getting closer to 1.96).
To halve the margin of error, you need about **four times** the sample size.
:::

---

# Confidence Interval for a Proportion

When the outcome of interest is a **proportion** (e.g., prevalence of a disease,
percentage of patients who respond to treatment), we use:

$$\hat{p} \pm z_{1-\alpha/2} \cdot \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

Where $\hat{p} = \dfrac{x}{n}$ is the sample proportion
($x$ = number with the characteristic, $n$ = total sample size).

::: {.callout-warning icon=false}
## When Is This Formula Valid?
This formula requires a **sufficiently large sample**. A common rule of thumb is:

$$n\hat{p} \geq 5 \quad \text{and} \quad n(1-\hat{p}) \geq 5$$

If this condition is not met, special methods are needed (not covered in this course).
:::

---

## Example 3: 95% CI for a Proportion

In a survey of **200 adults**, **68 were found to have hypertension**.

Construct a **95% confidence interval** for the true prevalence of
hypertension in this population.

::: {.callout-tip icon=false}
## Solution

**Given:** $n = 200$, $x = 68$

**Step 1 — Sample proportion:**
$$\hat{p} = \frac{68}{200} = 0.34 \quad (34\%)$$

**Step 2 — Check validity:**
$$n\hat{p} = 200 \times 0.34 = 68 \geq 5 \checkmark$$
$$n(1-\hat{p}) = 200 \times 0.66 = 132 \geq 5 \checkmark$$

**Step 3 — Standard Error:**
$$SE = \sqrt{\frac{0.34 \times 0.66}{200}} = \sqrt{\frac{0.2244}{200}} = \sqrt{0.001122} = 0.0335$$

**Step 4 — Confidence Interval** (using $z_{0.975} = 1.96$):
$$0.34 \pm 1.96 \times 0.0335 = 0.34 \pm 0.066$$

$$\textbf{95\% CI: } (0.274, \; 0.406) \quad \text{or} \quad (27.4\%, \; 40.6\%)$$

**Interpretation:** We are 95% confident that the true prevalence of
hypertension in this population is between **27.4% and 40.6%**.
:::

---

# Sample Size for Estimating a Mean

Before a study starts, researchers must decide **how many participants**
to recruit. Too few, and the CI will be too wide to be useful; too many,
and time, money and patients' effort are wasted.

## Where the Formula Comes From

The **margin of error** $E$ is the "±" part of the CI:

$$E = z_{1-\alpha/2} \cdot \frac{\sigma}{\sqrt{n}}$$

If we decide in advance how small we want $E$ to be, we can solve this
equation for $n$:

$$\sqrt{n} = \frac{z_{1-\alpha/2} \cdot \sigma}{E}
\qquad \Rightarrow \qquad
n = \left(\frac{z_{1-\alpha/2} \cdot \sigma}{E}\right)^2$$

::: {.callout-important icon=false}
## Three Rules for Sample Size
1. **We use z, not t.** The t value depends on $n$, which we do not know yet.
2. **We need a value for $\sigma$** before the study — usually from a
   pilot study or a previous publication.
3. **Always round UP** to the next whole number, even if the decimal is
   small (e.g., 96.04 becomes 97). Rounding down would give a margin of
   error slightly *larger* than we wanted.
:::

## Worked Example

A researcher wants to estimate the mean systolic blood pressure of
elderly patients in a district, with a **95% CI** and a margin of error of
**no more than 3 mmHg**. A previous study reported **SD = 15 mmHg**.

$$n = \left(\frac{1.96 \times 15}{3}\right)^2 = (9.8)^2 = 96.04
\;\rightarrow\; \textbf{97 patients}$$

Notice the effect of each part of the formula:

- A **smaller** $E$ (more precision) needs a **larger** $n$ — halving $E$
  multiplies $n$ by 4.
- A **larger** $\sigma$ (more variable patients) needs a **larger** $n$.
- A **higher** confidence level (99% uses 2.576) needs a **larger** $n$.

---

# Key Takeaways

::: {.callout-note icon=false}
## Chapter 4 Summary

1. A **point estimate** gives a single best-guess value for a population
   parameter; an **interval estimate** gives a range of plausible values.

2. A **confidence interval (CI)** = Point estimate $\pm$ Margin of error.
   The margin of error depends on the confidence level, the SE, and the
   critical value.

3. Use the **Z-interval** only when $\sigma$ is known. When $\sigma$ is
   unknown (almost always), use the **t-interval** with $df = n - 1$ —
   this is what SPSS does. For large $n$, the t and Z values are very close.
   The **Central Limit Theorem** is why these intervals work for large
   samples even when the data are skewed.

4. The **95% CI** is the standard in health research, using
   $z_{0.975} = 1.96$.

5. A wider CI means less precision; a larger sample size $n$ always
   produces a narrower (more precise) CI.

6. **Sample size for a mean:** $n = \left(z_{1-\alpha/2}\,\sigma / E\right)^2$,
   always rounded **up**.

7. The correct interpretation: *"We are 95% confident the true parameter
   lies within this interval"* — not a probability statement about
   a fixed parameter.
:::

---

## Formula Summary

| Quantity | Formula | Notes |
|:----------------|:----------------------------------------|:----------------------|
| SE of a mean | $SE_{\bar{x}} = \dfrac{s}{\sqrt{n}}$ | Gets smaller as $n$ increases |
| SE of a proportion | $SE_{\hat{p}} = \sqrt{\dfrac{\hat{p}(1-\hat{p})}{n}}$ | $\hat{p} = x/n$ |
| CI for a mean, $\sigma$ known | $\bar{x} \pm z_{1-\alpha/2} \cdot \dfrac{\sigma}{\sqrt{n}}$ | Rare in practice |
| CI for a mean, $\sigma$ unknown | $\bar{x} \pm t_{1-\alpha/2,\,n-1} \cdot \dfrac{s}{\sqrt{n}}$ | $df = n - 1$; what SPSS uses |
| CI for a proportion | $\hat{p} \pm z_{1-\alpha/2} \cdot \sqrt{\dfrac{\hat{p}(1-\hat{p})}{n}}$ | Needs $n\hat{p} \geq 5$ and $n(1-\hat{p}) \geq 5$ |
| Sample size for a mean | $n = \left(\dfrac{z_{1-\alpha/2} \cdot \sigma}{E}\right)^2$ | Always round **up** |
| Critical values of Z | $z_{0.95} = 1.645$, $z_{0.975} = 1.96$, $z_{0.995} = 2.576$ | 90%, 95%, 99% CI |

: Formulas in this chapter {.striped}

---

# Exercises

::: {.callout-caution icon=false}
## Exercise 1

A random sample of **25 patients** with chronic kidney disease had their
serum creatinine levels measured:

$$\bar{x} = 3.82 \text{ mg/dL}, \qquad s = 1.14 \text{ mg/dL}$$

**(a)** Calculate the standard error of the mean.

**(b)** Find the appropriate critical value for a 95% CI.

**(c)** Construct the **95% confidence interval** for the population mean
serum creatinine level.

**(d)** Interpret the confidence interval in plain language.
:::

::: {.callout-caution icon=false}
## Exercise 2

A hospital records that **45 out of 180 patients** discharged last month
were readmitted within 30 days.

**(a)** Calculate the sample proportion of 30-day readmissions.

**(b)** Check whether the large-sample conditions are satisfied.

**(c)** Construct a **95% confidence interval** for the true 30-day
readmission rate.

**(d)** The hospital's target readmission rate is 20%.
Based on the confidence interval, is there evidence that the true rate
differs from this target? Explain.
:::

::: {.callout-caution icon=false}
## Exercise 3

A researcher wants to estimate the mean diastolic blood pressure (DBP)
of adults in a city. From a pilot study, the SD is estimated to be
**12 mmHg**.

The researcher wants a **95% CI with a margin of error no greater than
2 mmHg**.

Using the formula:
$$n = \left(\frac{z_{1-\alpha/2} \cdot \sigma}{E}\right)^2$$

where $E$ is the desired margin of error, calculate the **minimum sample
size** required. Remember to round **up**.
:::

---

*End of Chapter 4*

---

::: {.callout-note icon=false}
## Coming Up Next
**Chapter 5: Statistical Testing** — Instead of estimating a parameter,
we will test specific claims about populations using
**hypothesis testing** — one of the most powerful tools in health research.
:::
