# Estimation

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(knitr)
library(kableExtra)
library(patchwork)
```

---

# Introduction {.unnumbered}

::: {.callout-note icon=false}
## 💡 Why Does This Matter?

In practice, we almost never have access to an entire population.
We collect data from a **sample** and use it to draw conclusions about
the **population** — a process called **statistical inference**.

**Estimation** is one of the two main branches of statistical inference
(the other being hypothesis testing, covered in Chapter 5).

In this chapter, we answer questions like:

- The mean serum cholesterol in our sample is 198 mg/dL —
  what can we say about the mean for the *entire* population?
- A survey found that 34% of respondents have hypertension —
  what is a reasonable range for the true population proportion?

The key tool for answering these questions is the **confidence interval**.
:::

---

# Concept of Estimation

## Point Estimate vs. Interval Estimate

There are two ways to estimate a population parameter from sample data:

### Point Estimate

A **point estimate** is a **single value** calculated from the sample
that serves as our best guess for the population parameter.

| Population Parameter | Symbol | Point Estimate | Symbol |
|---|:---:|---|:---:|
| Population mean | $\mu$ | Sample mean | $\bar{x}$ |
| Population proportion | $p$ | Sample proportion | $\hat{p}$ |
| Population SD | $\sigma$ | Sample SD | $s$ |

A point estimate is simple and easy to communicate, but it gives us
**no information about how accurate or reliable** that single number is.

---

### Interval Estimate (Confidence Interval)

An **interval estimate** gives a **range of plausible values** for the
population parameter, along with a stated level of confidence.

$$\text{Point Estimate} \pm \text{Margin of Error}$$

This range is called a **confidence interval (CI)**.

::: {.callout-important icon=false}
## 🔑 Key Concept: Confidence Level

The **confidence level** (commonly **90%, 95%, or 99%**) tells us:

> *"If we repeated this study many times and calculated a confidence interval
> each time, approximately X% of those intervals would contain the true
> population parameter."*

In health research, **95% confidence** is the most widely used standard.
:::

---

## The Role of the Standard Error

Before constructing a confidence interval, we need to understand the
**standard error (SE)** — a measure of how much the sample estimate
varies from sample to sample.

### Standard Error of the Mean

$$SE_{\bar{x}} = \frac{s}{\sqrt{n}}$$

Where $s$ is the sample standard deviation and $n$ is the sample size.

### Standard Error of a Proportion

$$SE_{\hat{p}} = \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

::: {.callout-tip icon=false}
## 💡 SE vs. SD — What Is the Difference?

| | Standard Deviation (SD) | Standard Error (SE) |
|---|---|---|
| **Describes** | Variability of individual observations | Variability of the sample mean |
| **Gets smaller with** | More homogeneous data | Larger sample size |
| **Used for** | Describing the spread of data | Constructing confidence intervals |

The SE always gets smaller as $n$ increases — larger samples give
more precise estimates.
:::

```{r}
#| label: fig-se
#| fig-cap: "As sample size increases, the standard error decreases and estimates become more precise"
#| fig-align: center
#| fig-height: 3.5
#| fig-width: 6

n_vals <- seq(10, 500, by = 10)
se_vals <- 15 / sqrt(n_vals)   # SD = 15, typical for SBP

df_se <- data.frame(n = n_vals, se = se_vals)

ggplot(df_se, aes(x = n, y = se)) +
  geom_line(color = "#2c7bb6", linewidth = 1.3) +
  geom_hline(yintercept = c(1, 2), linetype = "dashed",
             color = "#d7191c", alpha = 0.6) +
  annotate("text", x = 480, y = 2.3, label = "SE = 2", color = "#d7191c", size = 3.5) +
  annotate("text", x = 480, y = 1.3, label = "SE = 1", color = "#d7191c", size = 3.5) +
  labs(x = "Sample Size (n)", y = "Standard Error",
       title = "Standard Error vs. Sample Size (SD = 15)") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

---

# Confidence Interval for a Mean

## When the Population SD Is Known (Z-interval)

If the population standard deviation $\sigma$ is known and either
(a) the population is normal, or (b) the sample size is large ($n \geq 30$),
we use the **Z-interval**:

$$\bar{x} \pm z_{\alpha/2} \cdot \frac{\sigma}{\sqrt{n}}$$

The **critical value** $z_{\alpha/2}$ depends on the confidence level:

```{r}
#| label: tbl-z-critical
#| tbl-cap: "Critical Values for Common Confidence Levels"

crit <- data.frame(
  `Confidence Level` = c("90%", "95%", "99%"),
  `α` = c("0.10", "0.05", "0.01"),
  `α/2` = c("0.05", "0.025", "0.005"),
  `z(α/2)` = c("1.645", "1.960", "2.576")
)

kable(crit,
      col.names = c("Confidence Level", "α", "α/2", "z(α/2)"),
      align = c("c","c","c","c")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = 13) |>
  column_spec(1, bold = TRUE, color = "#2c7bb6") |>
  row_spec(2, background = "#eaf4fb")
```

::: {.callout-tip icon=false}
## 💡 The Most Used Value
In health research, the 95% CI with $z_{0.025} = 1.96$ is by far the
most commonly reported. You will see it in almost every clinical paper.
:::

---

## When the Population SD Is Unknown (t-interval)

In practice, $\sigma$ is almost never known. When we use the sample SD $s$
as an estimate, we must use the **t-distribution** instead of the Z-distribution:

$$\bar{x} \pm t_{\alpha/2,\, n-1} \cdot \frac{s}{\sqrt{n}}$$

Where $t_{\alpha/2,\, n-1}$ is the critical value from the **t-distribution**
with $n - 1$ **degrees of freedom (df)**.

### The t-Distribution

The t-distribution is similar to the standard normal distribution but has
**heavier tails** — reflecting the extra uncertainty from estimating $\sigma$
with $s$. As $n$ increases, the t-distribution approaches the Z-distribution.

```{r}
#| label: fig-t-dist
#| fig-cap: "t-distributions with different degrees of freedom compared to Z"
#| fig-align: center
#| fig-height: 4
#| fig-width: 7

x <- seq(-4, 4, length = 500)
df_plot <- data.frame(
  x = rep(x, 4),
  y = c(dnorm(x),
        dt(x, df = 3),
        dt(x, df = 10),
        dt(x, df = 30)),
  dist = rep(c("Z (normal)", "t (df=3)", "t (df=10)", "t (df=30)"), each = 500)
)

df_plot$dist <- factor(df_plot$dist,
                       levels = c("t (df=3)","t (df=10)","t (df=30)","Z (normal)"))

ggplot(df_plot, aes(x = x, y = y, color = dist, linetype = dist)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("#d7191c","#fdae61","#abd9e9","#2c7bb6")) +
  scale_linetype_manual(values = c("dashed","dashed","dashed","solid")) +
  labs(x = "Value", y = "Density", color = "Distribution", linetype = "Distribution") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right",
        panel.grid.minor = element_blank())
```

### t Critical Values

```{r}
#| label: tbl-t-critical
#| tbl-cap: "Selected t Critical Values (95% CI, two-tailed: α/2 = 0.025)"

df_vals <- c(5, 10, 15, 20, 25, 30, 40, 60, 120, Inf)
t_vals  <- qt(0.975, df = df_vals)
t_vals[length(t_vals)] <- 1.960

t_tab <- data.frame(
  df = c(as.character(df_vals[-length(df_vals)]), "∞ (Z)"),
  t  = round(t_vals, 3)
)

kable(t_tab, col.names = c("Degrees of Freedom (df = n−1)", "t(0.025, df)"),
      align = c("c","c")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = 13) |>
  column_spec(1, bold = TRUE, color = "#2c7bb6") |>
  row_spec(10, background = "#eaf4fb", bold = TRUE)
```

As df increases, the t critical value converges to **1.960** (the Z value).
For $n \geq 30$, the difference is small and Z is often used as an approximation.

---

# Confidence Interval for a Proportion

When the outcome of interest is a **proportion** (e.g., prevalence of a disease,
percentage of patients who respond to treatment), we use:

$$\hat{p} \pm z_{\alpha/2} \cdot \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

Where $\hat{p} = \dfrac{x}{n}$ is the sample proportion
($x$ = number with the characteristic, $n$ = total sample size).

::: {.callout-warning icon=false}
## ⚠️ When Is This Formula Valid?
This formula requires a **sufficiently large sample**. A common rule of thumb is:

$$n\hat{p} \geq 5 \quad \text{and} \quad n(1-\hat{p}) \geq 5$$

If this condition is not met, special methods are needed (not covered in this course).
:::

---

# Interpreting a Confidence Interval

::: {.callout-important icon=false}
## ✅ Correct Interpretation

> *"We are 95% confident that the true population mean lies between
> [lower bound] and [upper bound]."*

**What this means:** If we repeated the sampling process many times,
95% of the resulting confidence intervals would contain the true parameter.

**What this does NOT mean:** There is NOT a 95% probability that the
true parameter lies in this specific interval — the true parameter is fixed;
it either is or is not in the interval.
:::

```{r}
#| label: fig-ci-simulation
#| fig-cap: "Simulation: 50 confidence intervals from repeated samples. Green intervals contain the true mean (μ = 130); red ones do not."
#| fig-align: center
#| fig-height: 6
#| fig-width: 7

set.seed(42)
true_mu <- 130
sigma    <- 15
n        <- 40
n_sims   <- 50

sims <- replicate(n_sims, {
  x    <- rnorm(n, mean = true_mu, sd = sigma)
  xbar <- mean(x)
  se   <- sd(x) / sqrt(n)
  c(xbar, xbar - 1.96*se, xbar + 1.96*se)
})

ci_df <- data.frame(
  sim   = 1:n_sims,
  mean  = sims[1,],
  lower = sims[2,],
  upper = sims[3,],
  contains = sims[2,] <= true_mu & sims[3,] >= true_mu
)

ggplot(ci_df, aes(y = sim, color = contains)) +
  geom_segment(aes(x = lower, xend = upper, yend = sim), linewidth = 0.8) +
  geom_point(aes(x = mean), size = 1.5) +
  geom_vline(xintercept = true_mu, color = "black",
             linetype = "dashed", linewidth = 1) +
  scale_color_manual(values = c("TRUE" = "#1a9641", "FALSE" = "#d7191c"),
                     labels = c("TRUE" = "Contains μ", "FALSE" = "Does not contain μ"),
                     name = "") +
  annotate("text", x = true_mu + 0.5, y = 52,
           label = "True μ = 130", hjust = 0, size = 3.5) +
  labs(x = "Value", y = "Sample Number") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
```

The figure above shows 50 confidence intervals from 50 different samples
of the same population. Most (≈95%) contain the true mean; a few do not —
this is expected and not an error.

---

# Key Takeaways

::: {.callout-note icon=false}
## 📝 Chapter 4 Summary

1. A **point estimate** gives a single best-guess value for a population
   parameter; an **interval estimate** gives a range of plausible values.

2. A **confidence interval (CI)** = Point estimate $\pm$ Margin of error.
   The margin of error depends on the confidence level, the SE, and the
   critical value.

3. Use the **Z-interval** when $\sigma$ is known or $n \geq 30$;
   use the **t-interval** when $\sigma$ is unknown and $n$ is small,
   with $df = n - 1$.

4. The **95% CI** is the standard in health research, using
   $z_{0.025} = 1.96$.

5. A wider CI means less precision; a larger sample size $n$ always
   produces a narrower (more precise) CI.

6. The correct interpretation: *"We are 95% confident the true parameter
   lies within this interval"* — not a probability statement about
   a fixed parameter.
:::

---

# Examples

## Example 1: 95% CI for a Mean (σ unknown)

A study measured the fasting blood glucose (mg/dL) of a random sample of
**36 adults** and found:

$$\bar{x} = 105.4 \text{ mg/dL}, \qquad s = 18.6 \text{ mg/dL}$$

Construct a **95% confidence interval** for the population mean blood glucose.

::: {.callout-tip icon=false}
## ✅ Solution

**Given:** $n = 36$, $\bar{x} = 105.4$, $s = 18.6$, confidence level = 95%

**Step 1 — Standard Error:**
$$SE = \frac{s}{\sqrt{n}} = \frac{18.6}{\sqrt{36}} = \frac{18.6}{6} = 3.10$$

**Step 2 — Critical value:**

$df = n - 1 = 35$. From the t-table at $\alpha/2 = 0.025$:
$t_{0.025,\, 35} = 2.030$

*(Since $n = 36 \geq 30$, we could also use $z = 1.96$ as an approximation.)*

**Step 3 — Confidence Interval:**
$$105.4 \pm 2.030 \times 3.10 = 105.4 \pm 6.29$$

$$\textbf{95\% CI: } (99.1, \; 111.7) \text{ mg/dL}$$

**Interpretation:** We are 95% confident that the true mean fasting blood
glucose of adults in this population is between **99.1 and 111.7 mg/dL**.
:::

---

## Example 2: 95% CI for a Proportion

In a survey of **200 adults**, **68 were found to have hypertension**.

Construct a **95% confidence interval** for the true prevalence of
hypertension in this population.

::: {.callout-tip icon=false}
## ✅ Solution

**Given:** $n = 200$, $x = 68$

**Step 1 — Sample proportion:**
$$\hat{p} = \frac{68}{200} = 0.34 \quad (34\%)$$

**Step 2 — Check validity:**
$$n\hat{p} = 200 \times 0.34 = 68 \geq 5 \checkmark$$
$$n(1-\hat{p}) = 200 \times 0.66 = 132 \geq 5 \checkmark$$

**Step 3 — Standard Error:**
$$SE = \sqrt{\frac{0.34 \times 0.66}{200}} = \sqrt{\frac{0.2244}{200}} = \sqrt{0.001122} = 0.0335$$

**Step 4 — Confidence Interval** (using $z_{0.025} = 1.96$):
$$0.34 \pm 1.96 \times 0.0335 = 0.34 \pm 0.066$$

$$\textbf{95\% CI: } (0.274, \; 0.406) \quad \text{or} \quad (27.4\%, \; 40.6\%)$$

**Interpretation:** We are 95% confident that the true prevalence of
hypertension in this population is between **27.4% and 40.6%**.
:::

---

## Example 3: Effect of Sample Size on CI Width

Using the same data as Example 1 ($\bar{x} = 105.4$, $s = 18.6$),
compare the 95% CI when $n = 36$, $n = 100$, and $n = 400$.

::: {.callout-tip icon=false}
## ✅ Solution

```{r}
#| label: tbl-ci-width
#| tbl-cap: "Effect of Sample Size on CI Width (x̄ = 105.4, s = 18.6, 95% CI)"

ns   <- c(36, 100, 400)
se   <- 18.6 / sqrt(ns)
z    <- 1.96
me   <- z * se
lo   <- 105.4 - me
hi   <- 105.4 + me
wid  <- hi - lo

ci_tab <- data.frame(
  n     = ns,
  SE    = round(se, 3),
  ME    = round(me, 3),
  Lower = round(lo, 2),
  Upper = round(hi, 2),
  Width = round(wid, 2)
)

kable(ci_tab,
      col.names = c("n", "SE", "Margin of Error",
                    "Lower Bound", "Upper Bound", "CI Width"),
      align = rep("c", 6)) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = 13) |>
  column_spec(1, bold = TRUE, color = "#2c7bb6") |>
  column_spec(6, bold = TRUE)
```

**Conclusion:** Quadrupling the sample size (from 100 to 400) **halves**
the CI width — because SE depends on $\sqrt{n}$, not $n$ directly.
To halve the margin of error, you need to **quadruple** the sample size.
:::

---

# Exercises

::: {.callout-caution icon=false}
## ✏️ Exercise 1

A random sample of **25 patients** with chronic kidney disease had their
serum creatinine levels measured:

$$\bar{x} = 3.82 \text{ mg/dL}, \qquad s = 1.14 \text{ mg/dL}$$

**(a)** Calculate the standard error of the mean.

**(b)** Find the appropriate critical value for a 95% CI.

**(c)** Construct the **95% confidence interval** for the population mean
serum creatinine level.

**(d)** Interpret the confidence interval in plain language.
:::

::: {.callout-caution icon=false}
## ✏️ Exercise 2

A hospital records that **45 out of 180 patients** discharged last month
were readmitted within 30 days.

**(a)** Calculate the sample proportion of 30-day readmissions.

**(b)** Check whether the large-sample conditions are satisfied.

**(c)** Construct a **95% confidence interval** for the true 30-day
readmission rate.

**(d)** The hospital's target readmission rate is 20%.
Based on the confidence interval, is there evidence that the true rate
differs from this target? Explain.
:::

::: {.callout-caution icon=false}
## ✏️ Exercise 3

A researcher wants to estimate the mean diastolic blood pressure (DBP)
of adults in a city. From a pilot study, the SD is estimated to be
**12 mmHg**.

The researcher wants a **95% CI with a margin of error no greater than
2 mmHg**.

Using the formula:
$$n = \left(\frac{z_{\alpha/2} \cdot \sigma}{E}\right)^2$$

where $E$ is the desired margin of error, calculate the **minimum sample
size** required.
:::

---

*End of Chapter 4*

---

::: {.callout-note icon=false}
## 📚 Coming Up Next
**Chapter 5: Statistical Testing** — Instead of estimating a parameter,
we will test specific claims about populations using
**hypothesis testing** — one of the most powerful tools in health research.
:::
