# Statistical Testing

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(knitr)
library(kableExtra)
library(patchwork)
# Table helpers (used by the book and the slides)
# tbl_fs(): font size; PDF tables use 10pt so they fit the page width
# pdf_w():  fixed column width in the PDF only, so long text wraps
tbl_fs <- function(html_size = 13) if (knitr::is_latex_output()) 10 else html_size
pdf_w  <- function(x) if (knitr::is_latex_output()) x else NULL
```

---

# Introduction {.unnumbered}

::: {.callout-note icon=false}
## Why Does This Matter?

In health research, we constantly ask questions that require a decision:

- Does this new drug lower blood pressure *more* than the standard treatment?
- Is there a relationship between smoking and lung disease?
- Do recovery times differ between two surgical techniques?

We cannot answer these questions simply by looking at sample means or
proportions — random chance alone can create apparent differences between
groups. **Statistical testing** gives us a formal, objective framework
for deciding whether an observed difference is likely to be real or
simply due to chance.

This is the most widely used tool in health research, and understanding
it is essential for reading and interpreting scientific literature.
:::

---

# Concept of Hypothesis Testing

## The Logic of Hypothesis Testing

Statistical testing works by assuming the opposite of what we want to prove,
then asking: *"If nothing were really happening, how surprising would our
data be?"*

If the data would be very surprising under the assumption that nothing is
happening, we conclude that something probably *is* happening.

---

## The Two Hypotheses

Every hypothesis test involves two competing statements:

### Null Hypothesis ($H_0$)

The **null hypothesis** states that there is **no effect, no difference,
or no relationship** in the population. It is the assumption we start with
and attempt to disprove.

> *Examples:*
> - The mean blood pressure is 120 mmHg. ($H_0: \mu = 120$)
> - There is no difference in recovery time between the two groups. ($H_0: \mu_1 = \mu_2$)
> - There is no association between smoking status and disease. ($H_0:$ no association)

### Alternative Hypothesis ($H_1$ or $H_a$)

The **alternative hypothesis** states that there **is** an effect,
difference, or relationship. This is usually what the researcher
believes or wants to demonstrate.

> *Examples:*
> - The mean blood pressure is not 120 mmHg. ($H_1: \mu \neq 120$)
> - Recovery time differs between the two groups. ($H_1: \mu_1 \neq \mu_2$)
> - There is an association between smoking and disease. ($H_1:$ association exists)

---

## One-Tailed vs. Two-Tailed Tests

The alternative hypothesis can be **directional** or **non-directional**:

```{r}
#| label: tbl-tails
#| tbl-cap: "One-tailed vs. Two-tailed Tests"

tails <- data.frame(
  Type = c("Two-tailed", "One-tailed (right)", "One-tailed (left)"),
  `H1` = c("mu != mu0  (different from)", "mu > mu0  (greater than)", "mu < mu0  (less than)"),
  `Use When` = c(
    "Either direction (most common)",
    "Only an increase is of interest",
    "Only a decrease is of interest"
  )
)

kable(tails, col.names = c("Test Type", "H1", "Use When"),
      align = c("l","l","l")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = tbl_fs(13)) |>
  column_spec(1, bold = TRUE, color = "#2c7bb6") |>
  row_spec(1, background = "#eaf4fb")
```

::: {.callout-tip icon=false}
## Practical Advice
In most health research, use a **two-tailed test** unless you have a
strong, pre-specified reason to test in only one direction.
Two-tailed tests are more conservative and more widely accepted.
SPSS reports two-tailed p-values ("Sig. (2-tailed)"), and **all tests in
this course are two-tailed**.

After a significant two-tailed test, we still say **which direction** the
difference goes by looking at the sample means (or mean ranks).
:::

---

## The Test Statistic

A **test statistic** is a single number calculated from the sample that
measures **how far the data are from what $H_0$ predicts**. For the
t-tests it has the same form as the Z-score in Chapter 3:

$$\text{test statistic} = \frac{\text{what we observed} - \text{what } H_0 \text{ predicts}}{\text{standard error}}$$

- A test statistic **close to 0** means the data agree with $H_0$.
- A test statistic **far from 0** means the data are surprising if $H_0$ is true.

Each test in this chapter has its own test statistic ($t$, $\chi^2$, $U$, $W$)
and its own formula, but the idea is the same: the further the data are
from what $H_0$ predicts, the more extreme the test statistic.

---

## The p-value

The **p-value** is the probability of obtaining results *at least as extreme
as those observed*, assuming the null hypothesis is true.

$$p\text{-value} = P(\text{observed result or more extreme} \mid H_0 \text{ is true})$$

- A **small p-value** means the observed data would be very unlikely if
  $H_0$ were true $\rightarrow$ **evidence against $H_0$**
- A **large p-value** means the observed data are consistent with $H_0$
  $\rightarrow$ **insufficient evidence to reject $H_0$**

---

## The Significance Level (α)

The **significance level** $\alpha$ is the threshold we set in advance
for deciding whether the p-value is "small enough" to reject $H_0$.

The most common choice in health research is $\alpha = 0.05$.

| Decision Rule | |
|---|---|
| If $p\text{-value} \leq \alpha$ | **Reject $H_0$** — result is statistically significant |
| If $p\text{-value} > \alpha$ | **Fail to reject $H_0$** — insufficient evidence |

::: {.callout-warning icon=false}
## Common Misinterpretations

**"Failing to reject $H_0$" is NOT the same as proving $H_0$ is true.**
It simply means the data do not provide strong enough evidence against it.

**"Statistically significant" does NOT necessarily mean clinically important.**
With a very large sample, even a trivially small difference can become
statistically significant. Always consider the practical meaning of results.
:::

---

## Two Ways to Make the Decision: Critical Value or p-value

There are two equivalent ways to decide whether to reject $H_0$.
They **always give the same decision**.

**1. Critical value approach (calculating by hand).**
The **critical value** is the cut-off from a statistical table
(Z, t or $\chi^2$) that marks the most extreme $\alpha$ of the distribution.
We use the same left-cumulative notation as in Chapter 4. For a
two-tailed test at $\alpha = 0.05$ the cut-off is at $1 - \alpha/2 = 0.975$,
for example $t_{0.975,\,9} = 2.262$.

- If $|\text{test statistic}| >$ critical value $\rightarrow$ **reject $H_0$**
- If $|\text{test statistic}| \leq$ critical value $\rightarrow$ **do not reject $H_0$**

*(The rank tests later in this chapter also use this rule: we convert their
test statistic to a Z-score first.)*

**2. p-value approach (reading SPSS output).**
An exact p-value is hard to find by hand, but SPSS calculates it for us
(the column **Sig.**). We compare it with $\alpha$ as in the table above.

```{r}
#| label: fig-critical
#| fig-cap: "Two-tailed test at alpha = 0.05 with df = 9: reject H0 when the test statistic falls in a red tail beyond the critical value 2.262"
#| fig-align: center
#| fig-height: 3.2
#| fig-width: 7

x  <- seq(-5, 5, length = 600)
dd <- data.frame(x = x, y = dt(x, df = 9))
cv <- qt(0.975, df = 9)

ggplot(dd, aes(x, y)) +
  geom_area(data = subset(dd, x <= -cv), fill = "#d7191c", alpha = 0.6) +
  geom_area(data = subset(dd, x >=  cv), fill = "#d7191c", alpha = 0.6) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = c(-cv, cv), linetype = "dashed", color = "#d7191c") +
  annotate("text", x = 0, y = 0.12, label = "Do not reject H0\n(middle 95%)", size = 3.6) +
  annotate("text", x = -3.9, y = 0.07, label = "Reject H0\n(2.5%)", size = 3.3, color = "#d7191c") +
  annotate("text", x =  3.9, y = 0.07, label = "Reject H0\n(2.5%)", size = 3.3, color = "#d7191c") +
  scale_x_continuous(breaks = c(-round(cv, 3), 0, round(cv, 3))) +
  labs(x = "t", y = "") +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), axis.text.y = element_blank())
```

In this book, worked examples use the **critical value** for the hand
calculation and then show the **p-value** that SPSS would report, so you
can see that both lead to the same conclusion. In the lab, you will
use the p-value.

---

## Type I and Type II Errors

When making a decision about $H_0$, two types of errors are possible:

```{r}
#| label: tbl-errors
#| tbl-cap: "Type I and Type II Errors in Hypothesis Testing"

errors <- data.frame(
  ` ` = c("Reject H0", "Fail to reject H0"),
 `H0 is actually TRUE` = c("Type I Error (alpha) - False Positive",
                            "Correct Decision"),
  `H0 is actually FALSE` = c("Correct Decision (Power = 1 - beta)",
                             "Type II Error (beta) - False Negative")
)

kable(errors, col.names = c("Our Decision", "H0 is TRUE", "H0 is FALSE"),
      align = c("l","c","c")) |>
  kable_styling(bootstrap_options = c("bordered"),
                full_width = FALSE, font_size = tbl_fs(13)) |>
  column_spec(1, bold = TRUE) |>
  column_spec(2, background = c("#fde0d9","#d9f0d3")) |>
  column_spec(3, background = c("#d9f0d3","#fde0d9"))
```

- **Type I error (α):** Concluding there is an effect when there is none
  (*false positive*). Controlled by choosing α (usually 0.05).
- **Type II error (β):** Missing a real effect (*false negative*).
  The probability of correctly detecting a real effect is called **power** = $1 - \beta$.

---

## General Steps in Hypothesis Testing

::: {.callout-important icon=false}
## The 5-Step Framework

**Step 1 — State the hypotheses:** Write $H_0$ and $H_1$ clearly.

**Step 2 — Choose the significance level:** Usually $\alpha = 0.05$.

**Step 3 — Select and compute the test statistic:**
Choose the test that matches the data type and study design, and
calculate its test statistic.

**Step 4 — Compare:**
By hand, compare the test statistic with the **critical value** from the table;
with SPSS, read the **p-value** (Sig.).

**Step 5 — Make a decision and interpret:**
Reject or do not reject $H_0$, then state your conclusion in the context
of the problem — including the **direction** of any difference.
:::

---

# Statistical Testing on Categorical Data

## Chi-Square Test for Two Independent Groups

### When to Use

The **chi-square test of independence** is used when:

- The outcome variable is **categorical (nominal)**
- We are comparing **two or more independent groups**
- We want to test whether there is an **association** between two categorical variables

### The Contingency Table

Data are arranged in a **contingency table** (also called a cross-tabulation):

$$\begin{array}{|l|c|c|c|}
\hline
& \textbf{Outcome: Yes} & \textbf{Outcome: No} & \textbf{Row Total} \\
\hline
\textbf{Group 1} & a & b & a+b \\
\textbf{Group 2} & c & d & c+d \\
\hline
\textbf{Column Total} & a+c & b+d & n \\
\hline
\end{array}$$

### The Test Statistic

$$\chi^2 = \sum \frac{(O - E)^2}{E}$$

Where:

- $O$ = observed frequency (actual count in each cell)
- $E$ = expected frequency = $\dfrac{\text{Row total} \times \text{Column total}}{n}$

The sum is taken over **all cells** in the table. The expected frequency
is the count we would expect in a cell **if $H_0$ were true** (no association),
so a large $\chi^2$ means the observed counts are far from what $H_0$ predicts.

The test statistic follows a **chi-square distribution** with
$df = (\text{rows} - 1) \times (\text{columns} - 1)$.

For a $2 \times 2$ table: $df = (2-1)(2-1) = 1$.

Because $\chi^2$ is a sum of squares, it is never negative, and only
**large** values are evidence against $H_0$. The critical value therefore
uses the left cumulative probability $1 - \alpha$:
at $\alpha = 0.05$ and $df = 1$, $\chi^2_{0.95,\,1} = 3.841$.

::: {.callout-warning icon=false}
## Condition for Valid Chi-Square Test
All **expected frequencies** must be $\geq 5$.
If this condition is violated, use **Fisher's exact test** instead
(available in SPSS).
:::

### Chi-Square Critical Values

```{r}
#| label: tbl-chisq
#| tbl-cap: "Chi-Square Critical Values at alpha = 0.05 (left cumulative probability 0.95)"

df_chi  <- 1:10
cv_chi  <- round(qchisq(0.95, df = df_chi), 3)

chi_tab <- data.frame(df = df_chi, critical = cv_chi)

kable(chi_tab, col.names = c("Degrees of Freedom (df)", "chi2(0.95, df)"),
      align = c("c","c")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = tbl_fs(13)) |>
  column_spec(1, bold = TRUE, color = "#2c7bb6") |>
  row_spec(1, background = "#eaf4fb")
```

::: {.callout-tip icon=false}
## Reading the SPSS Output (Crosstabs)
SPSS prints several rows in the **Chi-Square Tests** table. In this course:

1. First read the **footnote** under the table, e.g.
   *"0 cells (0.0%) have expected count less than 5."*
2. If **no** cell has an expected count below 5 $\rightarrow$ read the
   **Pearson Chi-Square** row (Value, df, Asymp. Sig.).
3. If **any** cell has an expected count below 5 $\rightarrow$ read the
   **Fisher's Exact Test** row instead (Exact Sig. 2-sided).

The other rows (Continuity Correction, Likelihood Ratio,
Linear-by-Linear Association) are not used in this course.
:::

---

## Example 1: Chi-Square Test

A cross-sectional study investigates whether **smoking status** is
associated with **hypertension** in a sample of 200 adults.

|  | Hypertension | No Hypertension | Total |
|---|:---:|:---:|:---:|
| Smoker | 45 | 55 | 100 |
| Non-smoker | 25 | 75 | 100 |
| **Total** | **70** | **130** | **200** |

Test at $\alpha = 0.05$ whether smoking is associated with hypertension.

::: {.callout-tip icon=false}
## Solution

**Step 1 — Hypotheses:**

$H_0$: Smoking status and hypertension are independent (no association).

$H_1$: Smoking status and hypertension are associated.

**Step 2 —** $\alpha = 0.05$

**Step 3 — Expected frequencies and test statistic:**

$$E = \frac{\text{Row total} \times \text{Column total}}{n}$$

```{r}
#| label: tbl-expected
#| tbl-cap: "Observed (O) and Expected (E) Frequencies"

obs <- matrix(c(45, 55, 25, 75), nrow=2, byrow=TRUE)
row_totals <- rowSums(obs)
col_totals <- colSums(obs)
n <- sum(obs)

exp_freq <- outer(row_totals, col_totals) / n

result_tab <- data.frame(
  Group = c("Smoker", "Non-smoker"),
  O_hyp  = obs[,1],
  E_hyp  = exp_freq[,1],
  O_no   = obs[,2],
  E_no   = exp_freq[,2]
)

kable(result_tab,
      col.names = c("Group","O (Hyp.)","E (Hyp.)","O (No Hyp.)","E (No Hyp.)"),
      align = rep("c",5), digits = 2) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = tbl_fs(13)) |>
  column_spec(1, bold = TRUE)
```

All expected frequencies are $\geq 5$, so the chi-square test is valid.

$$\chi^2 = \frac{(45-35)^2}{35} + \frac{(55-65)^2}{65} +
           \frac{(25-35)^2}{35} + \frac{(75-65)^2}{65}$$

$$= \frac{100}{35} + \frac{100}{65} + \frac{100}{35} + \frac{100}{65}
= 2.857 + 1.538 + 2.857 + 1.538 = \mathbf{8.791}$$

$df = (2-1)(2-1) = 1$

**Step 4 — Compare:**

Critical value from the chi-square table: $\chi^2_{0.95,\,1} = 3.841$.
Since $8.791 > 3.841$, the test statistic falls in the rejection region.

SPSS (Pearson Chi-Square row) would report $p = 0.003$, which is
$< 0.05$ — the same decision.

**Step 5 — Decision and interpretation:**

We **reject $H_0$**. There is a statistically significant association between
smoking status and hypertension ($\chi^2 = 8.79$, $df = 1$, $p = 0.003$).
Smokers have a higher prevalence of hypertension (45%) than non-smokers (25%).

*Because this is a cross-sectional study, the result shows an
**association**, not that smoking **causes** hypertension.*
:::

---

# Statistical Testing on Quantitative Data

The two t-tests in this section compare **means**. Like the confidence
intervals in Chapter 4, they rely on the data (or the sample means) being
approximately **normal**.

## Checking the Normality Assumption

Before using a t-test, check whether the outcome is approximately normal:

- **Graphs:** a histogram should look roughly bell-shaped, and the points on
  a **normal Q–Q plot** should lie close to the straight line.
- **Tests:** SPSS *Explore* reports the **Kolmogorov–Smirnov** and
  **Shapiro–Wilk** tests (you practise these in the lab). Their hypotheses are

  $H_0$: the data come from a normal distribution

  $H_1$: the data do not come from a normal distribution

  so **Sig. > 0.05** means *no evidence against normality* $\rightarrow$
  a t-test is appropriate; **Sig. $\leq$ 0.05** means the data are not
  normal $\rightarrow$ consider a rank test (later in this chapter).

What to check depends on the design:

| Test | Check normality of |
|---|---|
| Independent samples t-test | The outcome **in each group separately** |
| Paired samples t-test | The **differences** $d_i$ within pairs |

With large samples ($n \geq 30$ per group) the **Central Limit Theorem**
(Chapter 4) makes the t-test robust even if the data are somewhat skewed.

---

## Independent Samples t-Test

### When to Use

The **independent samples t-test** is used when:

- The outcome variable is **continuous (numerical)**
- We are comparing the **means of two independent groups**
- The data are approximately **normally distributed** in each group
  (or sample sizes are large enough, $n \geq 30$ per group)

### Hypotheses

$$H_0: \mu_1 = \mu_2 \quad \text{(no difference in population means)}$$
$$H_1: \mu_1 \neq \mu_2 \quad \text{(the population means differ)}$$

### The Test Statistic

$$t = \frac{\bar{x}_1 - \bar{x}_2}{SE_{\bar{x}_1 - \bar{x}_2}}$$

The standard error of the difference depends on whether the two groups
have **equal or unequal variances**.

**Equal variances assumed (pooled):**

$$SE = s_p\sqrt{\frac{1}{n_1}+\frac{1}{n_2}}, \qquad
s_p = \sqrt{\frac{(n_1-1)s_1^2 + (n_2-1)s_2^2}{n_1+n_2-2}}$$

**Equal variances not assumed (Welch's):**

$$SE = \sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}$$

The test statistic follows a **t-distribution** with $df = n_1 + n_2 - 2$
(equal variances) or an approximated df (Welch's, calculated by SPSS).

### Which Version? Levene's Test

SPSS decides between the two versions with **Levene's test for equality
of variances**, printed in the first columns of the
*Independent Samples Test* table:

$H_0: \sigma_1^2 = \sigma_2^2$ (the two variances are equal)

$H_1: \sigma_1^2 \neq \sigma_2^2$ (the variances differ)

::: {.callout-important icon=false}
## Reading the Independent Samples Test Table — Two Steps
**Step 1 — Levene's test (Sig. in the Levene columns):**

- Sig. **> 0.05** $\rightarrow$ no evidence that the variances differ
  $\rightarrow$ read the **top row**, *"Equal variances assumed"*.
- Sig. **$\leq$ 0.05** $\rightarrow$ the variances differ
  $\rightarrow$ read the **bottom row**, *"Equal variances not assumed"*.

**Step 2 — the t-test (Sig. (2-tailed) in the row you chose):**
compare with $\alpha$ to decide whether the **means** differ.

Do not confuse the two Sig. values: the first is about **variances**,
the second is about **means**.
:::

When you calculate by hand in this course, the question will either give
you the Levene's test result or state whether the variances can be assumed
equal. You do not need to decide this yourself from the SDs.

---

## Example 2: Independent Samples t-Test

A clinical trial compares the mean reduction in systolic blood pressure (mmHg)
between patients receiving a **new drug** vs. a **placebo** after 8 weeks.

| Group | n | Mean reduction | SD |
|---|:---:|:---:|:---:|
| New drug | 40 | 12.4 | 6.8 |
| Placebo | 40 | 7.1 | 7.2 |

Assume the two population variances are **equal**. Test at $\alpha = 0.05$
whether the mean reduction in SBP differs between the new drug and placebo.

::: {.callout-tip icon=false}
## Solution

**Step 1 — Hypotheses:**

$H_0: \mu_1 = \mu_2$ (no difference in mean reduction)

$H_1: \mu_1 \neq \mu_2$ (two-tailed; difference exists)

**Step 2 —** $\alpha = 0.05$

**Step 3 — Pooled standard error.** The question states that the variances
are equal, so we use the pooled (equal variances) formula.

$$\begin{aligned}
s_p &= \sqrt{\frac{(40-1)(6.8)^2 + (40-1)(7.2)^2}{40+40-2}}
     = \sqrt{\frac{39(46.24) + 39(51.84)}{78}} \\
    &= \sqrt{\frac{1803.36 + 2021.76}{78}}
     = \sqrt{49.041} = 7.003
\end{aligned}$$

$$SE = 7.003 \times \sqrt{\frac{1}{40}+\frac{1}{40}}
= 7.003 \times \sqrt{0.05}
= 7.003 \times 0.2236 = 1.566$$

$$t = \frac{12.4 - 7.1}{1.566} = \frac{5.3}{1.566} = \mathbf{3.384}$$

$df = 40 + 40 - 2 = 78$

**Step 4 — Compare:**

$df = 78$ is not listed in the t-table, so we use the next smaller df, 70:
$t_{0.975,\,70} = 1.994$ (the exact value for $df = 78$ is 1.991).

Since $3.384 > 1.994$, the test statistic falls in the rejection region.
SPSS would report $p = 0.001$, which is $< 0.05$.

**Step 5 — Decision and interpretation:**

We **reject $H_0$**. The mean reduction in systolic blood pressure differs
significantly between the two groups ($t = 3.38$, $df = 78$, $p = 0.001$).
The new drug group had the **larger** mean reduction: 12.4 vs. 7.1 mmHg,
a difference of 5.3 mmHg (95% CI: 2.2 to 8.4 mmHg).
:::

---

## Paired Samples t-Test

### When to Use

The **paired samples t-test** is used when:

- The outcome variable is **continuous (numerical)**
- Observations come in **natural pairs** — the same subject measured twice
  (e.g., before and after treatment), or matched pairs of subjects
- We want to test whether the **mean difference** within pairs is zero
- The **differences** are approximately normally distributed

### Hypotheses

$$H_0: \mu_d = 0 \quad \text{(mean difference is zero)}$$
$$H_1: \mu_d \neq 0 \quad \text{(mean difference is not zero)}$$

Where $\mu_d$ is the population mean of the within-pair differences.

### The Test Statistic

For each pair $i$, compute the difference: $d_i = x_{1i} - x_{2i}$

Then:

$$t = \frac{\bar{d}}{s_d / \sqrt{n}}$$

Where:

- $\bar{d}$ = mean of the differences
- $s_d$ = standard deviation of the differences
- $n$ = number of pairs

The test statistic follows a **t-distribution** with $df = n - 1$.

::: {.callout-tip icon=false}
## Why Not Use the Independent t-Test for Paired Data?
Each patient is compared **with themselves**, so differences between
patients (age, weight, baseline health) cancel out. Working with the
differences $d_i$ turns two columns of data into **one**, and the test
becomes a one-sample test of whether the mean difference is 0.
:::

---

## Example 3: Paired Samples t-Test

A study measures the **diastolic blood pressure (DBP)** of 10 patients
**before** and **after** 4 weeks of a lifestyle intervention.

```{r}
#| label: tbl-paired
#| tbl-cap: "DBP Before and After Lifestyle Intervention (n = 10 patients)"

before <- c(92, 88, 95, 100, 85, 97, 90, 94, 88, 96)
after  <- c(85, 82, 89, 91,  80, 88, 84, 87, 83, 88)
d      <- before - after

pair_tab <- data.frame(
  Patient = paste0("Patient ", 1:10),
  Before  = before,
  After   = after,
  d       = d
)

kable(pair_tab,
      col.names = c("Patient", "DBP Before (mmHg)",
                    "DBP After (mmHg)", "Difference d = Before - After"),
      align = c("l","c","c","c")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = tbl_fs(13)) |>
  column_spec(4, bold = TRUE, color = "#2c7bb6") |>
  row_spec(0, bold = TRUE, color = "white", background = "#2c7bb6")
```

Test at $\alpha = 0.05$ whether the mean DBP changed after the intervention.
Assume the differences are approximately normally distributed.

::: {.callout-tip icon=false}
## Solution

**Step 1 — Hypotheses:**

$H_0: \mu_d = 0$ (no mean change in DBP)

$H_1: \mu_d \neq 0$ (mean DBP has changed)

**Step 2 —** $\alpha = 0.05$

**Step 3 — Compute $\bar{d}$, $s_d$ and the test statistic:**

$$\bar{d} = \frac{`r sum(d)`}{10} = `r mean(d)` \text{ mmHg}$$

$$s_d = \sqrt{\frac{\sum(d_i - \bar{d})^2}{n-1}} = `r round(sd(d), 3)`$$

$$t = \frac{\bar{d}}{s_d/\sqrt{n}} = \frac{`r mean(d)`}{`r round(sd(d),3)`/\sqrt{10}}
= \frac{`r mean(d)`}{`r round(sd(d)/sqrt(10),3)`} = `r round(mean(d)/(sd(d)/sqrt(10)),3)`$$

$df = n - 1 = 9$

**Step 4 — Compare:**

$t_{0.975,\,9} = 2.262$. Since $`r round(mean(d)/(sd(d)/sqrt(10)),3)` > 2.262$,
the test statistic falls in the rejection region.
SPSS would report $p < 0.001$.

**Step 5 — Decision and interpretation:**

We **reject $H_0$**. Mean diastolic blood pressure changed significantly
after the intervention ($t = `r round(mean(d)/(sd(d)/sqrt(10)),2)`$, $df = 9$,
$p < 0.001$). Because $\bar{d}$ = Before − After is positive, DBP
**decreased**, by **`r mean(d)` mmHg** on average.
:::

---

# Statistical Testing on Rank Data

The t-tests need **numerical** data that are approximately **normal**.
When the outcome is **ordinal** (such as a pain score) or the
normality check fails, we use **rank tests** instead. Rank tests replace
each value by its **rank** (1st, 2nd, 3rd, ...) and do not assume any
particular distribution — they are called **non-parametric** tests.

Each t-test has a rank-test partner:

| Design | Numerical, normal | Ordinal, or not normal |
|---|---|---|
| Two **independent** groups | Independent samples t-test | **Mann-Whitney U test** |
| Two **related** (paired) measurements | Paired samples t-test | **Wilcoxon signed-ranks test** |

::: {.callout-important icon=false}
## Rank Tests Use the Z-Table You Already Know
The rank tests have their own statistics ($U$ and $W$), but we do not need a
new table. We convert $U$ or $W$ to a **Z-score** — exactly what SPSS does —
and then use the same rule as in Chapter 3 and Chapter 4:

$$\text{reject } H_0 \text{ if } |z| > z_{0.975} = 1.96 \quad (\alpha = 0.05,\ \text{two-tailed})$$
:::

## Mann-Whitney U Test for Two Independent Groups

### When to Use

The **Mann-Whitney U test** (also called the Wilcoxon rank-sum test) is used when:

- The outcome variable is **ordinal**, or
- The outcome is **continuous but not normally distributed** (especially with
  small samples or obvious skewness)
- We are comparing **two independent groups**

It is the non-parametric alternative to the **independent samples t-test**.

### Hypotheses

$H_0$: the outcome values in the two groups are equal (the two groups come
from the same distribution)

$H_1$: the outcome values in the two groups are different

### How It Works

Instead of comparing means, the Mann-Whitney test compares the **ranks**
of observations between the two groups.

**Steps:**

1. Combine all observations from both groups and **rank them** from
   smallest (rank 1) to largest.
2. If there are **tied values**, assign the average rank to each tied observation.
3. Calculate the **sum of ranks** for each group: $W_1$ and $W_2$.
4. Calculate the test statistic $U$ from the rank sums (formula below).
5. Convert $U$ to a **Z-score** (below) and compare $|z|$ with 1.96.
   SPSS reports this Z-score and its p-value.

::: {.callout-tip icon=false}
## Intuition Behind the Test
If the two groups come from the same distribution, their ranks should be
**randomly mixed** between the groups — neither group should consistently
have higher or lower ranks. A significant result means one group
consistently has higher ranks (i.e., higher values) than the other.
:::

### The U Statistic

$$U_1 = n_1 n_2 + \frac{n_1(n_1+1)}{2} - W_1$$

$$U_2 = n_1 n_2 + \frac{n_2(n_2+1)}{2} - W_2$$

$$U = \min(U_1, U_2)$$

Where $n_1$, $n_2$ are the sample sizes and $W_1$, $W_2$ are the rank sums
for groups 1 and 2 respectively. As a check, $U_1 + U_2 = n_1 n_2$.

A **small** $U$ means the two groups hardly overlap — evidence against $H_0$.

### Converting U to a Z-Score

If $H_0$ is true, $U$ is centred at $n_1 n_2 / 2$ (half of the largest
possible value). We standardise it like any Z-score — *(value − mean) / SD*:

$$\mu_U = \frac{n_1 n_2}{2}, \qquad
\sigma_U = \sqrt{\frac{n_1 n_2 (n_1 + n_2 + 1)}{12}}, \qquad
z = \frac{U - \mu_U}{\sigma_U}$$

- **Decision:** reject $H_0$ if $|z| > 1.96$.
- **p-value (optional):** $p = 2 \times P(Z < -|z|) = 2\,[\,1 - P(Z < |z|)\,]$,
  read from the Z-table.

Because $U = \min(U_1, U_2)$ is the smaller value, $z$ is always zero or
negative. SPSS prints the same Z-score in its *Test Statistics* table.

::: {.callout-note icon=false}
## Why Your Z May Differ Slightly From SPSS
When there are **tied values**, SPSS adjusts $\sigma_U$ for the ties, so its
Z-score can differ from the hand calculation in the second decimal place.
The decision is the same in practice. The Z-approximation is most accurate
when each group has about 10 or more observations; SPSS uses the same
approximation for its *Asymp. Sig.* value.
:::

## Example 4: Mann-Whitney U Test

Pain score (0–10, **ordinal**) was recorded 24 hours after surgery in
**12 patients**: 6 received the standard analgesic regimen and 6 received
a new regimen. The two groups are different patients, so they are
**independent**.

| Regimen | Pain scores |
|---|---|
| Standard | 6, 7, 5, 8, 6, 7 |
| New | 3, 4, 5, 2, 4, 6 |

Test at $\alpha = 0.05$ whether pain scores differ between the two regimens.

::: {.callout-tip icon=false}
## Solution

**Hypotheses:** $H_0$: pain scores in the two groups are equal;
$H_1$: pain scores in the two groups are different. $\alpha = 0.05$.

```{r}
#| label: tbl-mw-ranks
#| tbl-cap: "Pain scores and their ranks (n1 = n2 = 6)"

std_pain <- c(6, 7, 5, 8, 6, 7)
new_pain <- c(3, 4, 5, 2, 4, 6)
all_pain <- c(std_pain, new_pain)
rk       <- rank(all_pain)

mw <- data.frame(
  Group = rep(c("Standard", "New"), each = 6),
  Pain  = all_pain,
  Rank  = rk
)
mw <- mw[order(mw$Pain), ]

kable(mw, row.names = FALSE,
      col.names = c("Group", "Pain score", "Rank (combined)"),
      align = c("l","c","c")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = tbl_fs(12)) |>
  row_spec(0, bold = TRUE, color = "white", background = "#2c7bb6")
```

**Step 1 — Rank all 12 scores together.** Ties share the average rank:
the two 4s occupy ranks 3 and 4, so each gets 3.5; the three 6s occupy
ranks 7, 8 and 9, so each gets 8.

**Step 2 — Sum the ranks in each group:**

$$W_{\text{Standard}} = 8 + 10.5 + 5.5 + 12 + 8 + 10.5 = 54.5$$
$$W_{\text{New}} = 2 + 3.5 + 5.5 + 1 + 3.5 + 8 = 23.5$$

(Check: the total of ranks 1 to 12 is $12 \times 13 / 2 = 78 = 54.5 + 23.5$.)

**Step 3 — Calculate U** with $n_1 = n_2 = 6$:

$$U_1 = 36 + \frac{6 \times 7}{2} - 54.5 = 36 + 21 - 54.5 = 2.5$$
$$U_2 = 36 + 21 - 23.5 = 33.5$$
$$U = \min(2.5,\; 33.5) = 2.5$$

**Step 4 — Convert U to a Z-score** with $n_1 = n_2 = 6$:

$$\mu_U = \frac{6 \times 6}{2} = 18, \qquad
\sigma_U = \sqrt{\frac{6 \times 6 \times 13}{12}} = \sqrt{39} = 6.245$$

$$z = \frac{2.5 - 18}{6.245} = -2.48$$

Since $|z| = 2.48 > 1.96$, the result is significant.
From the Z-table, $p = 2 \times (1 - 0.9934) = 0.013$.

SPSS reports $Z = -2.513$ and $p$ (Asymp. Sig. 2-tailed) $= 0.012$ — the
same decision (the small difference is the tie adjustment).

**Step 5 — Decision.** We **reject $H_0$**.
Pain scores differ significantly between the two regimens. The mean rank
of the new regimen ($23.5 / 6 = 3.9$) is lower than that of the standard
regimen ($54.5 / 6 = 9.1$), so patients on the **new regimen reported
less pain**.
:::

## Wilcoxon Signed-Ranks Test for Two Related Groups

### When to Use

The **Wilcoxon signed-ranks test** (also called the Wilcoxon matched-pairs
test) is the non-parametric alternative to the **paired samples t-test**.
It is used when:

- The outcome variable is **ordinal**, or
- The **differences** between pairs are **not normally distributed**
  (especially with small samples)
- The two sets of measurements are **related (paired)** — for example,
  the same patients measured before and after treatment, or two
  measurements taken on the same blood sample

::: {.callout-important icon=false}
## Mann-Whitney or Wilcoxon?
Both tests work with ranks, so they are easy to confuse. The difference
is the **study design**, exactly as with the two t-tests:

- **Two independent groups** (different people in each group)
  $\rightarrow$ **Mann-Whitney U test**
- **Two related groups** (the same people measured twice)
  $\rightarrow$ **Wilcoxon signed-ranks test**
:::

### Hypotheses

$H_0$: the two related measurements are equal

$H_1$: the two related measurements are different

### How It Works

The Wilcoxon test looks at the **differences within each pair**, and then
ranks the **sizes** of those differences.

**Steps:**

1. For each pair, calculate the difference $d_i = x_{i,1} - x_{i,2}$.
2. **Discard** any pair whose difference is zero.
3. Rank the **absolute** differences $|d_i|$ from smallest to largest,
   assigning average ranks to ties.
4. Attach to each rank the **sign** of its original difference.
5. Add the ranks belonging to the positive differences ($W^{+}$) and the
   ranks belonging to the negative differences ($W^{-}$).

### The Test Statistic

$$W = \min(W^{+},\; W^{-})$$

If the two conditions really are the same, the positive and negative
ranks should roughly balance, so $W^{+}$ and $W^{-}$ should be similar.
A **small** value of $W$ means the differences fall mostly in one
direction, which is evidence against $H_0$.

### Converting W to a Z-Score

$n$ is the number of pairs **after discarding zero differences**.
If $H_0$ is true, $W$ is centred at half of the total of all ranks:

$$\mu_W = \frac{n(n+1)}{4}, \qquad
\sigma_W = \sqrt{\frac{n(n+1)(2n+1)}{24}}, \qquad
z = \frac{W - \mu_W}{\sigma_W}$$

As with Mann-Whitney: **reject $H_0$ if $|z| > 1.96$**, and SPSS reports
this Z-score (adjusted for ties) in its *Test Statistics* table.

::: {.callout-tip icon=false}
## Reading the SPSS Output
Choose **Analyze** $\rightarrow$ **Nonparametric Tests** $\rightarrow$
**Legacy Dialogs** $\rightarrow$ **2 Related Samples**, and tick **Wilcoxon**
(older SPSS versions, as in the lab manual, have no *Legacy Dialogs* step).
The *Ranks* table shows the sum of the negative and positive ranks, and the
*Test Statistics* table reports $Z$ and the 2-tailed significance.
As always, decide by comparing the p-value with $\alpha$ — the sign of $Z$
only indicates the direction of the ranks, not the importance of the result.
:::

## Example 5: Wilcoxon Signed-Ranks Test

Blood clotting time (minutes) was measured for **8 patients** using two
solutions: an extract from shrimp shells ("Shrimp") and the standard
solution ("Standard"). Blood was taken from each patient and divided into two parts,
so the two measurements are **paired**.

| Patient | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 |
|---|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| Shrimp | 3 | 5 | 5 | 4 | 3 | 7 | 8 | 7 |
| Standard | 7 | 6 | 3 | 8 | 5 | 9 | 7 | 9 |

The sample is small ($n = 8$), so we cannot rely on the normality of the
differences. Test at $\alpha = 0.05$ whether the clotting times differ,
using the **Wilcoxon signed-ranks test**.

::: {.callout-tip icon=false}
## Solution

**Hypotheses:** $H_0$: the clotting times with the two solutions are equal;
$H_1$: they are different. $\alpha = 0.05$.

```{r}
#| label: tbl-wilcoxon
#| tbl-cap: "Blood clotting time using two solutions (n = 8 patients), with signed ranks"

pt   <- 1:8
sol1 <- c(3, 5, 5, 4, 3, 7, 8, 7)
gsol <- c(7, 6, 3, 8, 5, 9, 7, 9)
dif  <- sol1 - gsol
rk_abs <- rank(abs(dif))
signed <- ifelse(dif > 0, rk_abs, -rk_abs)

wil <- data.frame(Patient = pt, Shrimp = sol1, Standard = gsol, d = dif,
                  absd = abs(dif), rk = rk_abs, srk = signed)

kable(wil,
      col.names = c("Patient", "Shrimp", "Standard",
                    "d", "abs(d)", "Rank of abs(d)", "Signed rank"),
      align = rep("c", 7)) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = tbl_fs(12)) |>
  row_spec(0, bold = TRUE, color = "white", background = "#2c7bb6")
```

**Step 1 — Differences** $d = \text{Shrimp} - \text{Standard}$: none is zero,
so all 8 pairs are kept.

**Step 2 — Rank $|d|$ with average ranks for ties:**
the two 1s share ranks 1–2 (each 1.5); the four 2s share ranks 3–6
(each 4.5); the two 4s share ranks 7–8 (each 7.5).

**Step 3 — Sum the signed ranks:**

$$W^{+} = 4.5 \;(\text{patient } 3) + 1.5 \;(\text{patient } 7) = 6$$
$$W^{-} = 7.5 + 1.5 + 7.5 + 4.5 + 4.5 + 4.5 = 30 \quad (\text{patients } 1, 2, 4, 5, 6, 8)$$

(Check: $W^{+} + W^{-} = 8 \times 9 / 2 = 36$.)

$$W = \min(6,\; 30) = 6$$

**Step 4 — Convert W to a Z-score** with $n = 8$ pairs:

$$\mu_W = \frac{8 \times 9}{4} = 18, \qquad
\sigma_W = \sqrt{\frac{8 \times 9 \times 17}{24}} = \sqrt{51} = 7.141$$

$$z = \frac{6 - 18}{7.141} = -1.68$$

Since $|z| = 1.68 \leq 1.96$, the result is not significant.
From the Z-table, $p = 2 \times (1 - 0.9535) = 0.093$.

SPSS reports $Z = -1.706$ and $p = 0.088$ — the same decision.
*(SPSS calculates the difference as Standard − Shrimp, so its Ranks table
lists the 6 and the 30 the other way round; $W$ and the p-value are the same.)*

**Step 5 — Decision.** We **do not reject** $H_0$. There is not enough
evidence to say that the clotting times produced by the two solutions
differ, at the 0.05 level of significance.
:::

---

## Choosing the Right Test

```{r}
#| label: tbl-test-choice
#| tbl-cap: "Guide to Choosing the Correct Statistical Test (Two Groups)"

tests <- data.frame(
  `Data Type` = c("Categorical (Nominal)",
                   "Continuous (normal)",
                   "Continuous (normal)",
                   "Ordinal or non-normal continuous",
                   "Ordinal or non-normal continuous"),
  `Groups` = c("2 independent groups",
                "2 independent groups",
                "2 related (paired) groups",
                "2 independent groups",
                "2 related (paired) groups"),
  `Correct Test` = c("Chi-square test",
                      "Independent samples t-test",
                      "Paired samples t-test",
                      "Mann-Whitney U test",
                      "Wilcoxon signed-ranks test"),
 `SPSS Procedure` = c("Analyze -> Descriptive Statistics -> Crosstabs",
 "Analyze -> Compare Means -> Independent-Samples T Test",
 "Analyze -> Compare Means -> Paired-Samples T Test",
 "Analyze -> Nonparametric Tests -> Legacy Dialogs -> 2 Independent Samples",
 "Analyze -> Nonparametric Tests -> Legacy Dialogs -> 2 Related Samples")
)

kable(tests,
      col.names = c("Outcome Data Type", "Study Design",
                    "Statistical Test", "SPSS Procedure"),
      align = c("l","l","l","l")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = tbl_fs(12)) |>
  column_spec(1, width = pdf_w("1.2in")) |>
  column_spec(2, width = pdf_w("1.2in")) |>
  column_spec(3, bold = TRUE, color = "#2c7bb6", width = pdf_w("1.3in")) |>
  column_spec(4, width = pdf_w("1.7in"))
```

```{r}
#| label: fig-test-flowchart
#| fig-cap: "Decision flowchart for selecting the correct test (two-group comparisons)"
#| fig-align: center
#| fig-height: 5.4
#| fig-width: 9

boxes <- data.frame(
  x = c(5.5, 5.5, 2.0, 8.0, 2.0, 8.0, 5.6, 9.4, 4.2, 6.6, 8.6, 10.4),
  y = c(9.2, 7.7, 6.1, 6.1, 4.3, 4.3, 2.6, 2.6, 0.8, 0.8, 0.8, 0.8),
  label = c("Comparing 2 Groups", "Outcome variable type?",
            "Nominal\n(Categorical)", "Continuous\nor Ordinal",
            "Chi-square\nTest", "Groups independent\nor paired?",
            "Continuous\nand normal?", "Continuous\nand normal?",
            "Independent\nt-test", "Mann-Whitney\nU Test",
            "Paired\nt-test", "Wilcoxon\nsigned-ranks"),
  fill = c("#dce8f5","#dce8f5","#abd9e9","#abd9e9","#2c7bb6","#dce8f5",
           "#dce8f5","#dce8f5","#1a9641","#fdae61","#d7191c","#fdae61"),
  col  = c("black","black","black","black","white","black",
           "black","black","white","black","white","black")
)

seg <- data.frame(
  # 6th and 7th rows: "Groups independent or paired?" (x = 8.0) branches
  # left to the Independent node (5.6) and right to the Paired node (9.4)
  x    = c(5.5, 5.5, 5.5, 2.0, 8.0, 8.0, 8.0, 5.6, 5.6, 9.4, 9.4),
  xend = c(5.5, 2.2, 7.8, 2.0, 8.0, 5.8, 9.2, 4.4, 6.5, 8.7, 10.3),
  y    = c(8.9, 7.35, 7.35, 5.6, 5.6, 3.9, 3.9, 2.2, 2.2, 2.2, 2.2),
  yend = c(8.0, 6.6,  6.6,  4.7, 4.7, 3.0, 3.0, 1.2, 1.2, 1.2, 1.2)
)

edge_lab <- data.frame(
  x = c(3.4, 7.7, 6.4, 9.3, 4.3, 6.9, 8.5, 10.8),
  y = c(7.1, 7.1, 3.6, 3.6, 1.8, 1.8, 1.8, 1.8),
  label = c("Nominal","Continuous/\nOrdinal","Independent","Paired",
            "Yes","No /\nOrdinal","Yes","No /\nOrdinal")
)

ggplot() +
  geom_segment(data = seg, aes(x = x, xend = xend, y = y, yend = yend),
               arrow = arrow(length = unit(0.18, "cm")), color = "gray40") +
  geom_text(data = edge_lab, aes(x = x, y = y, label = label),
            size = 2.6, color = "gray30") +
  geom_label(data = boxes,
             aes(x = x, y = y, label = label, fill = fill, color = col),
             size = 2.8, label.r = unit(0.3, "lines"),
             label.padding = unit(0.35, "lines"), show.legend = FALSE) +
  scale_fill_identity() + scale_color_identity() +
  scale_x_continuous(limits = c(0.6, 11.6)) +
  scale_y_continuous(limits = c(0.1, 9.8)) +
  theme_void()
```

::: {.callout-warning icon=false}
## Ordinal Outcomes Always Go to a Rank Test
An ordinal outcome — such as a **pain score**, a satisfaction rating or a
stage of disease — is never analysed with a t-test, even if it is
written as numbers. Use the Mann-Whitney U test (independent groups)
or the Wilcoxon signed-ranks test (paired).
:::

---

# Key Takeaways

::: {.callout-note icon=false}
## Chapter 5 Summary

1. Hypothesis testing uses sample data to decide between two competing
   claims: the **null hypothesis ($H_0$)** and the
   **alternative hypothesis ($H_1$)**.

2. The **test statistic** measures how far the data are from $H_0$.
   By hand we compare it with a **critical value**; in SPSS we compare the
   **p-value** with $\alpha$. Both give the same decision.

3. The **p-value** is the probability of the observed data (or more extreme)
   if $H_0$ is true. A p-value $\leq \alpha$ (usually 0.05) leads us to
   **reject $H_0$**.

4. **Type I error (α):** rejecting a true $H_0$ (false positive).
   **Type II error (β):** failing to reject a false $H_0$ (false negative).

5. **Chi-square test:** for association between two categorical (nominal)
   variables in independent groups. Read the Pearson row, or Fisher's exact
   test if any expected count is below 5.

6. **Independent samples t-test:** for comparing means of two independent
   groups with continuous, approximately normal data. Read **Levene's test**
   first to choose the row.

7. **Paired samples t-test:** for comparing means of two related measurements
   (before/after, or matched pairs) when the differences are approximately
   normal.

8. **Mann-Whitney U test:** for comparing two independent groups when the
   outcome is ordinal or the normality assumption is not met.

9. **Wilcoxon signed-ranks test:** for comparing two **related** measurements
   when the outcome is ordinal or the differences are not normally
   distributed. It is the non-parametric alternative to the paired t-test.
   For both rank tests, convert $U$ or $W$ to a **Z-score** and
   **reject $H_0$ if $|z| > 1.96$** — the same Z-score that SPSS reports.

10. Always choose the test based on **data type** and **study design** —
    using the wrong test leads to invalid conclusions.
:::

---

## Formula Summary

| Quantity | Formula | Notes |
|:----------------|:----------------------------------------|:----------------------|
| Decision with SPSS | Reject $H_0$ if $p \leq \alpha$ | Usually $\alpha = 0.05$ |
| Expected frequency | $E = \dfrac{\text{Row total} \times \text{Column total}}{n}$ | All $E \geq 5$, otherwise Fisher's exact test |
| Chi-square | $\chi^2 = \sum \dfrac{(O - E)^2}{E}$ | $df = (r-1)(c-1)$; reject if $\chi^2 > \chi^2_{0.95,\,df}$ |
| Pooled SD | $s_p = \sqrt{\dfrac{(n_1-1)s_1^2 + (n_2-1)s_2^2}{n_1 + n_2 - 2}}$ | Equal variances (Levene Sig. $> 0.05$) |
| Independent t | $t = \dfrac{\bar{x}_1 - \bar{x}_2}{s_p \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}}$ | $df = n_1 + n_2 - 2$ |
| Welch SE | $SE = \sqrt{\dfrac{s_1^2}{n_1} + \dfrac{s_2^2}{n_2}}$ | Unequal variances; df from SPSS |
| Paired t | $t = \dfrac{\bar{d}}{s_d / \sqrt{n}}$ | $d_i = x_{1i} - x_{2i}$; $df = n - 1$ |
| Decision for t | Reject $H_0$ if $|t| > t_{0.975,\,df}$ | Two-tailed, $\alpha = 0.05$ |
| Mann-Whitney U | $U_1 = n_1 n_2 + \dfrac{n_1(n_1+1)}{2} - W_1$, $U = \min(U_1, U_2)$ | $W_1$ = rank sum of group 1 |
| U to Z | $z = \dfrac{U - n_1 n_2 / 2}{\sqrt{n_1 n_2 (n_1 + n_2 + 1)/12}}$ |  |
| Wilcoxon W | $W = \min(W^{+}, W^{-})$ | Drop zero differences first |
| W to Z | $z = \dfrac{W - n(n+1)/4}{\sqrt{n(n+1)(2n+1)/24}}$ | $n$ = number of non-zero pairs |
| Decision for U and W | Reject $H_0$ if $|z| > 1.96$ | Same Z as SPSS (up to ties) |

: Formulas in this chapter {.striped}

---

# Exercises

::: {.callout-caution icon=false}
## Exercise 1 — Chi-Square Test

A researcher investigates whether **physical activity level** is associated
with **diabetes status** in 300 adults:

|  | Diabetes | No Diabetes | Total |
|---|:---:|:---:|:---:|
| Active | 18 | 132 | 150 |
| Inactive | 42 | 108 | 150 |
| **Total** | **60** | **240** | **300** |

**(a)** State $H_0$ and $H_1$.

**(b)** Calculate all expected frequencies and verify the chi-square
condition is met.

**(c)** Calculate the chi-square test statistic.

**(d)** At $\alpha = 0.05$, what is your conclusion?

**(e)** In plain language, what does your result mean?
:::

::: {.callout-caution icon=false}
## Exercise 2 — Independent Samples t-Test

A study compares the **mean hemoglobin levels (g/dL)** between
male and female patients:

| Group | n | $\bar{x}$ | s |
|---|:---:|:---:|:---:|
| Male | 35 | 14.8 | 1.6 |
| Female | 35 | 12.9 | 1.4 |

In SPSS, Levene's test for these data gives Sig. = 0.41.

**(a)** State $H_0$ and $H_1$ (two-tailed).

**(b)** Based on Levene's test, should you use the "Equal variances assumed"
or the "Equal variances not assumed" version? Why?

**(c)** Calculate the pooled standard error.

**(d)** Calculate the t test statistic.

**(e)** At $\alpha = 0.05$ with appropriate df, state your conclusion.
:::

::: {.callout-caution icon=false}
## Exercise 3 — Paired Samples t-Test

A diabetes clinic measures the **capillary blood glucose (mg/dL)** of
8 patients with type 2 diabetes **before** and **2 hours after** a
structured 30-minute walking session:

| Patient | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 |
|---|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| Before | 182 | 165 | 210 | 148 | 195 | 172 | 160 | 188 |
| After  | 160 | 150 | 182 | 141 | 170 | 158 | 152 | 165 |

Assume the differences are approximately normally distributed.

**(a)** Calculate the difference $d_i$ = Before − After for each patient.

**(b)** Calculate $\bar{d}$ and $s_d$.

**(c)** Calculate the paired t statistic.

**(d)** At $\alpha = 0.05$, did blood glucose change significantly after
the walking session? In which direction? State your conclusion clearly.

**(e)** Which variable would you check for normality in SPSS before using
this test, and which SPSS procedure would you use?
:::

::: {.callout-caution icon=false}
## Exercise 4 — Mann-Whitney U Test

A physiotherapy team records the **time to first walk after knee surgery
(hours)**, which is strongly right-skewed, for two independent groups of
patients:

| Group | Time to first walk (hours) |
|---|---|
| Early mobilisation programme | 12, 18, 9, 30, 15 |
| Usual care | 22, 35, 28, 40, 26, 48 |

**(a)** Explain why the Mann-Whitney U test is used rather than the
independent samples t-test.

**(b)** Rank all 11 values together and find the rank sums $W_1$ and $W_2$.

**(c)** Calculate $U_1$, $U_2$ and $U$.

**(d)** Convert $U$ to a Z-score and state your conclusion at
$\alpha = 0.05$. Use the Z-table to find the approximate p-value.
:::

::: {.callout-caution icon=false}
## Exercise 5 — Wilcoxon Signed-Ranks Test

An anxiety score (0–20, **ordinal**) was recorded for 9 patients before and
after a pre-operative education session:

| Patient | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 |
|---|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| Before | 14 | 11 | 16 | 9 | 13 | 15 | 12 | 10 | 17 |
| After | 9 | 10 | 11 | 8 | 6 | 12 | 4 | 13 | 7 |

**(a)** Calculate $d$ = Before − After for each patient.

**(b)** Rank $|d|$ (use average ranks for ties) and find $W^{+}$, $W^{-}$
and $W$.

**(c)** Convert $W$ to a Z-score and state your conclusion at
$\alpha = 0.05$. Did anxiety go up or down?
:::

::: {.callout-caution icon=false}
## Exercise 6 — Choosing the Right Test

For each scenario below, state which statistical test is most appropriate
and briefly explain why:

**(a)** Comparing the proportion of post-operative complications between
patients who received standard care vs. enhanced recovery protocol.

**(b)** Comparing patient satisfaction scores (rated: Poor / Fair / Good /
Excellent) between two hospital wards.

**(c)** Measuring the effect of a new diet on body weight by recording
each participant's weight before and after a 12-week program
(the differences are approximately normal).

**(d)** Comparing the mean length of hospital stay (days) between two
age groups (under 60 vs. 60 and over), where length-of-stay data is
strongly right-skewed.

**(e)** Comparing pain scores (0–10, ordinal) recorded for the **same**
patients before and 24 hours after starting a new analgesic.
:::

---

*End of Chapter 5*

---

::: {.callout-note icon=false}
## Coming Up Next
**Chapter 6: Introduction to Statistical Modeling** — We take our first
steps beyond comparing two groups, exploring how one variable can be used
to explain or predict another through **linear regression**.
:::
