# Introduction to Statistical Modeling

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(knitr)
library(kableExtra)
library(patchwork)
# Table helpers (used by the book and the slides)
# tbl_fs(): font size; PDF tables use 10pt so they fit the page width
# pdf_w():  fixed column width in the PDF only, so long text wraps
tbl_fs <- function(html_size = 13) if (knitr::is_latex_output()) 10 else html_size
pdf_w  <- function(x) if (knitr::is_latex_output()) x else NULL
```

---

# Introduction {.unnumbered}

::: {.callout-note icon=false}
## Why Does This Matter?

So far we have described data, estimated population parameters, and tested
whether two groups differ. But many real-world health questions go beyond
simple comparisons:

- Does body weight *predict* blood pressure?
- As age *increases*, how does cholesterol *change*?
- Can we use a patient's BMI to *estimate* their fasting glucose level?

These questions require us to model the **relationship between two variables**.
**Statistical modeling** is the process of building a mathematical equation
that describes how one variable is related to another — and using that
equation to understand, explain, and predict outcomes.

In this chapter we focus on the simplest and most widely used model:
**simple linear regression**.
:::

---

# What Is a Statistical Model?

A **statistical model** is a mathematical equation that describes the
relationship between:

- An **outcome variable** (also called the *dependent variable*, *response
  variable*, or $Y$) — the variable we want to explain or predict.
- One or more **predictor variables** (also called *independent variables*,
  *explanatory variables*, or $X$) — the variables we use to explain or
  predict $Y$.

In this chapter we study **simple linear regression** — one outcome $Y$
and one predictor $X$, with a straight-line relationship.

---

# Correlation: Measuring the Relationship

Before building a model, we first ask: *Is there a relationship between
$X$ and $Y$, and how strong is it?*

## The Scatter Plot

A **scatter plot** displays pairs of values $(x_i, y_i)$ as points on a
graph. The pattern of points reveals the **direction** and **strength**
of the relationship.

```{r}
#| label: fig-scatter-types
#| fig-cap: "Four possible relationships between two variables"
#| fig-align: center
#| fig-height: 4
#| fig-width: 10

set.seed(42)
n <- 80

d1 <- data.frame(x = rnorm(n), y = 0.8*rnorm(n, mean=0) + rnorm(n,sd=0.5),
                 type = "Strong positive")
d1$y <- d1$x * 0.9 + rnorm(n, sd=0.3)

d2 <- data.frame(x = rnorm(n), y = -0.9*rnorm(n) + rnorm(n,sd=0.3),
                 type = "Strong negative")
d2$y <- -d2$x * 0.9 + rnorm(n, sd=0.3)

d3 <- data.frame(x = rnorm(n), y = 0.4*rnorm(n) + rnorm(n,sd=0.9),
                 type = "Weak positive")
d3$y <- d3$x * 0.3 + rnorm(n, sd=0.9)

d4 <- data.frame(x = rnorm(n), y = rnorm(n),
                 type = "No relationship")

all_d <- rbind(d1, d2, d3, d4)
all_d$type <- factor(all_d$type,
                     levels = c("Strong positive","Strong negative",
                                "Weak positive","No relationship"))

ggplot(all_d, aes(x = x, y = y)) +
  geom_point(alpha = 0.5, color = "#2c7bb6", size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#d7191c",
              linewidth = 1, formula = y ~ x) +
  facet_wrap(~ type, nrow = 1, scales = "free") +
  theme_minimal(base_size = 12) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        strip.text = element_text(face = "bold", size = 11))
```

---

## Pearson's Correlation Coefficient ($r$)

The **Pearson correlation coefficient** $r$ quantifies the **strength**
and **direction** of the *linear* relationship between two continuous variables.

$$r = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}
           {\sqrt{\sum_{i=1}^{n}(x_i-\bar{x})^2 \cdot \sum_{i=1}^{n}(y_i-\bar{y})^2}}$$

### Properties of $r$

- $r$ always lies between **−1 and +1**
- $r = +1$: perfect positive linear relationship
- $r = -1$: perfect negative linear relationship
- $r = 0$: no linear relationship

### Interpreting the Magnitude of $r$

```{r}
#| label: tbl-r-interpret
#| tbl-cap: "General guidelines for interpreting the correlation coefficient"

r_tab <- data.frame(
  `|r| range` = c("0.00 - 0.19", "0.20 - 0.39",
                   "0.40 - 0.59", "0.60 - 0.79", "0.80 - 1.00"),
  Strength = c("Negligible", "Weak", "Moderate", "Strong", "Very strong"),
  Direction = c("-", "Positive or negative depending on sign",
                "Positive or negative depending on sign",
                "Positive or negative depending on sign",
                "Positive or negative depending on sign")
)

kable(r_tab, col.names = c("Size of r (ignoring sign)", "Strength", "Note"),
      align = c("c","l","l")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = tbl_fs(13)) |>
  column_spec(1, bold = TRUE, color = "#2c7bb6")
```

### Testing Whether $r$ Is Significant

A sample correlation can be different from 0 just by chance, especially in
a small sample. We test the **population correlation** $\rho$ ("rho"):

$$H_0: \rho = 0 \quad \text{(no linear relationship in the population)}$$
$$H_1: \rho \neq 0 \quad \text{(there is a linear relationship)}$$

The test statistic is

$$t = \frac{r\sqrt{n-2}}{\sqrt{1-r^2}}, \qquad df = n - 2$$

In SPSS (**Analyze** $\rightarrow$ **Correlate** $\rightarrow$ **Bivariate**),
the **Correlations** table shows three numbers for each pair of variables:

| | | BMI | Glucose |
|---|---|:---:|:---:|
| BMI | Pearson Correlation | 1 | .612 |
| | Sig. (2-tailed) | | .000 |
| | N | 60 | 60 |

- **Pearson Correlation** = $r$ = 0.612 (strong positive)
- **Sig. (2-tailed)** = the p-value. SPSS shows **.000** when $p < 0.0005$;
  write it as $p < 0.001$, never as $p = 0$.
- **N** = number of pairs

Since $p < 0.05$, we reject $H_0$: there is a significant positive linear
relationship between BMI and fasting glucose.

::: {.callout-warning icon=false}
## Correlation Is Not Causation

A high correlation between $X$ and $Y$ does **not** mean that $X$
*causes* $Y$. Both variables may be influenced by a third variable,
or the relationship may be coincidental. Always interpret correlations
in the context of biological or clinical plausibility.
:::

---

# Simple Linear Regression

## The Regression Equation

Simple linear regression models the relationship between $X$ and $Y$
as a straight line:

$$\hat{Y} = b_0 + b_1 X$$

Where:

| Symbol | Name | Meaning |
|:---:|---|---|
| $\hat{Y}$ | Predicted value | The estimated value of $Y$ for a given $X$ |
| $b_0$ | Intercept | The predicted value of $Y$ when $X = 0$ |
| $b_1$ | Slope | The change in $\hat{Y}$ for each one-unit increase in $X$ |

The line above is calculated from a **sample**. It estimates the true
line in the **population**:

$$Y = \beta_0 + \beta_1 X + \varepsilon$$

where $\beta_0$ and $\beta_1$ (Greek "beta") are the population intercept
and slope — **parameters** — and $\varepsilon$ is random error.
The sample values $b_0$ and $b_1$ are the **statistics** we use to estimate
them, just as $\bar{x}$ estimates $\mu$ (Chapter 1).

For each patient, the **residual** $e_i$ is the difference between the
observed value and the value predicted by the sample line:

$$e_i = Y_i - \hat{Y}_i$$

---

## Estimating the Coefficients

The values of $b_0$ and $b_1$ are chosen to minimize the **sum of squared
residuals** — this method is called **Ordinary Least Squares (OLS)**:

$$\text{Minimize: } \sum_{i=1}^{n}(Y_i - \hat{Y}_i)^2 = \sum_{i=1}^{n} e_i^2$$

The OLS formulas are:

$$b_1 = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}
             {\sum_{i=1}^{n}(x_i - \bar{x})^2}
       = r \cdot \frac{s_y}{s_x}$$

$$b_0 = \bar{y} - b_1 \bar{x}$$

```{r}
#| label: fig-ols
#| fig-cap: "Least squares regression line minimizes the sum of squared residuals (vertical red lines)"
#| fig-align: center
#| fig-height: 4
#| fig-width: 6

set.seed(10)
x <- seq(20, 80, length = 20)
y <- 60 + 0.9 * x + rnorm(20, sd = 10)
fit <- lm(y ~ x)
df_ols <- data.frame(x = x, y = y,
                     yhat = fitted(fit),
                     resid = residuals(fit))

ggplot(df_ols, aes(x = x, y = y)) +
  geom_segment(aes(xend = x, yend = yhat),
               color = "#d7191c", linewidth = 0.7, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, color = "#2c7bb6",
              linewidth = 1.3, formula = y ~ x) +
  geom_point(size = 3, color = "#2c7bb6", alpha = 0.8) +
  labs(x = "X (Predictor)", y = "Y (Outcome)",
       caption = "Red dashed lines = residuals (e = Y − Ŷ)") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

---

## The Coefficient of Determination ($R^2$)

**$R^2$** (R-squared) measures how well the regression line fits the data —
specifically, the **proportion of the total variation in $Y$ that is
explained by $X$**:

$$R^2 = r^2 \quad \text{(in simple linear regression)}$$

$$R^2 = \frac{\text{Variation explained by the model}}{\text{Total variation in } Y}$$

- $R^2 = 1$: the model explains 100% of the variation in $Y$ (perfect fit)
- $R^2 = 0$: the model explains none of the variation in $Y$
- $R^2 = 0.65$ means: *"65% of the variation in $Y$ is explained by $X$"*

---

## Example 1: Calculating Correlation and Regression

A researcher records the **age (years)** and **systolic blood pressure (mmHg)**
of 8 adults:

```{r}
#| label: tbl-ex1-data
#| tbl-cap: "Age and Systolic Blood Pressure of 8 Adults"

age <- c(25, 32, 40, 48, 55, 62, 70, 78)
sbp <- c(118, 116, 131, 128, 146, 139, 160, 157)

ex1 <- data.frame(Subject = paste0("S", 1:8), Age = age, SBP = sbp)

kable(ex1, col.names = c("Subject", "Age (years)", "SBP (mmHg)"),
      align = c("c","c","c")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = tbl_fs(13))
```

**(a)** Calculate Pearson's $r$.

**(b)** Find the regression equation $\hat{Y} = b_0 + b_1 X$.

**(c)** Predict the SBP of a 50-year-old adult.

**(d)** Interpret $R^2$.

::: {.callout-tip icon=false}
## Solution

```{r}
#| label: tbl-ex1-calc
#| tbl-cap: "Calculation Table for Correlation and Regression"

xbar <- mean(age)
ybar <- mean(sbp)

calc <- data.frame(
  Subject = paste0("S", 1:8),
  x = age,
  y = sbp,
  xdev = round(age - xbar, 2),
  ydev = round(sbp - ybar, 2),
  xdev2 = round((age - xbar)^2, 2),
  ydev2 = round((sbp - ybar)^2, 2),
  xydev = round((age - xbar)*(sbp - ybar), 2)
)

totrow <- data.frame(
  Subject = "**Total**",
  x = NA, y = NA,
  xdev = NA, ydev = NA,
  xdev2 = round(sum((age-xbar)^2), 2),
  ydev2 = round(sum((sbp-ybar)^2), 2),
  xydev = round(sum((age-xbar)*(sbp-ybar)), 2)
)

calc_all <- rbind(calc, totrow)

kable(calc_all,
      col.names = c("Subject","x","y","x-xbar","y-ybar",
                    "(x-xbar)2","(y-ybar)2","(x-xbar)(y-ybar)"),
      align = rep("c", 8), na_print = "-") |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = tbl_fs(12)) |>
  row_spec(9, bold = TRUE) |>
  row_spec(0, bold = TRUE, color = "white", background = "#2c7bb6")
```

**(a) Pearson's r:**

$$r = \frac{\sum(x_i-\bar{x})(y_i-\bar{y})}
           {\sqrt{\sum(x_i-\bar{x})^2 \cdot \sum(y_i-\bar{y})^2}}
  = \frac{`r round(sum((age-xbar)*(sbp-ybar)),2)`}
         {\sqrt{`r round(sum((age-xbar)^2),2)` \times `r round(sum((sbp-ybar)^2),2)`}}
  = \frac{`r round(sum((age-xbar)*(sbp-ybar)),2)`}
         {`r round(sqrt(sum((age-xbar)^2)*sum((sbp-ybar)^2)),2)`}
  = \mathbf{`r round(cor(age,sbp),3)`}$$

This indicates a **very strong positive** linear relationship between
age and systolic blood pressure. *(Real clinical data are rarely this
tidy — with only 8 adults, a correlation this high is unusual.)*

**(b) Regression coefficients:**

$$b_1 = \frac{\sum(x_i-\bar{x})(y_i-\bar{y})}{\sum(x_i-\bar{x})^2}
      = \frac{`r round(sum((age-xbar)*(sbp-ybar)),2)`}
             {`r round(sum((age-xbar)^2),2)`}
      = `r round(coef(lm(sbp~age))[2],3)`$$

$$b_0 = \bar{y} - b_1\bar{x}
      = `r round(ybar,2)` - `r round(coef(lm(sbp~age))[2],3)` \times `r round(xbar,2)`
      = `r round(coef(lm(sbp~age))[1],2)`$$

$$\boxed{\hat{Y} = `r round(coef(lm(sbp~age))[1],2)` + `r round(coef(lm(sbp~age))[2],3)` X}$$

**(c) Prediction for age = 50:**

$$\hat{Y} = `r round(coef(lm(sbp~age))[1],2)` + `r round(coef(lm(sbp~age))[2],3)` \times 50
          = `r round(predict(lm(sbp~age), newdata=data.frame(age=50)),1)` \text{ mmHg}$$

**(d) $R^2$:**

$$R^2 = r^2 = (`r round(cor(age,sbp),3)`)^2 = `r round(cor(age,sbp)^2,3)`$$

**`r round(cor(age,sbp)^2*100,1)`% of the variation in systolic blood pressure
is explained by age.** The remaining `r round((1-cor(age,sbp)^2)*100,1)`%
is due to other factors not included in this model.
:::

```{r}
#| label: fig-ex1-fit
#| fig-cap: "Scatter plot with fitted regression line: Age vs. Systolic Blood Pressure"
#| fig-align: center
#| fig-height: 4
#| fig-width: 6

fit1 <- lm(sbp ~ age)
df1  <- data.frame(age = age, sbp = sbp)

ggplot(df1, aes(x = age, y = sbp)) +
  geom_point(size = 3.5, color = "#2c7bb6") +
  geom_smooth(method = "lm", se = TRUE, color = "#d7191c",
              fill = "#fdae61", alpha = 0.2, formula = y ~ x) +
  annotate("text", x = 30, y = 162,
           label = paste0("Ŷ = ", round(coef(fit1)[1],2),
                          " + ", round(coef(fit1)[2],3), "X\n",
                          "r = ", round(cor(age,sbp),3),
                          ",  R² = ", round(cor(age,sbp)^2,3)),
           hjust = 0, size = 4, color = "#d7191c") +
  labs(x = "Age (years)", y = "Systolic Blood Pressure (mmHg)") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

---

## Testing the Significance of the Regression

Before using the regression equation for interpretation or prediction,
we should test whether the population slope $\beta_1$ is different from zero.
If $\beta_1 = 0$, then $X$ provides no information about $Y$ — the line
is flat.

**Hypotheses:**

$$H_0: \beta_1 = 0 \quad \text{(X is not a useful predictor of Y)}$$
$$H_1: \beta_1 \neq 0 \quad \text{(X is a useful predictor of Y)}$$

The test statistic is:

$$t = \frac{b_1}{SE_{b_1}}$$

which follows a t-distribution with $df = n - 2$.

SPSS reports this p-value automatically in the **Coefficients table**
of the regression output — read the row of the **predictor**, not the
Constant row. In simple linear regression this test gives exactly the
same p-value as the test of Pearson's $r$ above: both ask whether there is
a linear relationship between $X$ and $Y$.

---

## Assumptions of Linear Regression

The regression model is valid only when the following assumptions are met:

```{r}
#| label: tbl-assumptions
#| tbl-cap: "Assumptions of Simple Linear Regression"

assump <- data.frame(
  Assumption = c("Linearity",
                 "Independence",
                 "Homoscedasticity",
                 "Normality of residuals"),
  Meaning = c(
    "The relationship between X and Y is linear (a straight line is appropriate)",
    "Observations are independent of each other",
    "The spread of residuals is roughly constant across all values of X",
    "Residuals are approximately normally distributed"
  ),
  `How to Check` = c(
    "Scatter plot of X vs. Y; residual plot",
    "Study design - satisfied by random sampling",
    "Plot of residuals vs. fitted values - no funnel shape",
    "Histogram or Q-Q plot of residuals"
  )
)

kable(assump, col.names = c("Assumption", "Meaning", "How to Check"),
      align = c("l","l","l")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = tbl_fs(13)) |>
  column_spec(1, bold = TRUE, color = "#2c7bb6", width = pdf_w("1.3in")) |>
  column_spec(2, width = pdf_w("2.2in")) |>
  column_spec(3, width = pdf_w("2.0in"))
```

---

## Reading SPSS Regression Output

A researcher studies whether **waist circumference (cm)** ($X$) predicts
**fasting blood glucose (mg/dL)** ($Y$) in **40 adults**
(**Analyze** $\rightarrow$ **Regression** $\rightarrow$ **Linear**).
SPSS produces three key tables. Here is how to read them:

```{r}
#| label: tbl-spss-model
#| tbl-cap: "SPSS Output: Model Summary"

model_sum <- data.frame(
  R = "0.821",
  R2 = "0.674",
  `Adjusted R2` = "0.666",
  `Std. Error of the Estimate` = "8.458"
)

kable(model_sum,
      col.names = c("R", "R2", "Adjusted R2", "Std. Error of the Estimate"),
      align = rep("c",4)) |>
  kable_styling(bootstrap_options = c("bordered"),
                full_width = FALSE, font_size = tbl_fs(13)) |>
  row_spec(0, bold = TRUE, color = "white", background = "#2c7bb6")
```

**R = 0.821** (very strong positive correlation); **R² = 0.674** — 67.4% of
the variation in fasting glucose is explained by waist circumference.
*(The Std. Error of the Estimate, 8.458 mg/dL, is the typical size of a
residual; it is not needed in this course.)*

```{r}
#| label: tbl-spss-anova
#| tbl-cap: "SPSS Output: ANOVA Table"

anova_tab <- data.frame(
  Source = c("Regression", "Residual", "Total"),
  SS = c("5621.4", "2718.6", "8340.0"),
  df = c("1", "38", "39"),
  MS = c("5621.4", "71.5", ""),
  F = c("78.62", "", ""),
  Sig = c("< 0.001", "", "")
)

kable(anova_tab,
      col.names = c("", "Sum of Squares", "df", "Mean Square", "F", "Sig."),
      align = c("l","c","c","c","c","c")) |>
  kable_styling(bootstrap_options = c("bordered"),
                full_width = FALSE, font_size = tbl_fs(13)) |>
  row_spec(0, bold = TRUE, color = "white", background = "#2c7bb6") |>
  row_spec(1, bold = TRUE)
```

The **ANOVA table** splits the total variation in $Y$ (Total Sum of Squares)
into the part explained by the line (Regression) and the part left over
(Residual). $R^2$ = Regression SS / Total SS = 5621.4 / 8340.0 = 0.674.

The **F** statistic tests $H_0: \beta_1 = 0$, the same hypothesis as the
slope test. In simple linear regression, $F = t^2$ for the slope
($8.87^2 \approx 78.7$, the small difference is rounding), so the two tests
always give the same p-value.

**F = 78.62, p < 0.001** — the regression model is statistically significant.

```{r}
#| label: tbl-spss-coef
#| tbl-cap: "SPSS Output: Coefficients Table"

coef_tab <- data.frame(
  ` ` = c("Constant", "Waist circumference"),
  B = c("32.15", "0.87"),
  SE = c("6.24", "0.10"),
  Beta = c("-", "0.821"),
  t = c("5.15", "8.87"),
  Sig = c("< 0.001", "< 0.001")
)

kable(coef_tab,
      col.names = c("", "B", "Std. Error", "Beta (standardized)", "t", "Sig."),
      align = c("l","c","c","c","c","c")) |>
  kable_styling(bootstrap_options = c("bordered"),
                full_width = FALSE, font_size = tbl_fs(13)) |>
  row_spec(0, bold = TRUE, color = "white", background = "#2c7bb6") |>
  row_spec(2, bold = TRUE)
```

**Regression equation:** $\widehat{\text{Glucose}} = 32.15 + 0.87 \times \text{Waist}$

The slope **B = 0.87** is statistically significant ($p < 0.001$, read from
the *Waist circumference* row): for each 1 cm increase in waist
circumference, fasting glucose increases by **0.87 mg/dL** on average.
The **Beta (standardized)** value equals $r$ in simple regression and is
not needed in this course.

---

## Example 2: Interpreting SPSS Output

A researcher uses SPSS to examine the relationship between
**BMI (kg/m²)** and **fasting blood glucose (mg/dL)** in 60 adults.
The following output is obtained:

**Model Summary:**

| R | R² | Adjusted R² | Std. Error |
|:---:|:---:|:---:|:---:|
| 0.612 | 0.375 | 0.364 | 14.23 |

**Coefficients:**

| | B | Std. Error | t | Sig. |
|---|:---:|:---:|:---:|:---:|
| Constant | 48.32 | 11.45 | 4.22 | < 0.001 |
| BMI | 2.91 | 0.46 | 6.33 | < 0.001 |

::: {.callout-tip icon=false}
## Interpretation

**Correlation:** $r = 0.612$ — a **strong positive** relationship between
BMI and fasting blood glucose (0.60–0.79 in the guideline table).

**Regression equation:**
$$\hat{Y} = 48.32 + 2.91 \times \text{BMI}$$

**Slope ($b_1 = 2.91$):** For each 1 kg/m² increase in BMI,
fasting blood glucose increases by **2.91 mg/dL** on average.

**Intercept ($b_0 = 48.32$):** The predicted fasting glucose when
BMI = 0 is 48.32 mg/dL. *(Note: BMI of 0 is not a meaningful value
in practice — the intercept is simply a mathematical anchor for the line.)*

**$R^2 = 0.375$:** BMI explains **37.5%** of the variation in fasting
blood glucose. The remaining 62.5% is due to other factors (diet, genetics,
physical activity, etc.).

**Significance ($p < 0.001$):** The slope is highly significant —
BMI is a statistically significant predictor of fasting blood glucose.

**Prediction example:** A person with BMI = 28 kg/m²:
$$\hat{Y} = 48.32 + 2.91 \times 28 = 48.32 + 81.48 = 129.8 \text{ mg/dL}$$
:::

---

# Key Takeaways

::: {.callout-note icon=false}
## Chapter 6 Summary

1. **Statistical modeling** describes the mathematical relationship between
   an outcome variable $Y$ and one or more predictor variables $X$.

2. **Pearson's $r$** measures the strength and direction of a linear
   relationship between two continuous variables ($-1 \leq r \leq +1$).
   Its significance is tested with $H_0: \rho = 0$ (SPSS: Sig. (2-tailed)
   in the Correlations table).

3. **Simple linear regression** fits the equation $\hat{Y} = b_0 + b_1 X$
   using the least squares method, which minimizes the sum of squared residuals.

4. The **slope $b_1$** tells us how much $Y$ changes for each one-unit
   increase in $X$; the **intercept $b_0$** is the predicted value of $Y$
   when $X = 0$.

5. **$R^2$** is the proportion of variation in $Y$ explained by the model.
   A higher $R^2$ means a better fit.

6. Always test whether the slope is significantly different from zero
   ($H_0: \beta_1 = 0$) before using the model for interpretation. In simple
   regression the slope t-test, the ANOVA F-test ($F = t^2$) and the test of
   $r$ all give the same p-value.

7. The model is only valid when the four assumptions — **linearity,
   independence, homoscedasticity, and normality of residuals** — are
   reasonably satisfied.
:::

---

## Formula Summary

| Quantity | Formula | Notes |
|:----------------|:----------------------------------------|:----------------------|
| Pearson correlation | $r = \dfrac{\sum (x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum (x_i - \bar{x})^2 \sum (y_i - \bar{y})^2}}$ | $-1 \leq r \leq 1$ |
| Test of $r$ | $t = \dfrac{r\sqrt{n-2}}{\sqrt{1 - r^2}}$ | $H_0: \rho = 0$; $df = n - 2$ |
| Slope | $b_1 = \dfrac{\sum (x_i - \bar{x})(y_i - \bar{y})}{\sum (x_i - \bar{x})^2} = r \cdot \dfrac{s_y}{s_x}$ | Change in $\hat{Y}$ per 1-unit increase in $X$ |
| Intercept | $b_0 = \bar{y} - b_1 \bar{x}$ | Predicted $Y$ when $X = 0$ |
| Prediction | $\hat{Y} = b_0 + b_1 X$ | Only within the range of the data |
| Residual | $e_i = Y_i - \hat{Y}_i$ | OLS minimises $\sum e_i^2$ |
| Coefficient of determination | $R^2 = r^2 = \dfrac{\text{Regression SS}}{\text{Total SS}}$ | Proportion of variation in $Y$ explained |
| Test of the slope | $t = \dfrac{b_1}{SE_{b_1}}$ | $H_0: \beta_1 = 0$; $df = n - 2$; $F = t^2$ |

: Formulas in this chapter {.striped}

---

# Exercises

::: {.callout-caution icon=false}
## Exercise 1

A study records the **hours of sleep per night** ($X$) and
**systolic blood pressure (mmHg)** ($Y$) for 6 adults:

| Subject | Sleep (hrs) | SBP (mmHg) |
|:---:|:---:|:---:|
| 1 | 5 | 145 |
| 2 | 6 | 138 |
| 3 | 7 | 130 |
| 4 | 7 | 125 |
| 5 | 8 | 120 |
| 6 | 9 | 115 |

**(a)** Draw a scatter plot and describe the relationship.

**(b)** Calculate Pearson's $r$. What does it indicate?

**(c)** Find the regression equation $\hat{Y} = b_0 + b_1 X$.

**(d)** Predict the SBP of a person who sleeps 6.5 hours per night.

**(e)** Calculate $R^2$ and interpret it.
:::

::: {.callout-caution icon=false}
## Exercise 2

A researcher uses SPSS to study the relationship between a patient's
**age (years)** and **total cholesterol (mg/dL)**.
The SPSS output gives the following results:

**Model Summary:** R = 0.548, R² = 0.300

**Coefficients:**

| | B | Std. Error | t | Sig. |
|---|:---:|:---:|:---:|:---:|
| Constant | 120.4 | 18.6 | 6.47 | < 0.001 |
| Age | 1.85 | 0.31 | 5.97 | < 0.001 |

**(a)** Write the regression equation.

**(b)** Interpret the slope in the context of this study.

**(c)** What proportion of the variation in cholesterol is explained
by age? Is this a strong model?

**(d)** Predict the total cholesterol of a 55-year-old patient.

**(e)** Is the predictor (age) statistically significant at $\alpha = 0.05$?
How do you know?
:::

::: {.callout-caution icon=false}
## Exercise 3 — Reflection Questions

Answer the following in 2–3 sentences each:

**(a)** A researcher finds that the correlation between ice cream sales
and drowning deaths is $r = 0.85$. Does this mean eating ice cream
causes drowning? Explain using the concept of correlation vs. causation.

**(b)** In a regression model predicting patient recovery time from
age ($R^2 = 0.22$), a colleague says *"this model is useless because
it only explains 22% of the variation."* Do you agree? What additional
information would help evaluate this claim?

**(c)** Why is it generally not appropriate to use a regression equation
to predict $Y$ for values of $X$ that are far outside the range of the
original data?
:::

---

*End of Chapter 6*

---

::: {.callout-note icon=false}
## End of Course

Congratulations on completing all six chapters of Introduction to Biostatistics!

You have covered the full journey from understanding what data is,
to summarizing it, to understanding distributions, making estimates,
testing hypotheses, and finally modeling relationships between variables.

These tools form the foundation of evidence-based practice in health science.
:::
