# Introduction to Statistical Modeling

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(knitr)
library(kableExtra)
library(patchwork)
```

---

# Introduction {.unnumbered}

::: {.callout-note icon=false}
## Why Does This Matter?

So far we have described data, estimated population parameters, and tested
whether two groups differ. But many real-world health questions go beyond
simple comparisons:

- Does body weight *predict* blood pressure?
- As age *increases*, how does cholesterol *change*?
- Can we use a patient's BMI to *estimate* their fasting glucose level?

These questions require us to model the **relationship between two variables**.
**Statistical modeling** is the process of building a mathematical equation
that describes how one variable is related to another — and using that
equation to understand, explain, and predict outcomes.

In this chapter we focus on the simplest and most widely used model:
**simple linear regression**.
:::

---

# What Is a Statistical Model?

A **statistical model** is a mathematical equation that describes the
relationship between:

- An **outcome variable** (also called the *dependent variable*, *response
  variable*, or $Y$) — the variable we want to explain or predict.
- One or more **predictor variables** (also called *independent variables*,
  *explanatory variables*, or $X$) — the variables we use to explain or
  predict $Y$.

In this chapter we study **simple linear regression** — one outcome $Y$
and one predictor $X$, with a straight-line relationship.

---

# Correlation: Measuring the Relationship

Before building a model, we first ask: *Is there a relationship between
$X$ and $Y$, and how strong is it?*

## The Scatter Plot

A **scatter plot** displays pairs of values $(x_i, y_i)$ as points on a
graph. The pattern of points reveals the **direction** and **strength**
of the relationship.

```{r}
#| label: fig-scatter-types
#| fig-cap: "Four possible relationships between two variables"
#| fig-align: center
#| fig-height: 4
#| fig-width: 10

set.seed(42)
n <- 80

d1 <- data.frame(x = rnorm(n), y = 0.8*rnorm(n, mean=0) + rnorm(n,sd=0.5),
                 type = "Strong positive")
d1$y <- d1$x * 0.9 + rnorm(n, sd=0.3)

d2 <- data.frame(x = rnorm(n), y = -0.9*rnorm(n) + rnorm(n,sd=0.3),
                 type = "Strong negative")
d2$y <- -d2$x * 0.9 + rnorm(n, sd=0.3)

d3 <- data.frame(x = rnorm(n), y = 0.4*rnorm(n) + rnorm(n,sd=0.9),
                 type = "Weak positive")
d3$y <- d3$x * 0.3 + rnorm(n, sd=0.9)

d4 <- data.frame(x = rnorm(n), y = rnorm(n),
                 type = "No relationship")

all_d <- rbind(d1, d2, d3, d4)
all_d$type <- factor(all_d$type,
                     levels = c("Strong positive","Strong negative",
                                "Weak positive","No relationship"))

ggplot(all_d, aes(x = x, y = y)) +
  geom_point(alpha = 0.5, color = "#2c7bb6", size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#d7191c",
              linewidth = 1, formula = y ~ x) +
  facet_wrap(~ type, nrow = 1, scales = "free") +
  theme_minimal(base_size = 12) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        strip.text = element_text(face = "bold", size = 11))
```

---

## Pearson's Correlation Coefficient ($r$)

The **Pearson correlation coefficient** $r$ quantifies the **strength**
and **direction** of the *linear* relationship between two continuous variables.

$$r = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}
           {\sqrt{\sum_{i=1}^{n}(x_i-\bar{x})^2 \cdot \sum_{i=1}^{n}(y_i-\bar{y})^2}}$$

### Properties of $r$

- $r$ always lies between **−1 and +1**
- $r = +1$: perfect positive linear relationship
- $r = -1$: perfect negative linear relationship
- $r = 0$: no linear relationship

### Interpreting the Magnitude of $r$

```{r}
#| label: tbl-r-interpret
#| tbl-cap: "General guidelines for interpreting the correlation coefficient"

r_tab <- data.frame(
  `|r| range` = c("0.00 - 0.19", "0.20 - 0.39",
                   "0.40 - 0.59", "0.60 - 0.79", "0.80 - 1.00"),
  Strength = c("Negligible", "Weak", "Moderate", "Strong", "Very strong"),
  Direction = c("-", "Positive or negative depending on sign",
                "Positive or negative depending on sign",
                "Positive or negative depending on sign",
                "Positive or negative depending on sign")
)

kable(r_tab, col.names = c("|r| value", "Strength", "Note"),
      align = c("c","l","l")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = 13) |>
  column_spec(1, bold = TRUE, color = "#2c7bb6")
```

::: {.callout-warning icon=false}
## Correlation Is Not Causation

A high correlation between $X$ and $Y$ does **not** mean that $X$
*causes* $Y$. Both variables may be influenced by a third variable,
or the relationship may be coincidental. Always interpret correlations
in the context of biological or clinical plausibility.
:::

---

# Simple Linear Regression

## The Regression Equation

Simple linear regression models the relationship between $X$ and $Y$
as a straight line:

$$\hat{Y} = b_0 + b_1 X$$

Where:

| Symbol | Name | Meaning |
|:---:|---|---|
| $\hat{Y}$ | Predicted value | The estimated value of $Y$ for a given $X$ |
| $b_0$ | Intercept | The predicted value of $Y$ when $X = 0$ |
| $b_1$ | Slope | The change in $\hat{Y}$ for each one-unit increase in $X$ |

The full model including random error is:

$$Y = b_0 + b_1 X + \varepsilon$$

Where $\varepsilon$ represents the **residual** — the difference between
the observed value and the predicted value:

$$\varepsilon_i = Y_i - \hat{Y}_i$$

---

## Estimating the Coefficients

The values of $b_0$ and $b_1$ are chosen to minimize the **sum of squared
residuals** — this method is called **Ordinary Least Squares (OLS)**:

$$\text{Minimize: } \sum_{i=1}^{n}(Y_i - \hat{Y}_i)^2 = \sum_{i=1}^{n}\varepsilon_i^2$$

The OLS formulas are:

$$b_1 = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}
             {\sum_{i=1}^{n}(x_i - \bar{x})^2}
       = r \cdot \frac{s_y}{s_x}$$

$$b_0 = \bar{y} - b_1 \bar{x}$$

```{r}
#| label: fig-ols
#| fig-cap: "Least squares regression line minimizes the sum of squared residuals (vertical red lines)"
#| fig-align: center
#| fig-height: 4
#| fig-width: 6

set.seed(10)
x <- seq(20, 80, length = 20)
y <- 60 + 0.9 * x + rnorm(20, sd = 10)
fit <- lm(y ~ x)
df_ols <- data.frame(x = x, y = y,
                     yhat = fitted(fit),
                     resid = residuals(fit))

ggplot(df_ols, aes(x = x, y = y)) +
  geom_segment(aes(xend = x, yend = yhat),
               color = "#d7191c", linewidth = 0.7, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, color = "#2c7bb6",
              linewidth = 1.3, formula = y ~ x) +
  geom_point(size = 3, color = "#2c7bb6", alpha = 0.8) +
  labs(x = "X (Predictor)", y = "Y (Outcome)",
       caption = "Red dashed lines = residuals (ε = Y − Ŷ)") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

---

## The Coefficient of Determination ($R^2$)

**$R^2$** (R-squared) measures how well the regression line fits the data —
specifically, the **proportion of the total variation in $Y$ that is
explained by $X$**:

$$R^2 = r^2 \quad \text{(in simple linear regression)}$$

$$R^2 = \frac{\text{Variation explained by the model}}{\text{Total variation in } Y}$$

- $R^2 = 1$: the model explains 100% of the variation in $Y$ (perfect fit)
- $R^2 = 0$: the model explains none of the variation in $Y$
- $R^2 = 0.65$ means: *"65% of the variation in $Y$ is explained by $X$"*

---

## Testing the Significance of the Regression

Before using the regression equation for interpretation or prediction,
we should test whether the slope $b_1$ is significantly different from zero.
If $b_1 = 0$, then $X$ provides no information about $Y$.

**Hypotheses:**

$$H_0: \beta_1 = 0 \quad \text{(X is not a useful predictor of Y)}$$
$$H_1: \beta_1 \neq 0 \quad \text{(X is a useful predictor of Y)}$$

The test statistic is:

$$t = \frac{b_1}{SE_{b_1}}$$

which follows a t-distribution with $df = n - 2$.

SPSS reports this p-value automatically in the **Coefficients table**
of the regression output.

---

## Assumptions of Linear Regression

The regression model is valid only when the following assumptions are met:

```{r}
#| label: tbl-assumptions
#| tbl-cap: "Assumptions of Simple Linear Regression"

assump <- data.frame(
  Assumption = c("Linearity",
                 "Independence",
                 "Homoscedasticity",
                 "Normality of residuals"),
  Meaning = c(
    "The relationship between X and Y is linear (a straight line is appropriate)",
    "Observations are independent of each other",
    "The spread of residuals is roughly constant across all values of X",
    "Residuals are approximately normally distributed"
  ),
  `How to Check` = c(
    "Scatter plot of X vs. Y; residual plot",
    "Study design - satisfied by random sampling",
    "Plot of residuals vs. fitted values - no funnel shape",
    "Histogram or Q-Q plot of residuals"
  )
)

kable(assump, col.names = c("Assumption", "Meaning", "How to Check"),
      align = c("l","l","l")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = 13) |>
  column_spec(1, bold = TRUE, color = "#2c7bb6")
```

---

## Reading SPSS Regression Output

SPSS produces three key tables for simple linear regression.
Here is how to read them:

```{r}
#| label: tbl-spss-model
#| tbl-cap: "SPSS Output: Model Summary"

model_sum <- data.frame(
  R = "0.821",
  R2 = "0.674",
  `Adjusted R2` = "0.666",
  `Std. Error of the Estimate` = "8.432"
)

kable(model_sum,
      col.names = c("R", "R2", "Adjusted R2", "Std. Error of the Estimate"),
      align = rep("c",4)) |>
  kable_styling(bootstrap_options = c("bordered"),
                full_width = FALSE, font_size = 13) |>
  row_spec(0, bold = TRUE, color = "white", background = "#2c7bb6")
```

 **R = 0.821** (strong positive correlation); **R² = 0.674** (67.4% of
variation in Y explained by X).

```{r}
#| label: tbl-spss-anova
#| tbl-cap: "SPSS Output: ANOVA Table"

anova_tab <- data.frame(
  Source = c("Regression", "Residual", "Total"),
  SS = c("5621.4", "2718.6", "8340.0"),
  df = c("1", "38", "39"),
  MS = c("5621.4", "71.5", ""),
  F = c("78.62", "", ""),
  Sig = c("< 0.001", "", "")
)

kable(anova_tab,
      col.names = c("", "Sum of Squares", "df", "Mean Square", "F", "Sig."),
      align = c("l","c","c","c","c","c")) |>
  kable_styling(bootstrap_options = c("bordered"),
                full_width = FALSE, font_size = 13) |>
  row_spec(0, bold = TRUE, color = "white", background = "#2c7bb6") |>
  row_spec(1, bold = TRUE)
```

 **F = 78.62, p < 0.001** — the overall regression model is
statistically significant.

```{r}
#| label: tbl-spss-coef
#| tbl-cap: "SPSS Output: Coefficients Table"

coef_tab <- data.frame(
  ` ` = c("(Constant)", "X (predictor)"),
  B = c("32.15", "0.87"),
  SE = c("6.24", "0.10"),
  Beta = c("-", "0.821"),
  t = c("5.15", "8.87"),
  Sig = c("< 0.001", "< 0.001")
)

kable(coef_tab,
      col.names = c("", "B", "Std. Error", "Beta (standardized)", "t", "Sig."),
      align = c("l","c","c","c","c","c")) |>
  kable_styling(bootstrap_options = c("bordered"),
                full_width = FALSE, font_size = 13) |>
  row_spec(0, bold = TRUE, color = "white", background = "#2c7bb6") |>
  row_spec(2, bold = TRUE, background = "#eaf4fb")
```

 **Regression equation:** $\hat{Y} = 32.15 + 0.87X$

 The slope **B = 0.87** is statistically significant ($p < 0.001$):
for each one-unit increase in $X$, $Y$ increases by **0.87 units** on average.

---

# Key Takeaways

::: {.callout-note icon=false}
## Chapter 6 Summary

1. **Statistical modeling** describes the mathematical relationship between
   an outcome variable $Y$ and one or more predictor variables $X$.

2. **Pearson's $r$** measures the strength and direction of a linear
   relationship between two continuous variables ($-1 \leq r \leq +1$).

3. **Simple linear regression** fits the equation $\hat{Y} = b_0 + b_1 X$
   using the least squares method, which minimizes the sum of squared residuals.

4. The **slope $b_1$** tells us how much $Y$ changes for each one-unit
   increase in $X$; the **intercept $b_0$** is the predicted value of $Y$
   when $X = 0$.

5. **$R^2$** is the proportion of variation in $Y$ explained by the model.
   A higher $R^2$ means a better fit.

6. Always test whether $b_1$ is significantly different from zero ($H_0: \beta_1 = 0$)
   before using the model for interpretation.

7. The model is only valid when the four assumptions — **linearity,
   independence, homoscedasticity, and normality of residuals** — are
   reasonably satisfied.
:::

---

# Examples

## Example 1: Calculating Correlation and Regression

A researcher records the **age (years)** and **systolic blood pressure (mmHg)**
of 8 adults:

```{r}
#| label: tbl-ex1-data
#| tbl-cap: "Age and Systolic Blood Pressure of 8 Adults"

age <- c(25, 32, 40, 48, 55, 62, 70, 78)
sbp <- c(115, 120, 128, 135, 142, 148, 158, 165)

ex1 <- data.frame(Subject = paste0("S", 1:8), Age = age, SBP = sbp)

kable(ex1, col.names = c("Subject", "Age (years)", "SBP (mmHg)"),
      align = c("c","c","c")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = 13)
```

**(a)** Calculate Pearson's $r$.

**(b)** Find the regression equation $\hat{Y} = b_0 + b_1 X$.

**(c)** Predict the SBP of a 50-year-old adult.

**(d)** Interpret $R^2$.

::: {.callout-tip icon=false}
## Solution

```{r}
#| label: tbl-ex1-calc
#| tbl-cap: "Calculation Table for Correlation and Regression"

xbar <- mean(age)
ybar <- mean(sbp)

calc <- data.frame(
  Subject = paste0("S", 1:8),
  x = age,
  y = sbp,
  xdev = round(age - xbar, 2),
  ydev = round(sbp - ybar, 2),
  xdev2 = round((age - xbar)^2, 2),
  ydev2 = round((sbp - ybar)^2, 2),
  xydev = round((age - xbar)*(sbp - ybar), 2)
)

totrow <- data.frame(
  Subject = "**Total**",
  x = NA, y = NA,
  xdev = NA, ydev = NA,
  xdev2 = round(sum((age-xbar)^2), 2),
  ydev2 = round(sum((sbp-ybar)^2), 2),
  xydev = round(sum((age-xbar)*(sbp-ybar)), 2)
)

calc_all <- rbind(calc, totrow)

kable(calc_all,
      col.names = c("Subject","x","y","x-xbar","y-ybar",
                    "(x-xbar)2","(y-ybar)2","(x-xbar)(y-ybar)"),
      align = rep("c", 8), na_print = "-") |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = 12) |>
  row_spec(9, bold = TRUE, background = "#eaf4fb") |>
  row_spec(0, bold = TRUE, color = "white", background = "#2c7bb6")
```

**(a) Pearson's r:**

$$r = \frac{\sum(x_i-\bar{x})(y_i-\bar{y})}
           {\sqrt{\sum(x_i-\bar{x})^2 \cdot \sum(y_i-\bar{y})^2}}
  = \frac{`r round(sum((age-xbar)*(sbp-ybar)),2)`}
         {\sqrt{`r round(sum((age-xbar)^2),2)` \times `r round(sum((sbp-ybar)^2),2)`}}
  = \frac{`r round(sum((age-xbar)*(sbp-ybar)),2)`}
         {`r round(sqrt(sum((age-xbar)^2)*sum((sbp-ybar)^2)),2)`}
  = \mathbf{`r round(cor(age,sbp),3)`}$$

This indicates a **very strong positive** linear relationship between
age and systolic blood pressure.

**(b) Regression coefficients:**

$$b_1 = \frac{\sum(x_i-\bar{x})(y_i-\bar{y})}{\sum(x_i-\bar{x})^2}
      = \frac{`r round(sum((age-xbar)*(sbp-ybar)),2)`}
             {`r round(sum((age-xbar)^2),2)`}
      = `r round(coef(lm(sbp~age))[2],3)`$$

$$b_0 = \bar{y} - b_1\bar{x}
      = `r round(ybar,2)` - `r round(coef(lm(sbp~age))[2],3)` \times `r round(xbar,2)`
      = `r round(coef(lm(sbp~age))[1],2)`$$

$$\boxed{\hat{Y} = `r round(coef(lm(sbp~age))[1],2)` + `r round(coef(lm(sbp~age))[2],3)` X}$$

**(c) Prediction for age = 50:**

$$\hat{Y} = `r round(coef(lm(sbp~age))[1],2)` + `r round(coef(lm(sbp~age))[2],3)` \times 50
          = `r round(predict(lm(sbp~age), newdata=data.frame(age=50)),1)` \text{ mmHg}$$

**(d) $R^2$:**

$$R^2 = r^2 = (`r round(cor(age,sbp),3)`)^2 = `r round(cor(age,sbp)^2,3)`$$

**`r round(cor(age,sbp)^2*100,1)`% of the variation in systolic blood pressure
is explained by age.** The remaining `r round((1-cor(age,sbp)^2)*100,1)`%
is due to other factors not included in this model.
:::

```{r}
#| label: fig-ex1-fit
#| fig-cap: "Scatter plot with fitted regression line: Age vs. Systolic Blood Pressure"
#| fig-align: center
#| fig-height: 4
#| fig-width: 6

fit1 <- lm(sbp ~ age)
df1  <- data.frame(age = age, sbp = sbp)

ggplot(df1, aes(x = age, y = sbp)) +
  geom_point(size = 3.5, color = "#2c7bb6") +
  geom_smooth(method = "lm", se = TRUE, color = "#d7191c",
              fill = "#fdae61", alpha = 0.2, formula = y ~ x) +
  annotate("text", x = 30, y = 162,
           label = paste0("Ŷ = ", round(coef(fit1)[1],2),
                          " + ", round(coef(fit1)[2],3), "X\n",
                          "r = ", round(cor(age,sbp),3),
                          ",  R² = ", round(cor(age,sbp)^2,3)),
           hjust = 0, size = 4, color = "#d7191c") +
  labs(x = "Age (years)", y = "Systolic Blood Pressure (mmHg)") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

---

## Example 2: Interpreting SPSS Output

A researcher uses SPSS to examine the relationship between
**BMI (kg/m²)** and **fasting blood glucose (mg/dL)** in 60 adults.
The following output is obtained:

**Model Summary:**

| R | R² | Adjusted R² | Std. Error |
|:---:|:---:|:---:|:---:|
| 0.612 | 0.375 | 0.364 | 14.23 |

**Coefficients:**

| | B | Std. Error | t | Sig. |
|---|:---:|:---:|:---:|:---:|
| (Constant) | 48.32 | 11.45 | 4.22 | < 0.001 |
| BMI | 2.91 | 0.46 | 6.33 | < 0.001 |

::: {.callout-tip icon=false}
## Interpretation

**Correlation:** $r = 0.612$ — a **moderate-to-strong positive**
relationship between BMI and fasting blood glucose.

**Regression equation:**
$$\hat{Y} = 48.32 + 2.91 \times \text{BMI}$$

**Slope ($b_1 = 2.91$):** For each 1 kg/m² increase in BMI,
fasting blood glucose increases by **2.91 mg/dL** on average,
holding other factors constant.

**Intercept ($b_0 = 48.32$):** The predicted fasting glucose when
BMI = 0 is 48.32 mg/dL. *(Note: BMI of 0 is not a meaningful value
in practice — the intercept is simply a mathematical anchor for the line.)*

**$R^2 = 0.375$:** BMI explains **37.5%** of the variation in fasting
blood glucose. The remaining 62.5% is due to other factors (diet, genetics,
physical activity, etc.).

**Significance ($p < 0.001$):** The slope is highly significant —
BMI is a statistically significant predictor of fasting blood glucose.

**Prediction example:** A person with BMI = 28 kg/m²:
$$\hat{Y} = 48.32 + 2.91 \times 28 = 48.32 + 81.48 = 129.8 \text{ mg/dL}$$
:::

---

# Exercises

::: {.callout-caution icon=false}
## Exercise 1

A study records the **hours of sleep per night** ($X$) and
**systolic blood pressure (mmHg)** ($Y$) for 6 adults:

| Subject | Sleep (hrs) | SBP (mmHg) |
|:---:|:---:|:---:|
| 1 | 5 | 145 |
| 2 | 6 | 138 |
| 3 | 7 | 130 |
| 4 | 7 | 125 |
| 5 | 8 | 120 |
| 6 | 9 | 115 |

**(a)** Draw a scatter plot and describe the relationship.

**(b)** Calculate Pearson's $r$. What does it indicate?

**(c)** Find the regression equation $\hat{Y} = b_0 + b_1 X$.

**(d)** Predict the SBP of a person who sleeps 6.5 hours per night.

**(e)** Calculate $R^2$ and interpret it.
:::

::: {.callout-caution icon=false}
## Exercise 2

A researcher uses SPSS to study the relationship between a patient's
**age (years)** and **total cholesterol (mg/dL)**.
The SPSS output gives the following results:

**Model Summary:** R = 0.548, R² = 0.300

**Coefficients:**

| | B | Std. Error | t | Sig. |
|---|:---:|:---:|:---:|:---:|
| (Constant) | 120.4 | 18.6 | 6.47 | < 0.001 |
| Age | 1.85 | 0.31 | 5.97 | < 0.001 |

**(a)** Write the regression equation.

**(b)** Interpret the slope in the context of this study.

**(c)** What proportion of the variation in cholesterol is explained
by age? Is this a strong model?

**(d)** Predict the total cholesterol of a 55-year-old patient.

**(e)** Is the predictor (age) statistically significant at $\alpha = 0.05$?
How do you know?
:::

::: {.callout-caution icon=false}
## Exercise 3 — Reflection Questions

Answer the following in 2–3 sentences each:

**(a)** A researcher finds that the correlation between ice cream sales
and drowning deaths is $r = 0.85$. Does this mean eating ice cream
causes drowning? Explain using the concept of correlation vs. causation.

**(b)** In a regression model predicting patient recovery time from
age ($R^2 = 0.22$), a colleague says *"this model is useless because
it only explains 22% of the variation."* Do you agree? What additional
information would help evaluate this claim?

**(c)** Why is it generally not appropriate to use a regression equation
to predict $Y$ for values of $X$ that are far outside the range of the
original data?
:::

---

*End of Chapter 6*

---

::: {.callout-note icon=false}
## End of Course

Congratulations on completing all six chapters of Introduction to Biostatistics!

You have covered the full journey from understanding what data is,
to summarizing it, to understanding distributions, making estimates,
testing hypotheses, and finally modeling relationships between variables.

These tools form the foundation of evidence-based practice in health science.
:::
