# Random Variable and Probability Distribution

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(knitr)
library(kableExtra)
library(dplyr)
# Table helpers (used by the book and the slides)
# tbl_fs(): font size; PDF tables use 10pt so they fit the page width
# pdf_w():  fixed column width in the PDF only, so long text wraps
tbl_fs <- function(html_size = 13) if (knitr::is_latex_output()) 10 else html_size
pdf_w  <- function(x) if (knitr::is_latex_output()) x else NULL
```

---

# Introduction {.unnumbered}

::: {.callout-note icon=false}
## Why Does This Matter?

When we collect data — patient ages, blood pressure readings, test results —
we rarely get the same value twice. Outcomes vary, and that variation
follows patterns we can describe mathematically.

Understanding **how data is distributed** allows us to answer questions like:

- What is the probability that a randomly selected patient has a fever above 39°C?
- What proportion of adults have a BMI in the healthy range?
- How unusual is a blood pressure reading of 160 mmHg in this population?

In this chapter, we start from **frequency distributions** (what we
observed in a sample), move to **probability** and **random variables**
(what we expect in the population), and finish with the
**normal distribution** — the most important probability distribution
in all of biostatistics.
:::

---

# Frequency Distributions

## What Is a Frequency Distribution?

A **frequency distribution** is a systematic way of organizing raw data
into a table that shows how often each value (or range of values) occurs.

It answers the simple question: *"How many times does each value appear?"*

---

## Frequency Distribution for Categorical Data

For categorical variables, we simply count how many observations fall
into each category.

::: {.callout-tip icon=false}
## Example
Blood types of 200 patients admitted to a hospital last month:

```{r}
#| label: tbl-cat-freq
#| tbl-cap: "Frequency Distribution: Blood Type of 200 Patients"

blood <- data.frame(
  `Blood Type` = c("A", "B", "AB", "O", "Total"),
  Frequency = c(76, 54, 30, 40, 200),
  `Relative Frequency` = c("76/200 = 0.38", "54/200 = 0.27",
                             "30/200 = 0.15", "40/200 = 0.20",
                             "200/200 = 1.00"),
  Percentage = c("38%", "27%", "15%", "20%", "100%")
)

kable(blood,
      col.names = c("Blood Type", "Frequency (f)",
                    "Relative Frequency (f/n)", "Percentage"),
      align = c("c","c","c","c")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = tbl_fs(13)) |>
  row_spec(5, bold = TRUE, background = "#f0f0f0") |>
  column_spec(1, bold = TRUE)
```
:::

**Relative frequency** = $\dfrac{f}{n}$, where $f$ is the count for that
category and $n$ is the total number of observations.

---

## Frequency Distribution for Continuous Data

For continuous variables, we group values into **class intervals**
(also called *bins*) and count how many observations fall in each interval.

**Steps to construct a frequency distribution table:**

1. Find the **range** = Maximum − Minimum
2. Choose the **number of classes** (typically 5–10)
3. Calculate **class width** = $\dfrac{\text{Range}}{\text{Number of classes}}$ (round up)
4. List the class intervals and count observations in each

::: {.callout-tip icon=false}
## Example
Systolic blood pressure (mmHg) of 80 patients:

```{r}
#| label: tbl-cont-freq
#| tbl-cap: "Frequency Distribution: Systolic Blood Pressure of 80 Patients"

set.seed(123)
sbp <- round(rnorm(80, mean = 130, sd = 15))

breaks <- seq(95, 175, by = 10)
labels <- paste0(breaks[-length(breaks)], "-", breaks[-1] - 1)
sbp_cut <- cut(sbp, breaks = breaks, right = FALSE, labels = labels)
freq_table <- as.data.frame(table(sbp_cut))
colnames(freq_table) <- c("Class Interval (mmHg)", "Frequency (f)")
freq_table$`Relative Frequency` <- round(freq_table$`Frequency (f)` / 80, 3)
freq_table$`Cumulative Frequency` <- cumsum(freq_table$`Frequency (f)`)

kable(freq_table, align = c("c","c","c","c")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = tbl_fs(13)) |>
  column_spec(1, bold = TRUE)
```
:::

**Cumulative frequency** tells us how many observations fall *at or below*
the upper boundary of each class — useful for answering questions like
*"What proportion of patients have SBP below 135 mmHg?"*
(read the cumulative frequency of the class that ends at 134, then divide by 80).

---

## Example 1: Constructing a Frequency Distribution Table

The following are the ages (in years) of 20 patients:

$$45, 62, 38, 55, 71, 48, 60, 33, 52, 67,$$
$$41, 58, 74, 50, 65, 44, 57, 69, 36, 53$$

Construct a frequency distribution table using **5 class intervals**.

::: {.callout-tip icon=false}
## Solution

**Step 1 — Find the range:**
$$\text{Range} = 74 - 33 = 41$$

**Step 2 — Calculate class width:**
$$\text{Class width} = \frac{41}{5} = 8.2 \rightarrow \text{round up to } 9$$

**Step 3 — Construct the table** (starting from the minimum, 33).
Each class is 9 years wide, so the classes are 33–41, 42–50, 51–59,
60–68 and 69–77. Every value belongs to **exactly one** class — for
example, age 60 is counted in 60–68, not in 51–59.

```{r}
#| label: tbl-ex1
#| tbl-cap: "Frequency Distribution of Patient Ages (n = 20)"

ages <- c(45,62,38,55,71,48,60,33,52,67,41,58,74,50,65,44,57,69,36,53)
breaks2 <- c(33, 42, 51, 60, 69, 78)
labels2 <- c("33-41","42-50","51-59","60-68","69-77")
# right = FALSE: each class includes its lower limit and excludes the next
# class's lower limit, so age 60 falls in 60-68 (not 51-59)
age_cut <- cut(ages, breaks = breaks2, right = FALSE, labels = labels2)
ft <- as.data.frame(table(age_cut))
colnames(ft) <- c("Age Group (years)", "Frequency (f)")
ft$`Relative Frequency` <- round(ft$`Frequency (f)` / 20, 2)
ft$Percentage <- paste0(ft$`Relative Frequency` * 100, "%")
ft$`Cumulative Frequency` <- cumsum(ft$`Frequency (f)`)

kable(ft, align = c("c","c","c","c","c")) |>
  kable_styling(bootstrap_options = c("striped","hover","bordered"),
                full_width = FALSE, font_size = tbl_fs(13)) |>
  column_spec(1, bold = TRUE)
```

**Interpretation:** The largest group of patients is aged **51–59 years** (5 patients, 25%).
From the cumulative frequency column, 8 of the 20 patients (40%) are aged
50 years or younger, and 13 patients (65%) are aged 59 years or younger.
:::

---

## Visualizing Frequency Distributions

A frequency distribution table can be visualized as a **histogram**,
where the height of each bar represents the frequency of that class interval.

```{r}
#| label: fig-hist-freq
#| fig-cap: "Histogram of Systolic Blood Pressure (80 patients)"
#| fig-align: center
#| fig-height: 4
#| fig-width: 6

set.seed(123)
sbp_data <- data.frame(sbp = round(rnorm(80, mean = 130, sd = 15)))

ggplot(sbp_data, aes(x = sbp)) +
  geom_histogram(breaks = seq(95, 175, by = 10),
                 fill = "#2c7bb6", color = "white", alpha = 0.85) +
  labs(x = "Systolic Blood Pressure (mmHg)", y = "Frequency") +
  scale_x_continuous(breaks = seq(95, 175, by = 10)) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

The **shape** of the histogram is important — it tells us about the
**distribution** of the data, which leads us to the next section.

---

# From Frequency to Probability

## Probability

A **probability** is a number between 0 and 1 that describes how likely
an event is:

- $P = 0$ — the event never happens
- $P = 1$ — the event always happens
- $P = 0.25$ — the event happens about 25% of the time in the long run

The **relative frequency** we calculated above is our link to probability.
In the blood-type table, 38% of the 200 patients had blood type A.
If we pick one patient **at random** from this group, the probability that
the patient has blood type A is $P(\text{type A}) = 0.38$.

When we only have a sample, the relative frequency is our best
*estimate* of the probability in the population.

## Random Variable

A **random variable** is a variable whose value depends on **chance** —
we do not know its value until we select an individual and measure it.

> **Example:** "The systolic blood pressure of a randomly selected patient"
> is a random variable. Before we choose the patient, we cannot say what
> the value will be, but we *can* say which values are more likely.

Random variables follow the same classification as data in Chapter 1:

| Random variable | Example | Probabilities are shown as |
|---|---|---|
| **Discrete** | Number of falls on a ward per month | A bar for each possible count |
| **Continuous** | Systolic blood pressure | The **area under a smooth curve** |

A **probability distribution** lists all possible values of a random
variable and how likely each one is.

For a continuous random variable, imagine collecting more and more
patients and making the class intervals of the histogram narrower and
narrower: the bars smooth out into a **curve**. The probability that a
value falls in a range equals the **area under the curve** over that
range, and the **total area is 1** (100%).

The most important of these curves is the **normal distribution**.

---

# The Normal Distribution

## What Is the Normal Distribution?

The **normal distribution** (also called the *Gaussian distribution* or
*bell curve*) is the most important probability distribution in statistics.

Many biological and clinical measurements naturally follow — or approximately
follow — a normal distribution:

- Adult height
- Blood pressure in a healthy population
- Hemoglobin level in healthy adults
- Birth weight of full-term newborns

Not every clinical measurement is normal. Length of hospital stay,
for example, is strongly **right-skewed** (Chapter 2), so it does not
follow a normal distribution.

```{r}
#| label: fig-normal-intro
#| fig-cap: "The Normal Distribution: a symmetric, bell-shaped curve"
#| fig-align: center
#| fig-height: 4
#| fig-width: 7

x <- seq(-4, 4, length = 500)
y <- dnorm(x)
df <- data.frame(x = x, y = y)

ggplot(df, aes(x, y)) +
  geom_line(color = "#2c7bb6", linewidth = 1.4) +
  geom_area(fill = "#abd9e9", alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#d7191c", linewidth = 1) +
  annotate("text", x = 0, y = 0.43, label = "Mean = Median = Mode",
           color = "#d7191c", size = 4, hjust = 0.5) +
  scale_x_continuous(breaks = -3:3,
                     labels = c("μ−3σ","μ−2σ","μ−σ","μ","μ+σ","μ+2σ","μ+3σ")) +
  labs(x = "", y = "Probability Density") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

---

## Key Properties of the Normal Distribution

A normal distribution is completely described by just **two parameters**:

| Parameter | Symbol | Meaning |
|---|:---:|---|
| Mean | $\mu$ | Center of the distribution (peak of the bell) |
| Standard deviation | $\sigma$ | Width of the bell — larger σ = wider, flatter curve |

**Key properties:**

1. **Symmetric** around the mean — the left and right halves are mirror images
2. **Mean = Median = Mode** — all three are equal and located at the center
3. The **total area** under the curve = 1 (representing 100% of all probability)
4. The curve extends infinitely in both directions but never touches the x-axis

---

## The 68–95–99.7 Rule (Empirical Rule)

One of the most useful facts about the normal distribution is how much
of the data falls within 1, 2, or 3 standard deviations of the mean.

```{r}
#| label: fig-empirical
#| fig-cap: "The 68–95–99.7 Rule for the Normal Distribution"
#| fig-align: center
#| fig-height: 5
#| fig-width: 8

x <- seq(-3.8, 3.8, length = 1000)
y <- dnorm(x)
df <- data.frame(x = x, y = y)

ggplot(df, aes(x, y)) +
  # 3 SD shading
  geom_area(data = subset(df, x >= -3 & x <= 3),
            aes(x, y), fill = "#fdae61", alpha = 0.5) +
  # 2 SD shading
  geom_area(data = subset(df, x >= -2 & x <= 2),
            aes(x, y), fill = "#abd9e9", alpha = 0.6) +
  # 1 SD shading
  geom_area(data = subset(df, x >= -1 & x <= 1),
            aes(x, y), fill = "#2c7bb6", alpha = 0.6) +
  geom_line(color = "gray30", linewidth = 1.2) +
  # Vertical lines
  geom_vline(xintercept = c(-3,-2,-1,0,1,2,3),
             linetype = "dashed", color = "gray50", linewidth = 0.5) +
  # Percentage annotations
  annotate("text", x = 0, y = 0.20,
           label = "68%", color = "white", size = 5, fontface = "bold") +
  annotate("text", x = 1.5, y = 0.07,
           label = "95%", color = "#2c7bb6", size = 4.5, fontface = "bold") +
  annotate("text", x = 2.5, y = 0.02,
           label = "99.7%", color = "#d7191c", size = 4, fontface = "bold") +
  scale_x_continuous(breaks = -3:3,
                     labels = c("μ−3σ","μ−2σ","μ−σ","μ","μ+σ","μ+2σ","μ+3σ")) +
  labs(x = "", y = "Probability Density",
       caption = "Blue = 68% | Blue+Teal = 95% | All shaded = 99.7%") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

::: {.callout-important icon=false}
## The Empirical Rule

| Range | Proportion of data |
|---|:---:|
| Within **1 SD** of the mean: $\mu \pm \sigma$ | **≈ 68%** |
| Within **2 SD** of the mean: $\mu \pm 2\sigma$ | **≈ 95%** |
| Within **3 SD** of the mean: $\mu \pm 3\sigma$ | **≈ 99.7%** |

This means only about **5%** of values fall more than 2 SD from the mean,
and only **0.3%** fall more than 3 SD away — these are considered unusual or extreme.

*More precisely, 95% of values lie within* **1.96** *SD of the mean.
The rounded value 2 is fine for quick estimates; the exact value 1.96
is the one we will use for confidence intervals in Chapter 4.*
:::

---

## Example 2: Applying the Empirical Rule

The systolic blood pressure (SBP) of adults in a community follows a normal
distribution with **mean = 125 mmHg** and **SD = 14 mmHg**.

**(a)** Between what two values does the SBP of the **middle 95%** of adults fall?

**(b)** What percentage of adults have SBP **above 153 mmHg**?

::: {.callout-tip icon=false}
## Solution

**(a)** The middle 95% falls within $\mu \pm 2\sigma$:

$$125 - 2(14) = 97 \text{ mmHg} \quad \text{to} \quad 125 + 2(14) = 153 \text{ mmHg}$$

So **95% of adults** have SBP between **97 and 153 mmHg**.

**(b)** SBP = 153 mmHg is exactly $\mu + 2\sigma$.
Since 95% of values fall *within* 2 SD, the remaining 5% fall *outside*.
By symmetry, **2.5%** fall above 153 mmHg and 2.5% fall below 97 mmHg.

**Answer:** Approximately **2.5%** of adults have SBP above 153 mmHg.
:::

---

## The Standard Normal Distribution and Z-Score

### The Z-Score

Different normal distributions have different means and SDs. To compare values
across different distributions, we **standardize** them using the **Z-score**:

$$z = \frac{x - \mu}{\sigma}$$

Where:

- $x$ = the observed value
- $\mu$ = the population mean
- $\sigma$ = the population standard deviation
- $z$ = the number of standard deviations $x$ is away from the mean

A **positive Z** means the value is above the mean;
a **negative Z** means it is below the mean.

---

### The Standard Normal Distribution

When we convert all values to Z-scores, the resulting distribution is called
the **standard normal distribution**, with:

$$\mu = 0 \quad \text{and} \quad \sigma = 1$$

This is written as $Z \sim N(0, 1)$.

```{r}
#| label: fig-znorm
#| fig-cap: "Standard Normal Distribution N(0,1)"
#| fig-align: center
#| fig-height: 4
#| fig-width: 7

x <- seq(-3.8, 3.8, length = 500)
df <- data.frame(x = x, y = dnorm(x))

ggplot(df, aes(x, y)) +
  geom_area(fill = "#abd9e9", alpha = 0.5) +
  geom_line(color = "#2c7bb6", linewidth = 1.3) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "#d7191c", linewidth = 1) +
  scale_x_continuous(breaks = -3:3) +
  labs(x = "Z", y = "Probability Density",
       title = "Standard Normal Distribution: N(0, 1)") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

---

### Finding Probabilities Using the Z-Table

The **Z-table** (also called the standard normal table) gives us the
**area to the left** of any Z-score — which equals the **probability**
that a randomly selected observation falls below that value.

```{r}
#| label: tbl-ztable
#| tbl-cap: "Standard Normal Table (area to the LEFT of Z)"

z_vals <- seq(0.0, 3.4, by = 0.1)
decimals <- seq(0.00, 0.09, by = 0.01)

z_matrix <- outer(z_vals, decimals, function(z, d) {
  round(pnorm(z + d), 4)
})

rownames(z_matrix) <- formatC(z_vals, format = "f", digits = 1)
colnames(z_matrix) <- formatC(decimals, format = "f", digits = 2)

kable(z_matrix, caption = "Area to the left of Z (rows = Z integer + 1st decimal, columns = 2nd decimal)") |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE, font_size = tbl_fs(11)) |>
  column_spec(1, bold = TRUE, color = "#2c7bb6")
```

**How to use the table:**
To find $P(Z < 1.23)$, look at row **1.2** and column **0.03**, then read
off the value: $P(Z < 1.23) = 0.8907$.

### Negative Z-Scores: Use Symmetry

The table above only lists **positive** Z-scores. Because the normal curve
is **symmetric** around 0, the area to the left of $-z$ is the same as the
area to the right of $+z$:

$$P(Z < -z) = P(Z > z) = 1 - P(Z < z)$$

> **Example:** $P(Z < -1.50) = 1 - P(Z < 1.50) = 1 - 0.9332 = 0.0668$

---

### Common Z-Table Probability Rules

```{r}
#| label: fig-z-rules
#| fig-cap: "Three useful probability rules for the normal distribution"
#| fig-align: center
#| fig-height: 3
#| fig-width: 10

x <- seq(-3.5, 3.5, length = 500)
y <- dnorm(x)
df <- data.frame(x = x, y = y)
z_val <- 1.5

p1 <- ggplot(df, aes(x, y)) +
  geom_area(data = subset(df, x <= z_val), fill = "#2c7bb6", alpha = 0.6) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = z_val, linetype = "dashed", color = "#d7191c") +
  annotate("text", x = -1, y = 0.15,
           label = paste0("P(Z < ", z_val, ")"), size = 3.5, color = "white") +
  labs(title = "P(Z < z)", x = "Z", y = "") +
  scale_x_continuous(breaks = c(0, z_val),
                     labels = c("0", paste0("z=",z_val))) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank())

p2 <- ggplot(df, aes(x, y)) +
  geom_area(data = subset(df, x >= z_val), fill = "#d7191c", alpha = 0.6) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = z_val, linetype = "dashed", color = "#d7191c") +
  annotate("text", x = 2.5, y = 0.15,
           label = paste0("P(Z > ", z_val, ")"), size = 3.5, color = "white") +
  labs(title = "P(Z > z) = 1 − P(Z < z)", x = "Z", y = "") +
  scale_x_continuous(breaks = c(0, z_val),
                     labels = c("0", paste0("z=",z_val))) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank())

p3 <- ggplot(df, aes(x, y)) +
  geom_area(data = subset(df, x >= -z_val & x <= z_val),
            fill = "#1a9641", alpha = 0.6) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = c(-z_val, z_val),
             linetype = "dashed", color = "#d7191c") +
  annotate("text", x = 0, y = 0.15,
           label = "P(−z < Z < z)", size = 3.5, color = "white") +
  labs(title = "P(z₁ < Z < z₂)", x = "Z", y = "") +
  scale_x_continuous(breaks = c(-z_val, 0, z_val),
                     labels = c(paste0("-",z_val), "0", z_val)) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank())

library(patchwork)
p1 + p2 + p3
```

::: {.callout-important icon=false}
## Key Probability Rules

| Situation | Formula |
|---|---|
| Probability **below** z | $P(Z < z)$ — read directly from Z-table |
| Probability **above** z | $P(Z > z) = 1 - P(Z < z)$ |
| Probability **between** two values | $P(z_1 < Z < z_2) = P(Z < z_2) - P(Z < z_1)$ |
| Probability below a **negative** z | $P(Z < -z) = 1 - P(Z < z)$ |
:::

---

## Example 3: Using Z-Scores and the Z-Table

Body weight of adult males in a region is approximately normally distributed
with **mean = 70 kg** and **SD = 10 kg**.

**(a)** What is the probability that a randomly selected adult male weighs
**less than 85 kg**?

**(b)** What is the probability that he weighs **more than 55 kg**?

**(c)** What is the probability that he weighs **between 60 and 80 kg**?

::: {.callout-tip icon=false}
## Solution

**(a) P(X < 85):**

$$z = \frac{85 - 70}{10} = \frac{15}{10} = 1.50$$

From Z-table: $P(Z < 1.50) = \mathbf{0.9332}$

**93.3%** of adult males weigh less than 85 kg.

---

**(b) P(X > 55):**

$$z = \frac{55 - 70}{10} = \frac{-15}{10} = -1.50$$

Using the symmetry rule, $P(Z < -1.50) = 1 - 0.9332 = 0.0668$, so

$$P(Z > -1.50) = 1 - P(Z < -1.50) = 1 - 0.0668 = \mathbf{0.9332}$$

**93.3%** of adult males weigh more than 55 kg.
*(This makes sense by symmetry — 55 and 85 are equidistant from the mean.)*

---

**(c) P(60 < X < 80):**

$$z_1 = \frac{60 - 70}{10} = -1.00 \qquad z_2 = \frac{80 - 70}{10} = 1.00$$

Using symmetry, $P(Z < -1.00) = 1 - 0.8413 = 0.1587$, so

$$P(-1.00 < Z < 1.00) = P(Z < 1.00) - P(Z < -1.00)$$
$$= 0.8413 - 0.1587 = \mathbf{0.6826}$$

**68.3%** of adult males weigh between 60 and 80 kg —
consistent with the 68–95–99.7 rule.
:::

---

# Key Takeaways

::: {.callout-note icon=false}
## Chapter 3 Summary

1. A **frequency distribution** organizes data into a table showing how often
   each value or range of values occurs. It can show absolute frequency,
   relative frequency, or cumulative frequency.

2. The **normal distribution** is a symmetric, bell-shaped distribution
   described by its mean ($\mu$) and standard deviation ($\sigma$).

3. A **probability** is a number from 0 to 1; for a continuous
   **random variable** it is the **area under the curve**. The
   **68–95–99.7 rule**: approximately 68%, 95%, and 99.7% of values
   fall within 1, 2, and 3 standard deviations of the mean, respectively.

4. A **Z-score** measures how many standard deviations a value is from the mean:
   $z = \dfrac{x - \mu}{\sigma}$

5. The **Z-table** gives the probability (area) to the left of any Z-score,
   allowing us to find the probability that a measurement falls in any range.
:::

---

## Formula Summary

| Quantity | Formula | Notes |
|:----------------|:----------------------------------------|:----------------------|
| Relative frequency | $\dfrac{f}{n}$ | Multiply by 100 for a percentage |
| Class width | $\dfrac{\text{Range}}{\text{Number of classes}}$, rounded up | Usually 5 to 10 classes |
| Empirical rule | $\mu \pm \sigma$: 68%, $\mu \pm 2\sigma$: 95%, $\mu \pm 3\sigma$: 99.7% | Normal distribution only |
| Z-score | $z = \dfrac{x - \mu}{\sigma}$ | Number of SDs from the mean |
| Probability below | $P(Z < z)$ | Read directly from the Z-table |
| Probability above | $P(Z > z) = 1 - P(Z < z)$ |  |
| Probability between | $P(z_1 < Z < z_2) = P(Z < z_2) - P(Z < z_1)$ |  |
| Negative Z | $P(Z < -z) = 1 - P(Z < z)$ | Symmetry of the normal curve |

: Formulas in this chapter {.striped}

---

# Exercises

::: {.callout-caution icon=false}
## Exercise 1

The fasting blood glucose levels (mg/dL) of 25 patients are:

$$95, 112, 88, 134, 102, 119, 145, 97, 108, 125,$$
$$91, 130, 103, 116, 88, 142, 99, 121, 107, 138,$$
$$94, 115, 126, 100, 111$$

**(a)** Construct a frequency distribution table with **5 class intervals**.
Include columns for frequency, relative frequency, percentage, and
cumulative frequency.

**(b)** Which class interval contains the most patients?

**(c)** Using the cumulative frequency column, what percentage of patients
have a blood glucose level below 112 mg/dL?
:::

::: {.callout-caution icon=false}
## Exercise 2

The birth weight of full-term newborns is approximately normally distributed
with **mean = 3,300 g** and **SD = 400 g**.

Using the **68–95–99.7 rule**:

**(a)** What range of birth weights includes the **middle 68%** of newborns?

**(b)** What range includes the **middle 99.7%** of newborns?

**(c)** A newborn weighing less than **2,500 g** is classified as low birth weight.
Based on the empirical rule, is this an unusual outcome? Explain.
:::

::: {.callout-caution icon=false}
## Exercise 3

The resting heart rate of healthy adults follows a normal distribution
with **mean = 72 bpm** and **SD = 8 bpm**.

Find the following probabilities using the Z-table. Show the Z-score
calculation and draw a sketch of the normal curve for each part.

**(a)** $P(\text{heart rate} < 80)$

**(b)** $P(\text{heart rate} > 88)$

**(c)** $P(64 < \text{heart rate} < 88)$

**(d)** A heart rate below 60 bpm is considered **bradycardia**.
What proportion of healthy adults would be classified as having bradycardia
based on this distribution?
:::

---

*End of Chapter 3*

---

::: {.callout-note icon=false}
## Coming Up Next
**Chapter 4: Estimation** — Using what we know about the normal distribution,
we will learn how to estimate population parameters from sample data
and construct **confidence intervals** that quantify our uncertainty.
:::
